---
title: "Gut Content Metabarcoding Study: Sample Metadata Summary"
author: ""
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  word_document:
    reference_docx: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 7)

library(dplyr)
library(readr)
library(flextable)

# Set flextable defaults for publication quality
set_flextable_defaults(
  font.family = "Times New Roman",
  font.size = 11,
  padding = 6
)
```

```{r load-data, include=FALSE}
# Read sample-to-fish mapping files for both primer sets
# Skip the #q2:types row (comment = "#")
ps1_samples <- read_tsv(
  "../input_data/metadata/primerset1_sample_to_fish.tsv",
  comment = "#",
  show_col_types = FALSE
) %>%
  filter(!is.na(fish_id)) %>%  # Remove BLANK samples
  mutate(Primer_Set = "PS1")

ps2_samples <- read_tsv(
  "../input_data/metadata/primerset2_sample_to_fish.tsv",
  comment = "#",
  show_col_types = FALSE
) %>%
  filter(!is.na(fish_id)) %>%  # Remove BLANK samples
  mutate(Primer_Set = "PS2")

# Combine both primer sets
all_samples <- bind_rows(ps1_samples, ps2_samples)

# Read fish metadata (also skip #q2:types row)
fish_metadata <- read_tsv(
  "../input_data/metadata/fish_id_metadata.tsv",
  comment = "#",
  show_col_types = FALSE
)

# Join sample data with fish metadata
data <- all_samples %>%
  left_join(fish_metadata, by = c("fish_id" = "sampleid")) %>%
  mutate(Locality_Name = trimws(Locality_Name)) %>%
  rename(Sample_Name = fish_id, Sample_ID = sampleid)
```

# Dataset Overview

This report summarizes the metadata for `r n_distinct(data$Sample_Name)`
distinct fish samples analyzed in the gut content metabarcoding study.
Of these, `r data %>% filter(Primer_Set == "PS1") %>% distinct(Sample_Name) %>% nrow()`
samples were sequenced with primer set PS1 and
`r data %>% filter(Primer_Set == "PS2") %>% distinct(Sample_Name) %>% nrow()`
samples were sequenced with primer set PS2, resulting in
`r nrow(data)` total sequencing runs.

---

# Dataset-Wide Statistics

## Total Sample Count

**Total distinct samples:** `r n_distinct(data$Sample_Name)`

## Samples by Species

```{r species-table}
samples_by_species <- data %>%
  group_by(Genus, Species) %>%
  summarise(n_samples = n_distinct(Sample_Name), .groups = "drop") %>%
  arrange(Genus, Species)

samples_by_species %>%
  flextable() %>%
  set_header_labels(
    Genus = "Genus",
    Species = "Species",
    n_samples = "N Samples"
  ) %>%
  theme_booktabs() %>%
  autofit() %>%
  align(j = 3, align = "center", part = "all") %>%
  bold(part = "header")
```

**Total species:** `r nrow(samples_by_species)` across `r n_distinct(samples_by_species$Genus)` genera

## Primer Set Coverage

**Samples sequenced with both primer sets (PS1 and PS2):** `r data %>% group_by(Sample_Name) %>% summarise(n_primer_sets = n_distinct(Primer_Set), .groups = "drop") %>% filter(n_primer_sets == 2) %>% nrow()`

## Samples by Sex

```{r sex-table}
samples_by_sex <- data %>%
  group_by(Sex) %>%
  summarise(n_samples = n_distinct(Sample_Name), .groups = "drop") %>%
  mutate(Sex = ifelse(is.na(Sex), "Unknown", Sex)) %>%
  arrange(Sex)

samples_by_sex %>%
  flextable() %>%
  set_header_labels(
    Sex = "Sex",
    n_samples = "N Samples"
  ) %>%
  theme_booktabs() %>%
  autofit() %>%
  align(j = 2, align = "center", part = "all") %>%
  bold(part = "header")
```

## Parasite Prevalence

**Samples with parasites detected:** `r data %>% filter(Parasite == 1) %>% summarise(n_samples = n_distinct(Sample_Name)) %>% pull(n_samples)` of `r n_distinct(data$Sample_Name)` (`r round(100 * (data %>% filter(Parasite == 1) %>% summarise(n_samples = n_distinct(Sample_Name)) %>% pull(n_samples)) / n_distinct(data$Sample_Name), 1)`%)

---

# Locality-Level Breakdown

```{r locality-stats, results='asis'}
localities <- data %>%
  distinct(Locality_Name) %>%
  pull(Locality_Name) %>%
  sort()

for (locality in localities) {
  cat("\n## ", locality, "\n\n", sep = "")

  locality_data <- data %>% filter(Locality_Name == !!locality)

  # Total samples
  loc_total <- locality_data %>%
    summarise(n_samples = n_distinct(Sample_Name)) %>%
    pull(n_samples)

  cat("**Total samples:** ", loc_total, "\n\n", sep = "")

  # Species composition
  loc_species <- locality_data %>%
    group_by(Genus, Species) %>%
    summarise(n_samples = n_distinct(Sample_Name), .groups = "drop") %>%
    arrange(Genus, Species)

  if (nrow(loc_species) > 0) {
    cat("**Species composition:**\n\n")
    ft <- loc_species %>%
      flextable() %>%
      set_header_labels(
        Genus = "Genus",
        Species = "Species",
        n_samples = "N"
      ) %>%
      theme_booktabs() %>%
      autofit() %>%
      align(j = 3, align = "center", part = "all") %>%
      bold(part = "header") %>%
      fontsize(size = 10, part = "all")

    ft <- ft %>% flextable::width(j = 1, width = 1.2) %>%
      flextable::width(j = 2, width = 2.5) %>%
      flextable::width(j = 3, width = 0.6)

    flextable_to_rmd(ft)
    cat("\n\n")
  }

  # Primer set coverage
  loc_both_primers <- locality_data %>%
    group_by(Sample_Name) %>%
    summarise(n_primer_sets = n_distinct(Primer_Set), .groups = "drop") %>%
    filter(n_primer_sets == 2) %>%
    nrow()

  cat("**Samples with both primer sets:** ", loc_both_primers, " of ", loc_total,
      " (", round(100 * loc_both_primers / loc_total, 1), "%)\n\n", sep = "")

  # Sex distribution
  loc_sex <- locality_data %>%
    group_by(Sex) %>%
    summarise(n_samples = n_distinct(Sample_Name), .groups = "drop") %>%
    mutate(Sex = ifelse(is.na(Sex), "Unknown", Sex)) %>%
    arrange(Sex)

  if (nrow(loc_sex) > 0) {
    cat("**Sex distribution:**\n\n")
    ft <- loc_sex %>%
      flextable() %>%
      set_header_labels(
        Sex = "Sex",
        n_samples = "N"
      ) %>%
      theme_booktabs() %>%
      autofit() %>%
      align(j = 2, align = "center", part = "all") %>%
      bold(part = "header") %>%
      fontsize(size = 10, part = "all")

    ft <- ft %>% flextable::width(j = 1, width = 1.5) %>%
      flextable::width(j = 2, width = 0.6)

    flextable_to_rmd(ft)
    cat("\n\n")
  }

  # Parasites
  loc_parasites <- locality_data %>%
    filter(Parasite == 1) %>%
    summarise(n_samples = n_distinct(Sample_Name)) %>%
    pull(n_samples)

  cat("**Samples with parasites:** ", loc_parasites, " of ", loc_total,
      " (", round(100 * loc_parasites / loc_total, 1), "%)\n\n", sep = "")

  cat("---\n\n")
}
```

# Summary

This dataset comprises samples from `r length(localities)` localities across the study region, with the majority of samples belonging to the genus *Paramormyrops* (particularly *P. kingsleyae*). Most samples (`r round(100 * (data %>% group_by(Sample_Name) %>% summarise(n_primer_sets = n_distinct(Primer_Set)) %>% filter(n_primer_sets == 2) %>% nrow()) / n_distinct(data$Sample_Name), 1)`%) were successfully sequenced with both primer sets, providing comprehensive coverage for downstream metabarcoding analyses.
