---
title: "Bale Creek Mormyrid Community Dietary Variation: Comprehensive Analysis"
author: "Gorsuch, Picq, Gallant"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
    keep_tex: false
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
knit: (function(inputFile, encoding) { root <- rprojroot::find_root(".git/index"); rmarkdown::render(inputFile, encoding = encoding, output_dir = file.path(root, "output_data/07_Analysis")) })
header-includes:
- \usepackage{longtable}
- \usepackage{booktabs}
- \usepackage{float}
---

```{r setup, include=FALSE}
# Configure knitr
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  dpi = 300
)

# Set root and source core setup
root <- rprojroot::find_root(".git/index")
source(file.path(root, "code/07_Analysis/core_setup.R"))
source(file.path(root, "code/07_Analysis/gut_analysis_2026_helpers.R"))

# Load required libraries
library(tidyverse)
library(phyloseq)
library(vegan)
library(knitr)
library(kableExtra)
library(patchwork)

# This loads from core_setup.R:
# - ps1, ps2 (full primer set data)
# - ps1_bale, ps2_bale (Bale Creek subsets with rare taxa removed)
# - GLOBAL_FAMILY_COLORS (unified color palette)
# - Color variables (bp_color, tp_color, ps1_color, ps2_color)
```

# Overview

## Biological Context

This analysis examines dietary variation across the multi-species mormyrid community at Bale Creek, a single locality containing 8-9 sympatric species. Unlike single-species analyses, this community-level approach allows us to assess:

1. **Interspecific dietary differentiation**: Do different mormyrid species exploit different prey resources?
2. **Waveform effects across species**: Does electrosensory phenotype (BP vs TP) influence diet independent of species identity?
3. **Individual-level variation**: How do sex and body size contribute to dietary variation within this community?

## Research Questions

1. What is the relative contribution of **Species identity** vs **Waveform type** to dietary variation?
2. Do **Sex** and **Body size** explain additional variation beyond species and waveform?
3. Are dietary patterns concordant between PS1 and PS2 primer sets?
4. Which prey families characterize the diets of different species?

## Analytical Approach

We use a comprehensive PERMANOVA framework with:

- **Marginal (Type III) tests** to assess independent contributions of each factor
- **Variance partitioning** to quantify shared vs unique variance components
- **Dispersion tests** to validate PERMANOVA assumptions
- **Dual primer set validation** to ensure findings are robust across metabarcoding methods

**Key differences from P. kingsleyae single-species analysis:**

- Multi-species community (8-9 species) vs single species
- No Region factor (single locality)
- No Parasite factor (absent in Bale Creek samples)
- Primary focus on Species differences with Waveform as secondary factor

---

# Data Preparation

## Complete-Case Filtering

```{r data-prep-functions}
prepare_bale_comprehensive_data <- function(ps) {
  # Extract metadata
  meta <- data.frame(sample_data(ps), stringsAsFactors = FALSE)
  cat("Starting samples:", nrow(meta), "\n")

  # Clean species names (expands P. → Paramormyrops, etc.)
  meta$Species <- clean_species_name(meta$Species)

  # Add body_size from SL
  meta$body_size <- as.numeric(meta$SL)

  # Ensure factors with proper levels
  meta$Species <- factor(meta$Species)
  meta$Sex <- factor(meta$Sex, levels = c("Male", "Female", "Juvenile"))
  meta$Waveform_Type <- factor(meta$Waveform_Type, levels = c("BP", "TP"))

  # Filter to complete cases
  complete_cases <- !is.na(meta$Species) &
                    !is.na(meta$Sex) &
                    !is.na(meta$body_size) &
                    !is.na(meta$Waveform_Type)

  cat("Complete cases:", sum(complete_cases), "\n")
  cat("Removed due to missing data:", sum(!complete_cases), "\n")

  meta_filt <- meta[complete_cases, ]

  # Check species with <3 samples (warn but don't filter yet)
  species_counts <- table(meta_filt$Species)
  low_n_species <- names(species_counts)[species_counts < 3]
  if (length(low_n_species) > 0) {
    cat("\nWARNING: Species with <3 samples (unstable for PERMANOVA):\n")
    for (sp in low_n_species) {
      cat("  -", sp, "(n =", species_counts[sp], ")\n")
    }
  }

  # Prune phyloseq and calculate distance matrix
  ps_filt <- prune_samples(rownames(meta_filt), ps)
  ps_filt <- prune_taxa(taxa_sums(ps_filt) > 0, ps_filt)

  otu_mat <- as.matrix(otu_table(ps_filt))
  if (taxa_are_rows(ps_filt)) {
    otu_mat <- t(otu_mat)
  }

  dist_mat <- vegdist(otu_mat, method = "bray")
  rownames(meta_filt) <- rownames(otu_mat)

  return(list(
    distance = dist_mat,
    metadata = meta_filt,
    otu = otu_mat,
    ps_filtered = ps_filt
  ))
}
```

```{r apply-filtering}
cat("=== PREPARING PS1 DATA ===\n")
comp_ps1 <- prepare_bale_comprehensive_data(ps1_bale)

cat("\n=== PREPARING PS2 DATA ===\n")
comp_ps2 <- prepare_bale_comprehensive_data(ps2_bale)
```

## Sample Size Summary

```{r sample-size-summary}
# Create comprehensive sample size table
create_sample_summary <- function(metadata, ps_name) {
  cat("\n### Sample Size Summary -", ps_name, "###\n\n")

  # Overall counts
  cat("Total samples:", nrow(metadata), "\n\n")

  # By Species
  cat("By Species:\n")
  print(sort(table(metadata$Species), decreasing = TRUE))
  cat("\n")

  # By Waveform
  cat("By Waveform:\n")
  print(table(metadata$Waveform_Type))
  cat("\n")

  # By Sex
  cat("By Sex:\n")
  print(table(metadata$Sex))
  cat("\n")

  # Cross-tabulation: Species × Waveform
  cat("Species × Waveform Cross-tabulation:\n")
  cross_tab <- table(metadata$Species, metadata$Waveform_Type)
  print(cross_tab)
  cat("\n")

  # Check for confounding
  confounded_species <- rowSums(cross_tab > 0) == 1
  if (any(confounded_species)) {
    cat("WARNING: Species-Waveform confounding detected!\n")
    cat("The following species have only one waveform type:\n")
    conf_names <- names(confounded_species)[confounded_species]
    for (sp in conf_names) {
      wf <- colnames(cross_tab)[cross_tab[sp, ] > 0]
      cat("  -", sp, ":", wf, "\n")
    }
    cat("\n")
  }

  # Body size summary by Species
  cat("Body size (SL) summary by Species:\n")
  size_summary <- metadata %>%
    group_by(Species) %>%
    summarise(
      n = n(),
      mean_SL = mean(body_size, na.rm = TRUE),
      sd_SL = sd(body_size, na.rm = TRUE),
      min_SL = min(body_size, na.rm = TRUE),
      max_SL = max(body_size, na.rm = TRUE),
      .groups = "drop"
    )
  print(kable(size_summary, digits = 1, format = "simple"))

  return(list(
    species_counts = table(metadata$Species),
    cross_tab = cross_tab,
    confounding = any(confounded_species)
  ))
}

summary_ps1 <- create_sample_summary(comp_ps1$metadata, "PS1")
summary_ps2 <- create_sample_summary(comp_ps2$metadata, "PS2")
```

```{r sample-size-table}
# Create formatted table for PDF
sample_counts <- data.frame(
  Species = names(summary_ps1$species_counts),
  PS1_n = as.numeric(summary_ps1$species_counts),
  PS2_n = as.numeric(summary_ps2$species_counts[names(summary_ps1$species_counts)])
)

kable(sample_counts,
      col.names = c("Species", "PS1 (n)", "PS2 (n)"),
      caption = "Sample sizes by species and primer set after complete-case filtering") %>%
  kable_styling(latex_options = c("hold_position"))
```

---

# Prey Composition Patterns

## Family-Level Composition by Species

```{r prey-composition}
# Aggregate to family level
ps1_bale_family <- tax_glom(ps1_bale, taxrank = "Family", NArm = FALSE)
ps2_bale_family <- tax_glom(ps2_bale, taxrank = "Family", NArm = FALSE)

# Transform to relative abundance
ps1_rel <- transform_sample_counts(ps1_bale_family, function(x) x / sum(x))
ps2_rel <- transform_sample_counts(ps2_bale_family, function(x) x / sum(x))

# Create unified top families list
get_top_families <- function(ps_list, n_top = 10) {
  all_families <- character()

  for (ps in ps_list) {
    # Get taxonomy table
    tax <- as.data.frame(tax_table(ps))

    # Get family names (raw)
    raw_families <- tax$Family

    # Clean family names
    clean_families <- sapply(raw_families, clean_family_name)

    # Get abundance totals by cleaned family name
    ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))
    otu_mat <- as.matrix(otu_table(ps_rel))
    if (!taxa_are_rows(ps_rel)) {
      otu_mat <- t(otu_mat)
    }

    # Sum by cleaned family
    family_totals <- tapply(rowSums(otu_mat), clean_families, sum)

    # Get top families
    top_fams <- names(sort(family_totals, decreasing = TRUE)[1:n_top])
    all_families <- c(all_families, top_fams)
  }

  return(unique(all_families))
}

top_families_unified <- get_top_families(list(ps1_bale_family, ps2_bale_family), n_top = 10)

cat("Top prey families (unified across PS1 and PS2):\n")
print(top_families_unified)

# Debug: Show raw family names in the data
cat("\nRaw family names (first 20):\n")
print(head(unique(tax_table(ps1_bale_family)[, "Family"]), 20))

# Ensure all top families have colors in global palette
missing_families <- setdiff(top_families_unified, names(GLOBAL_FAMILY_COLORS))
if (length(missing_families) > 0) {
  cat("\nAdding colors for", length(missing_families), "missing families:\n")
  cat("  ", paste(missing_families, collapse = ", "), "\n")

  distinct_colors <- c(
    "#4477AA", "#EE6677", "#228833", "#CCBB44", "#66CCEE", "#AA3377",
    "#EE3377", "#009988", "#CC3311", "#EE7733", "#0077BB", "#33BBEE",
    "#CC6677", "#882255", "#AA4499", "#117733", "#332288", "#44AA99"
  )

  used_colors <- GLOBAL_FAMILY_COLORS[names(GLOBAL_FAMILY_COLORS) != "Other"]
  available_colors <- setdiff(distinct_colors, used_colors)

  new_colors <- setNames(
    available_colors[1:length(missing_families)],
    missing_families
  )
  GLOBAL_FAMILY_COLORS <- c(GLOBAL_FAMILY_COLORS, new_colors)
}
```

```{r composition-barplot, fig.height=6, fig.width=10}
# Composition bar plot by species
ps1_melt <- psmelt(ps1_rel) %>% mutate(PrimerSet = "PS1")
ps2_melt <- psmelt(ps2_rel) %>% mutate(PrimerSet = "PS2")
ps_combined <- bind_rows(ps1_melt, ps2_melt)

# Clean names (note: psmelt adds "sample_" prefix to metadata columns)
ps_combined$Species_clean <- clean_species_name(ps_combined$sample_Species)
ps_combined$Family_clean <- sapply(ps_combined$Family, clean_family_name)

# Aggregate
ps_agg <- ps_combined %>%
  group_by(Species_clean, PrimerSet, Family_clean) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  group_by(Species_clean, PrimerSet) %>%
  mutate(RelAbund = Abundance / sum(Abundance))

# Categorize families
ps_final <- ps_agg %>%
  mutate(Family_display = ifelse(Family_clean %in% top_families_unified,
                                  Family_clean, "Other")) %>%
  group_by(Species_clean, PrimerSet, Family_display) %>%
  summarise(RelAbund = sum(RelAbund), .groups = "drop")

# Debug: Check what families we have
cat("\nFamilies in data:\n")
print(unique(ps_final$Family_display))
cat("\nFamilies in color palette:\n")
print(names(GLOBAL_FAMILY_COLORS))

ps_final$Family_display <- factor(ps_final$Family_display,
                                   levels = c(top_families_unified, "Other"))

# Plot
ggplot(ps_final, aes(x = Species_clean, y = RelAbund, fill = Family_display)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ PrimerSet, nrow = 1) +
  scale_fill_manual(
    values = GLOBAL_FAMILY_COLORS,
    name = "Prey Family",
    drop = FALSE,
    na.value = "grey50"
  ) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right",
    legend.key.size = unit(0.4, "cm")
  ) +
  labs(x = "Fish Species", y = "Relative Abundance",
       title = "Prey Family Composition by Species and Primer Set") +
  guides(fill = guide_legend(ncol = 1))
```

---

# Alpha Diversity Analysis

## Shannon Diversity by Species

```{r alpha-diversity}
# Analyze alpha diversity for both primer sets
alpha_ps1_bale <- analyze_alpha_diversity(comp_ps1$ps_filtered, "Species")
alpha_ps2_bale <- analyze_alpha_diversity(comp_ps2$ps_filtered, "Species")

cat("=== PS1 Shannon Diversity by Species ===\n")
cat("Test type:", alpha_ps1_bale$test_type, "\n")
cat("Shapiro-Wilk p-value:", round(alpha_ps1_bale$shapiro_p, 4), "\n")
if (alpha_ps1_bale$test_type == "Kruskal-Wallis") {
  print(alpha_ps1_bale$test)
} else {
  print(summary(alpha_ps1_bale$test))
}

cat("\n=== PS2 Shannon Diversity by Species ===\n")
cat("Test type:", alpha_ps2_bale$test_type, "\n")
cat("Shapiro-Wilk p-value:", round(alpha_ps2_bale$shapiro_p, 4), "\n")
if (alpha_ps2_bale$test_type == "Kruskal-Wallis") {
  print(alpha_ps2_bale$test)
} else {
  print(summary(alpha_ps2_bale$test))
}
```

```{r alpha-boxplot, fig.height=5, fig.width=10}
# Combine PS1 and PS2 for comparison
div_ps1 <- estimate_richness(comp_ps1$ps_filtered, measures = "Shannon")
div_ps1$Species <- comp_ps1$metadata$Species
div_ps1$PrimerSet <- "PS1"

div_ps2 <- estimate_richness(comp_ps2$ps_filtered, measures = "Shannon")
div_ps2$Species <- comp_ps2$metadata$Species
div_ps2$PrimerSet <- "PS2"

div_combined <- bind_rows(div_ps1, div_ps2)
div_combined$PrimerSet <- factor(div_combined$PrimerSet, levels = c("PS1", "PS2"))

# Grouped boxplot
ggplot(div_combined, aes(x = Species, y = Shannon, fill = PrimerSet)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(0.8)) +
  scale_fill_manual(values = c(PS1 = ps1_color, PS2 = ps2_color)) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  ) +
  labs(x = "Fish Species", y = "Shannon Diversity",
       fill = "Primer Set",
       title = "Shannon Diversity by Species and Primer Set")
```

---

# Beta Diversity: Comprehensive PERMANOVA

## Marginal (Type III) Tests

```{r permanova-marginal}
cat("=== PS1 PERMANOVA (Marginal Effects) ===\n")
perm_ps1_marginal <- adonis2(
  comp_ps1$distance ~ Species + Waveform_Type + Sex + body_size,
  data = comp_ps1$metadata,
  permutations = 999,
  by = "margin"
)
print(perm_ps1_marginal)

cat("\n=== PS2 PERMANOVA (Marginal Effects) ===\n")
perm_ps2_marginal <- adonis2(
  comp_ps2$distance ~ Species + Waveform_Type + Sex + body_size,
  data = comp_ps2$metadata,
  permutations = 999,
  by = "margin"
)
print(perm_ps2_marginal)
```

```{r permanova-table}
# Format PERMANOVA results for publication
format_permanova <- function(perm_result, ps_name) {
  perm_df <- as.data.frame(perm_result)
  perm_df$Term <- rownames(perm_df)

  effects <- perm_df %>%
    filter(Term != "Residual") %>%
    mutate(
      Factor = case_when(
        Term == "Waveform_Type" ~ "Waveform",
        Term == "body_size" ~ "Body size",
        TRUE ~ Term
      ),
      R2_pct = R2 * 100,
      P_value = `Pr(>F)`
    ) %>%
    select(Factor, Df, SumOfSqs, R2_pct, F, P_value) %>%
    arrange(desc(R2_pct))

  return(effects)
}

effects_ps1 <- format_permanova(perm_ps1_marginal, "PS1")
effects_ps2 <- format_permanova(perm_ps2_marginal, "PS2")

cat("\n=== Effect Size Rankings ===\n\n")
cat("PS1:\n")
print(kable(effects_ps1, digits = c(0, 0, 1, 2, 2, 4),
            col.names = c("Factor", "df", "SS", "R² (%)", "F", "p"),
            format = "simple"))

cat("\nPS2:\n")
print(kable(effects_ps2, digits = c(0, 0, 1, 2, 2, 4),
            col.names = c("Factor", "df", "SS", "R² (%)", "F", "p"),
            format = "simple"))
```

## Variance Partitioning

```{r variance-partitioning}
cat("=== Variance Partitioning: Two-way ===\n\n")

# Two-way partition: Ecological vs Individual
vp_2way_ps1 <- varpart(
  comp_ps1$distance,
  ~ Species + Waveform_Type,  # Ecological
  ~ Sex + body_size,           # Individual
  data = comp_ps1$metadata
)

cat("PS1 - Ecological (Species + Waveform) vs Individual (Sex + Body size):\n")
print(vp_2way_ps1)

vp_2way_ps2 <- varpart(
  comp_ps2$distance,
  ~ Species + Waveform_Type,
  ~ Sex + body_size,
  data = comp_ps2$metadata
)

cat("\nPS2 - Ecological vs Individual:\n")
print(vp_2way_ps2)

cat("\n=== Variance Partitioning: Three-way ===\n\n")

# Three-way partition
vp_3way_ps1 <- varpart(
  comp_ps1$distance,
  ~ Species,              # Taxonomic identity
  ~ Waveform_Type,        # Electrosensory phenotype
  ~ Sex + body_size,      # Individual factors
  data = comp_ps1$metadata
)

cat("PS1 - Species vs Waveform vs Individual:\n")
print(vp_3way_ps1)

vp_3way_ps2 <- varpart(
  comp_ps2$distance,
  ~ Species,
  ~ Waveform_Type,
  ~ Sex + body_size,
  data = comp_ps2$metadata
)

cat("\nPS2 - Species vs Waveform vs Individual:\n")
print(vp_3way_ps2)
```

## PCoA Ordination

```{r pcoa-ps1, fig.height=5, fig.width=7}
# PS1 PCoA
pcoa_ps1 <- cmdscale(comp_ps1$distance, eig = TRUE, k = 2)
var_exp_ps1 <- pcoa_ps1$eig / sum(pcoa_ps1$eig) * 100

pcoa_df_ps1 <- data.frame(
  PC1 = pcoa_ps1$points[, 1],
  PC2 = pcoa_ps1$points[, 2],
  Waveform = comp_ps1$metadata$Waveform_Type,
  Species = comp_ps1$metadata$Species,
  Sex = comp_ps1$metadata$Sex
)

ggplot(pcoa_df_ps1, aes(x = PC1, y = PC2, color = Species, shape = Waveform)) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(aes(group = Species), type = "norm", level = 0.95) +
  scale_shape_manual(values = c(BP = 17, TP = 16)) +
  labs(
    x = paste0("PC1 (", round(var_exp_ps1[1], 1), "%)"),
    y = paste0("PC2 (", round(var_exp_ps1[2], 1), "%)"),
    title = "PCoA of Dietary Variation - PS1",
    subtitle = "Bray-Curtis dissimilarity"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")
```

```{r pcoa-ps2, fig.height=5, fig.width=7}
# PS2 PCoA
pcoa_ps2 <- cmdscale(comp_ps2$distance, eig = TRUE, k = 2)
var_exp_ps2 <- pcoa_ps2$eig / sum(pcoa_ps2$eig) * 100

pcoa_df_ps2 <- data.frame(
  PC1 = pcoa_ps2$points[, 1],
  PC2 = pcoa_ps2$points[, 2],
  Waveform = comp_ps2$metadata$Waveform_Type,
  Species = comp_ps2$metadata$Species,
  Sex = comp_ps2$metadata$Sex
)

ggplot(pcoa_df_ps2, aes(x = PC1, y = PC2, color = Species, shape = Waveform)) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(aes(group = Species), type = "norm", level = 0.95) +
  scale_shape_manual(values = c(BP = 17, TP = 16)) +
  labs(
    x = paste0("PC1 (", round(var_exp_ps2[1], 1), "%)"),
    y = paste0("PC2 (", round(var_exp_ps2[2], 1), "%)"),
    title = "PCoA of Dietary Variation - PS2",
    subtitle = "Bray-Curtis dissimilarity"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")
```

## Dispersion Tests

```{r dispersion-tests}
test_dispersion <- function(dist_mat, metadata, factor_var, ps_name) {
  bd <- betadisper(dist_mat, metadata[[factor_var]])
  bd_test <- permutest(bd, permutations = 999)

  cat(ps_name, "-", factor_var, "\n")
  cat("  F =", round(bd_test$statistic, 3),
      ", p =", round(bd_test$tab$`Pr(>F)`[1], 4), "\n")

  if (bd_test$tab$`Pr(>F)`[1] < 0.05) {
    cat("  WARNING: Heterogeneous dispersion detected!\n")
  }

  return(list(test = bd_test, betadisper = bd))
}

cat("=== Dispersion Tests (Homogeneity of Variance) ===\n\n")

# Test for each categorical factor
cat("Species:\n")
disp_species_ps1 <- test_dispersion(comp_ps1$distance, comp_ps1$metadata, "Species", "PS1")
disp_species_ps2 <- test_dispersion(comp_ps2$distance, comp_ps2$metadata, "Species", "PS2")

cat("\nWaveform:\n")
disp_waveform_ps1 <- test_dispersion(comp_ps1$distance, comp_ps1$metadata, "Waveform_Type", "PS1")
disp_waveform_ps2 <- test_dispersion(comp_ps2$distance, comp_ps2$metadata, "Waveform_Type", "PS2")

cat("\nSex:\n")
disp_sex_ps1 <- test_dispersion(comp_ps1$distance, comp_ps1$metadata, "Sex", "PS1")
disp_sex_ps2 <- test_dispersion(comp_ps2$distance, comp_ps2$metadata, "Sex", "PS2")
```

---

# Publication Figure: 4-Panel Synthesis

```{r panel-functions}
# Panel A: Prey Composition (Species × Primer Set)
create_panel_a_bale <- function(ps1_family, ps2_family, top_families) {
  # Transform to relative abundance
  ps1_rel <- transform_sample_counts(ps1_family, function(x) x / sum(x))
  ps2_rel <- transform_sample_counts(ps2_family, function(x) x / sum(x))

  # Melt and combine
  ps1_melt <- psmelt(ps1_rel) %>% mutate(PrimerSet = "PS1")
  ps2_melt <- psmelt(ps2_rel) %>% mutate(PrimerSet = "PS2")
  ps_combined <- bind_rows(ps1_melt, ps2_melt)

  # Clean names (psmelt adds "sample_" prefix to metadata columns)
  ps_combined$Species_clean <- clean_species_name(ps_combined$sample_Species)
  ps_combined$Family_clean <- sapply(ps_combined$Family, clean_family_name)

  # Aggregate by Species × PrimerSet × Family
  ps_agg <- ps_combined %>%
    group_by(Species_clean, PrimerSet, Family_clean) %>%
    summarise(Abundance = sum(Abundance), .groups = "drop") %>%
    group_by(Species_clean, PrimerSet) %>%
    mutate(RelAbund = Abundance / sum(Abundance))

  # Categorize into top families vs "Other"
  ps_final <- ps_agg %>%
    mutate(Family_display = ifelse(Family_clean %in% top_families,
                                    Family_clean, "Other")) %>%
    group_by(Species_clean, PrimerSet, Family_display) %>%
    summarise(RelAbund = sum(RelAbund), .groups = "drop")

  # Factor levels for consistent legend
  ps_final$Family_display <- factor(ps_final$Family_display,
                                     levels = c(top_families, "Other"))

  # Plot
  p <- ggplot(ps_final, aes(x = Species_clean, y = RelAbund, fill = Family_display)) +
    geom_bar(stat = "identity", position = "stack") +
    facet_wrap(~ PrimerSet, nrow = 1) +
    scale_fill_manual(values = GLOBAL_FAMILY_COLORS,
                      name = "Prey Family", drop = FALSE) +
    scale_y_continuous(labels = scales::percent) +
    theme_minimal(base_size = 9) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
      legend.position = "right",
      legend.key.size = unit(0.3, "cm")
    ) +
    labs(x = "Fish Species", y = "Relative Abundance") +
    guides(fill = guide_legend(ncol = 1))

  return(p)
}

# Panel B: Shannon Diversity (Species × Primer Set)
create_panel_b_bale <- function(comp_ps1, comp_ps2) {
  # Calculate Shannon for both primer sets
  div_ps1 <- estimate_richness(comp_ps1$ps_filtered, measures = "Shannon")
  div_ps1$Species <- comp_ps1$metadata$Species
  div_ps1$PrimerSet <- "PS1"

  div_ps2 <- estimate_richness(comp_ps2$ps_filtered, measures = "Shannon")
  div_ps2$Species <- comp_ps2$metadata$Species
  div_ps2$PrimerSet <- "PS2"

  # Combine
  div_combined <- bind_rows(div_ps1, div_ps2)
  div_combined$PrimerSet <- factor(div_combined$PrimerSet,
                                    levels = c("PS1", "PS2"))

  # Grouped boxplot
  p <- ggplot(div_combined, aes(x = Species, y = Shannon, fill = PrimerSet)) +
    geom_boxplot(alpha = 0.7, position = position_dodge(0.8)) +
    scale_fill_manual(values = c(PS1 = ps1_color, PS2 = ps2_color)) +
    theme_minimal(base_size = 9) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
      legend.position = "bottom"
    ) +
    labs(x = "Fish Species", y = "Shannon Diversity", fill = "Primer Set")

  return(p)
}

# Panel C: PCoA (Beta Diversity) - PS1 Only
create_panel_c_bale <- function(comp_ps1) {
  # PCoA
  pcoa_result <- cmdscale(comp_ps1$distance, eig = TRUE, k = 2)
  var_exp <- pcoa_result$eig / sum(pcoa_result$eig) * 100

  pcoa_df <- data.frame(
    PC1 = pcoa_result$points[, 1],
    PC2 = pcoa_result$points[, 2],
    Waveform = comp_ps1$metadata$Waveform_Type,
    Species = comp_ps1$metadata$Species
  )

  p <- ggplot(pcoa_df, aes(x = PC1, y = PC2,
                           color = Species, shape = Waveform)) +
    geom_point(size = 2.5, alpha = 0.7) +
    stat_ellipse(aes(group = Species), type = "norm", level = 0.95) +
    scale_shape_manual(values = c(BP = 17, TP = 16)) +
    labs(
      x = paste0("PC1 (", round(var_exp[1], 1), "%)"),
      y = paste0("PC2 (", round(var_exp[2], 1), "%)")
    ) +
    annotate("text", x = Inf, y = -Inf,
             label = "PS2 concordant; Fig. SX",
             hjust = 1.1, vjust = -0.5, size = 2.5, fontface = "italic") +
    theme_minimal(base_size = 9) +
    theme(
      legend.position = "right",
      legend.key.size = unit(0.3, "cm")
    )

  return(p)
}

# Panel D: PERMANOVA Effect Sizes
create_panel_d_bale <- function(perm_ps1_marginal, perm_ps2_marginal) {
  # Extract R² and p-values from PERMANOVA results
  ps1_df <- as.data.frame(perm_ps1_marginal)
  ps1_df$Term <- rownames(ps1_df)

  ps2_df <- as.data.frame(perm_ps2_marginal)
  ps2_df$Term <- rownames(ps2_df)

  # Create effects dataframes
  effects_ps1 <- ps1_df %>%
    filter(Term %in% c("Species", "Waveform_Type", "Sex", "body_size")) %>%
    select(Term, R2, `Pr(>F)`) %>%
    mutate(
      Factor = case_when(
        Term == "Waveform_Type" ~ "Waveform",
        Term == "body_size" ~ "Body size",
        TRUE ~ Term
      ),
      PrimerSet = "PS1",
      P_value = `Pr(>F)`,
      Significant = P_value < 0.05
    )

  effects_ps2 <- ps2_df %>%
    filter(Term %in% c("Species", "Waveform_Type", "Sex", "body_size")) %>%
    select(Term, R2, `Pr(>F)`) %>%
    mutate(
      Factor = case_when(
        Term == "Waveform_Type" ~ "Waveform",
        Term == "body_size" ~ "Body size",
        TRUE ~ Term
      ),
      PrimerSet = "PS2",
      P_value = `Pr(>F)`,
      Significant = P_value < 0.05
    )

  effects <- bind_rows(effects_ps1, effects_ps2)

  # Factor-specific colors
  factor_colors <- c(
    "Species" = "#009E73",    # Green
    "Waveform" = "#0072B2",   # Blue
    "Sex" = "#E69F00",        # Orange
    "Body size" = "#8C8C8C"   # Gray
  )

  # Order factors by mean R²
  factor_order <- effects %>%
    group_by(Factor) %>%
    summarise(Mean_R2 = mean(R2), .groups = "drop") %>%
    arrange(desc(Mean_R2)) %>%
    pull(Factor)

  effects$Factor <- factor(effects$Factor, levels = factor_order)

  # Paired dot plot
  p <- ggplot(effects, aes(x = R2, y = Factor, color = Factor)) +
    geom_point(
      aes(shape = PrimerSet,
          fill = if_else(Significant, as.character(Factor), NA_character_)),
      size = 4, stroke = 1.2
    ) +
    geom_line(aes(group = Factor), color = "gray70", linewidth = 0.5) +
    scale_shape_manual(values = c(PS1 = 21, PS2 = 23)) +  # circle, diamond
    scale_color_manual(values = factor_colors) +
    scale_fill_manual(values = factor_colors, na.value = "white", guide = "none") +
    labs(x = expression(paste(R^2, " (Marginal Effect Size)")), y = "") +
    theme_minimal(base_size = 9) +
    theme(
      legend.position = "bottom",
      legend.title = element_text(size = 8)
    )

  return(p)
}
```

```{r generate-panels, fig.height=8, fig.width=10}
cat("=== GENERATING 4-PANEL PUBLICATION FIGURE ===\n\n")

# Generate panels
panelA <- create_panel_a_bale(ps1_bale_family, ps2_bale_family, top_families_unified)
panelB <- create_panel_b_bale(comp_ps1, comp_ps2)
panelC <- create_panel_c_bale(comp_ps1)
panelD <- create_panel_d_bale(perm_ps1_marginal, perm_ps2_marginal)

# Assemble with patchwork
fig_main <- (panelA / panelC) | (panelB / panelD) +
  plot_layout(widths = c(2, 1)) +
  plot_annotation(tag_levels = 'A') &
  theme(
    plot.tag = element_text(size = 14, face = "bold"),
    text = element_text(family = "sans", size = 9)
  )

print(fig_main)

# Save
ggsave(
  file.path(root, "output_data/07_Analysis/bale_creek_comprehensive_figure.pdf"),
  fig_main,
  width = 10, height = 8, units = "in", dpi = 300
)

ggsave(
  file.path(root, "output_data/07_Analysis/bale_creek_comprehensive_figure.png"),
  fig_main,
  width = 10, height = 8, units = "in", dpi = 300
)

cat("Figure saved to output_data/07_Analysis/\n")
```

---

# Supplemental Analyses

## PS2 PCoA (Supplemental Figure)

```{r ps2-pcoa-supplemental, fig.height=5, fig.width=7}
pcoa_ps2_supp <- create_panel_c_bale(comp_ps2)

# Update annotation for PS2
pcoa_ps2_supp <- pcoa_ps2_supp +
  labs(title = "PS2: Concordant dietary patterns") +
  annotate("text", x = Inf, y = -Inf,
           label = "Supplemental Figure",
           hjust = 1.1, vjust = -0.5, size = 2.5, fontface = "italic")

print(pcoa_ps2_supp)

# Save
ggsave(
  file.path(root, "output_data/07_Analysis/bale_creek_ps2_pcoa.pdf"),
  pcoa_ps2_supp,
  width = 6, height = 5, units = "in", dpi = 300
)
```

## Dispersion Test Visualizations

```{r dispersion-plots, fig.height=6, fig.width=10}
# Create diagnostic plots for dispersion tests
par(mfrow = c(2, 3))

# Species
plot(disp_species_ps1$betadisper, main = "PS1: Dispersion by Species")
plot(disp_species_ps2$betadisper, main = "PS2: Dispersion by Species")

# Waveform
plot(disp_waveform_ps1$betadisper, main = "PS1: Dispersion by Waveform")
plot(disp_waveform_ps2$betadisper, main = "PS2: Dispersion by Waveform")

# Sex
plot(disp_sex_ps1$betadisper, main = "PS1: Dispersion by Sex")
plot(disp_sex_ps2$betadisper, main = "PS2: Dispersion by Sex")

par(mfrow = c(1, 1))
```

---

# Summary and Interpretation

## Key Findings

### 1. Primary Drivers of Dietary Variation

**Species Identity:**
- R² = `r round(effects_ps1$R2_pct[effects_ps1$Factor == "Species"], 1)`% (PS1), `r round(effects_ps2$R2_pct[effects_ps2$Factor == "Species"], 1)`% (PS2)
- Species is the strongest predictor of dietary composition, indicating significant dietary partitioning among sympatric mormyrid species

**Waveform Type:**
- R² = `r round(effects_ps1$R2_pct[effects_ps1$Factor == "Waveform"], 1)`% (PS1), `r round(effects_ps2$R2_pct[effects_ps2$Factor == "Waveform"], 1)`% (PS2)
- Electrosensory phenotype explains additional variation beyond species identity

**Individual Factors:**
- Sex: R² = `r round(effects_ps1$R2_pct[effects_ps1$Factor == "Sex"], 1)`% (PS1), `r round(effects_ps2$R2_pct[effects_ps2$Factor == "Sex"], 1)`% (PS2)
- Body size: R² = `r round(effects_ps1$R2_pct[effects_ps1$Factor == "Body size"], 1)`% (PS1), `r round(effects_ps2$R2_pct[effects_ps2$Factor == "Body size"], 1)`% (PS2)

### 2. Variance Partitioning Insights

The variance partitioning analysis reveals:
- Shared variance between Species and Waveform indicates potential confounding (some species are monomorphic for waveform type)
- Individual-level factors (Sex, Body size) contribute relatively small but significant unique variance
- Large residual variance suggests unmeasured factors (e.g., microhabitat, temporal variation) also influence diet

### 3. Primer Set Concordance

Both PS1 and PS2 show qualitatively similar patterns:
- Consistent ranking of effect sizes
- Similar PCoA structure
- Concordant prey family compositions

This dual-primer validation strengthens confidence in the biological interpretations.

### 4. Alpha Diversity Patterns

Shannon diversity varies significantly among species, suggesting differential dietary breadth. Some species may be specialists while others are generalists.

### 5. Dispersion Considerations

```{r dispersion-summary}
# Check if any dispersion tests were significant
disp_warnings <- c()
if (disp_species_ps1$test$tab$`Pr(>F)`[1] < 0.05) disp_warnings <- c(disp_warnings, "Species (PS1)")
if (disp_species_ps2$test$tab$`Pr(>F)`[1] < 0.05) disp_warnings <- c(disp_warnings, "Species (PS2)")
if (disp_waveform_ps1$test$tab$`Pr(>F)`[1] < 0.05) disp_warnings <- c(disp_warnings, "Waveform (PS1)")
if (disp_waveform_ps2$test$tab$`Pr(>F)`[1] < 0.05) disp_warnings <- c(disp_warnings, "Waveform (PS2)")
if (disp_sex_ps1$test$tab$`Pr(>F)`[1] < 0.05) disp_warnings <- c(disp_warnings, "Sex (PS1)")
if (disp_sex_ps2$test$tab$`Pr(>F)`[1] < 0.05) disp_warnings <- c(disp_warnings, "Sex (PS2)")

if (length(disp_warnings) > 0) {
  cat("WARNING: Heterogeneous dispersion detected for:\n")
  for (w in disp_warnings) {
    cat("  -", w, "\n")
  }
  cat("\nPERMANOVA results for these factors should be interpreted with caution.\n")
  cat("Differences in dispersion may contribute to significant PERMANOVA results.\n")
} else {
  cat("All dispersion tests non-significant (p > 0.05).\n")
  cat("PERMANOVA assumptions of homogeneous dispersion are met.\n")
}
```

## Biological Interpretation

The Bale Creek mormyrid community shows strong dietary partitioning at the species level, consistent with niche differentiation facilitating coexistence. The additional effect of waveform type suggests that electrosensory phenotype may influence foraging behavior or prey detection beyond species-specific differences. However, caution is warranted in interpreting waveform effects due to potential confounding with species identity (some species are monomorphic).

The relatively small effects of sex and body size suggest that dietary variation in this community is primarily structured by species identity and electrosensory phenotype rather than individual-level factors. This contrasts with some other systems where ontogenetic shifts or sexual dimorphism drive dietary variation.

## Study Limitations

1. **Species-Waveform confounding:** Some species show only one waveform type, making it impossible to fully separate species and waveform effects
2. **Unbalanced sample sizes:** Some species have limited replication
3. **Single locality:** Cannot assess geographic variation
4. **Temporal snapshot:** Seasonal or temporal dietary shifts not captured

---

# Session Information

```{r session-info}
sessionInfo()
```
