---
title: "Comprehensive PERMANOVA: P. kingsleyae Diet Composition"
author: "Gorsuch, Picq, Gallant"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
    keep_tex: false
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
    df_print: paged
knit: (function(inputFile, encoding) { root <- rprojroot::find_root(".git/index"); rmarkdown::render(inputFile, encoding = encoding, output_dir = file.path(root, "output_data/07_Analysis")) })
header-includes:
- \usepackage{longtable}
- \usepackage{booktabs}
- \usepackage{float}
- \usepackage{array}
- \usepackage{pdflscape}
---

```{r setup, include=FALSE}
# Configure knitr
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  dpi = 300
)

# Set root path and source core setup
root <- rprojroot::find_root(".git/index")
source(file.path(root, "code/07_Analysis/core_setup.R"))
```

# Overview

**Biological context:** This analysis combines population-level factors (Sex, body_size, Parasitism) from level3_pk_population_factors with geographic/waveform factors (Region, Waveform) from level4_pk_bp_tp into a single comprehensive PERMANOVA model. The goal is to assess the relative importance of all these factors in explaining dietary variation in *Paramormyrops kingsleyae*.

**Key question:** Understanding whether waveform-based niche partitioning persists after accounting for confounding factors (geography, sex dimorphism, body size, parasitism status).

**Design approach:** Strictly additive PERMANOVA model (Region + Waveform_Type + Sex + body_size + Parasite_status). Uses marginal (Type III) effects to test each factor's independent contribution while controlling for all others.

---

# 1. Data Preparation

```{r data-prep}
cat("=== COMPREHENSIVE DATA PREPARATION ===\n\n")

#' Prepare comprehensive dataset for full PERMANOVA
#'
#' Filters to complete cases (all variables present)
#' Includes ALL fish (juveniles, males, females)
#'
#' @param ps phyloseq object (P. kingsleyae only)
#' @return List with distance matrix, metadata, OTU matrix, filtered phyloseq
prepare_comprehensive_data <- function(ps) {

  # Extract metadata
  meta <- data.frame(sample_data(ps), stringsAsFactors = FALSE)

  cat("Starting samples:", nrow(meta), "\n")

  # Add Region variable (North vs South)
  # North: Mengono Creek, Iboundji Creek
  # South: Bambomo Creek, Ipassa Creek
  meta$Region <- ifelse(
    meta$Locality_Name %in% c("Mengono Creek", "Iboundji Creek"),
    "North", "South"
  )
  meta$Region <- factor(meta$Region, levels = c("North", "South"))

  # Transform Parasitism to Yes/No
  meta$Parasite_status <- ifelse(meta$Parasite == 1, "Yes", "No")
  meta$Parasite_status <- factor(meta$Parasite_status, levels = c("No", "Yes"))

  # Add body_size from SL (standard length)
  meta$body_size <- as.numeric(meta$SL)

  # Ensure Sex is a factor with all levels
  meta$Sex <- factor(meta$Sex, levels = c("Male", "Female", "Juvenile"))

  # Ensure Waveform_Type is a factor
  meta$Waveform_Type <- factor(meta$Waveform_Type, levels = c("BP", "TP"))

  # Filter to complete cases (all variables present)
  complete_cases <- !is.na(meta$Sex) &
                    !is.na(meta$body_size) &
                    !is.na(meta$Parasite) &
                    !is.na(meta$Waveform_Type) &
                    !is.na(meta$Region)

  cat("Complete cases (all variables):", sum(complete_cases), "\n")
  cat("Removed due to missing data:", sum(!complete_cases), "\n")

  # Filter metadata
  meta_filt <- meta[complete_cases, ]

  # Prune phyloseq to matching samples
  ps_filt <- prune_samples(rownames(meta_filt), ps)

  # Remove any taxa not present in these samples
  ps_filt <- prune_taxa(taxa_sums(ps_filt) > 0, ps_filt)

  cat("Filtered phyloseq:", nsamples(ps_filt), "samples,", ntaxa(ps_filt), "ASVs\n")

  # Extract OTU table for beta diversity
  otu_mat <- as.matrix(otu_table(ps_filt))
  if (taxa_are_rows(ps_filt)) {
    otu_mat <- t(otu_mat)
  }

  # Calculate Bray-Curtis distance
  dist_mat <- vegdist(otu_mat, method = "bray")

  # Ensure rownames alignment
  rownames(meta_filt) <- rownames(otu_mat)

  cat("\n")

  return(list(
    distance = dist_mat,
    metadata = meta_filt,
    otu = otu_mat,
    ps_filtered = ps_filt
  ))
}

# Apply to both primer sets
comp_ps1 <- prepare_comprehensive_data(ps1_pk)
comp_ps2 <- prepare_comprehensive_data(ps2_pk)
```

## 1.1 Sample Size Summary

```{r sample-sizes}
cat("=== SAMPLE SIZE SUMMARY ===\n\n")

# Function to create summary tables
summarize_samples <- function(meta, ps_name) {
  cat(ps_name, "\n")
  cat("Total samples:", nrow(meta), "\n\n")

  # By Sex
  cat("By Sex:\n")
  print(table(meta$Sex))
  cat("\n")

  # By Region
  cat("By Region:\n")
  print(table(meta$Region))
  cat("\n")

  # By Waveform
  cat("By Waveform:\n")
  print(table(meta$Waveform_Type))
  cat("\n")

  # By Parasite status
  cat("By Parasite Status:\n")
  print(table(meta$Parasite_status))
  cat("\n")

  # Cross-tabulation: Region × Waveform
  cat("Region × Waveform:\n")
  print(table(meta$Region, meta$Waveform_Type))
  cat("\n")

  # Cross-tabulation: Region × Waveform × Sex
  cat("Region × Waveform × Sex:\n")
  cross_tab <- meta %>%
    group_by(Region, Waveform_Type, Sex) %>%
    summarise(n = n(), .groups = "drop") %>%
    pivot_wider(names_from = Sex, values_from = n, values_fill = 0)
  print(cross_tab)
  cat("\n")

  # Body size summary
  cat("Body size (SL) summary:\n")
  print(summary(meta$body_size))
  cat("\n")

  # Body size by Sex
  cat("Body size by Sex:\n")
  body_by_sex <- meta %>%
    group_by(Sex) %>%
    summarise(
      n = n(),
      mean_SL = mean(body_size, na.rm = TRUE),
      sd_SL = sd(body_size, na.rm = TRUE),
      min_SL = min(body_size, na.rm = TRUE),
      max_SL = max(body_size, na.rm = TRUE),
      .groups = "drop"
    )
  print(body_by_sex)
  cat("\n\n")
}

# Summarize both primer sets
summarize_samples(comp_ps1$metadata, "PRIMERSET 1")
summarize_samples(comp_ps2$metadata, "PRIMERSET 2")
```

**Key observations:**

- Sample includes all life stages: Males, Females, and Juveniles
- Body size and Sex are partially confounded (juveniles smaller, sexual size dimorphism)
- Balanced design for Region × Waveform allows independent testing
- Complete case filtering ensures all PERMANOVA terms can be tested

---

# 2. Comprehensive PERMANOVA Analysis

## 2.1 Primerset 1: Marginal Effects (Additive Model)

```{r ps1-permanova-marginal}
cat("=== PRIMERSET 1: COMPREHENSIVE PERMANOVA ===\n\n")
cat("Model: distance ~ Region + Waveform_Type + Sex + body_size + Parasite_status\n\n")

# PRIMARY MODEL: Marginal tests (additive)
cat("MARGINAL EFFECTS (Type III) - Tests each term independently:\n\n")
perm_ps1_marginal <- adonis2(
  comp_ps1$distance ~ Region + Waveform_Type + Sex + body_size + Parasite_status,
  data = comp_ps1$metadata,
  permutations = 999,
  by = "margin"
)
print(perm_ps1_marginal)

# Extract effect sizes for summary
perm_ps1_df <- as.data.frame(perm_ps1_marginal)
perm_ps1_df$Term <- rownames(perm_ps1_df)

# Remove residual row for effect size ranking
effects_ps1 <- perm_ps1_df %>%
  filter(Term != "Residual") %>%
  arrange(desc(R2)) %>%
  select(Term, Df, SumOfSqs, R2, F, `Pr(>F)`)

cat("\n\nEFFECT SIZE RANKING (by R²):\n")
print(effects_ps1, digits = 4)
```

### Marginal Effects Table

```{r ps1-marginal-table}
kable(effects_ps1,
  digits = 4,
  col.names = c("Term", "Df", "Sum of Squares", "R²", "F", "P-value"),
  caption = "Primerset 1: Marginal PERMANOVA Effects (Type III)",
  booktabs = TRUE
) %>%
  kable_styling(
    latex_options = c("striped", "HOLD_position"),
    full_width = FALSE
  )
```

## 2.2 Primerset 1: Sequential Tests (Term Order Sensitivity)

```{r ps1-permanova-sequential}
cat("\n=== SEQUENTIAL TESTS (Sensitivity to Term Order) ===\n\n")

# Test 1: Geographic/waveform factors first
cat("Geographic/Waveform factors first:\n")
perm_ps1_geo_first <- adonis2(
  comp_ps1$distance ~ Region + Waveform_Type + Sex + body_size + Parasite_status,
  data = comp_ps1$metadata,
  permutations = 999,
  by = "terms"
)
print(perm_ps1_geo_first)

cat("\n\nIndividual factors first:\n")
perm_ps1_ind_first <- adonis2(
  comp_ps1$distance ~ Sex + body_size + Parasite_status + Region + Waveform_Type,
  data = comp_ps1$metadata,
  permutations = 999,
  by = "terms"
)
print(perm_ps1_ind_first)

# Compare R² values
cat("\n\nCOMPARISON: R² values by term order\n")
r2_comparison_ps1 <- data.frame(
  Term = c("Region", "Waveform_Type", "Sex", "body_size", "Parasite_status"),
  Marginal_R2 = perm_ps1_df[1:5, "R2"],
  Geo_First_R2 = as.data.frame(perm_ps1_geo_first)[1:5, "R2"],
  Ind_First_R2 = as.data.frame(perm_ps1_ind_first)[c(4, 5, 1, 2, 3), "R2"]
)
print(r2_comparison_ps1, digits = 4)
```

## 2.3 Primerset 2: Marginal Effects (Additive Model)

```{r ps2-permanova-marginal}
cat("\n\n=== PRIMERSET 2: COMPREHENSIVE PERMANOVA ===\n\n")

# PRIMARY MODEL: Marginal tests (additive)
cat("MARGINAL EFFECTS (Type III):\n\n")
perm_ps2_marginal <- adonis2(
  comp_ps2$distance ~ Region + Waveform_Type + Sex + body_size + Parasite_status,
  data = comp_ps2$metadata,
  permutations = 999,
  by = "margin"
)
print(perm_ps2_marginal)

# Extract effect sizes
perm_ps2_df <- as.data.frame(perm_ps2_marginal)
perm_ps2_df$Term <- rownames(perm_ps2_df)

effects_ps2 <- perm_ps2_df %>%
  filter(Term != "Residual") %>%
  arrange(desc(R2)) %>%
  select(Term, Df, SumOfSqs, R2, F, `Pr(>F)`)

cat("\n\nEFFECT SIZE RANKING (by R²):\n")
print(effects_ps2, digits = 4)
```

### Marginal Effects Table

```{r ps2-marginal-table}
kable(effects_ps2,
  digits = 4,
  col.names = c("Term", "Df", "Sum of Squares", "R²", "F", "P-value"),
  caption = "Primerset 2: Marginal PERMANOVA Effects (Type III)",
  booktabs = TRUE
) %>%
  kable_styling(
    latex_options = c("striped", "HOLD_position"),
    full_width = FALSE
  )
```

## 2.4 Primerset 2: Sequential Tests

```{r ps2-permanova-sequential}
cat("\n=== PRIMERSET 2: SEQUENTIAL TESTS ===\n\n")

cat("Geographic/Waveform factors first:\n")
perm_ps2_geo_first <- adonis2(
  comp_ps2$distance ~ Region + Waveform_Type + Sex + body_size + Parasite_status,
  data = comp_ps2$metadata,
  permutations = 999,
  by = "terms"
)
print(perm_ps2_geo_first)

cat("\n\nIndividual factors first:\n")
perm_ps2_ind_first <- adonis2(
  comp_ps2$distance ~ Sex + body_size + Parasite_status + Region + Waveform_Type,
  data = comp_ps2$metadata,
  permutations = 999,
  by = "terms"
)
print(perm_ps2_ind_first)

# Compare R² values
cat("\n\nCOMPARISON: R² values by term order\n")
r2_comparison_ps2 <- data.frame(
  Term = c("Region", "Waveform_Type", "Sex", "body_size", "Parasite_status"),
  Marginal_R2 = perm_ps2_df[1:5, "R2"],
  Geo_First_R2 = as.data.frame(perm_ps2_geo_first)[1:5, "R2"],
  Ind_First_R2 = as.data.frame(perm_ps2_ind_first)[c(4, 5, 1, 2, 3), "R2"]
)
print(r2_comparison_ps2, digits = 4)
```

---

# 3. Variance Partitioning

## 3.1 Two-Way Partition: Ecological vs. Individual Factors

```{r variance-partition-2way}
cat("=== TWO-WAY VARIANCE PARTITIONING ===\n\n")
cat("Group 1: Ecological (Region + Waveform_Type)\n")
cat("Group 2: Individual (Sex + body_size + Parasite_status)\n\n")

# Primerset 1
cat("PRIMERSET 1:\n")
vp_2way_ps1 <- varpart(
  comp_ps1$distance,
  ~ Region + Waveform_Type,
  ~ Sex + body_size + Parasite_status,
  data = comp_ps1$metadata
)
print(vp_2way_ps1)
plot(vp_2way_ps1, digits = 3,
     Xnames = c("Ecological", "Individual"),
     bg = c("#0072B2", "#D55E00"))
title(main = "Primerset 1: Ecological vs. Individual Variance")

# Extract fractions
vp_2way_ps1_frac <- vp_2way_ps1$part$indfract
cat("\n\nVariance Components (Adjusted R²):\n")
cat(sprintf("  Pure Ecological: %.4f\n", vp_2way_ps1_frac$Adj.R.square[1]))
cat(sprintf("  Pure Individual: %.4f\n", vp_2way_ps1_frac$Adj.R.square[2]))
cat(sprintf("  Shared: %.4f\n", vp_2way_ps1_frac$Adj.R.square[3]))
cat(sprintf("  Unexplained: %.4f\n", 1 - sum(vp_2way_ps1_frac$Adj.R.square[1:3])))

# Primerset 2
cat("\n\nPRIMERSET 2:\n")
vp_2way_ps2 <- varpart(
  comp_ps2$distance,
  ~ Region + Waveform_Type,
  ~ Sex + body_size + Parasite_status,
  data = comp_ps2$metadata
)
print(vp_2way_ps2)
plot(vp_2way_ps2, digits = 3,
     Xnames = c("Ecological", "Individual"),
     bg = c("#0072B2", "#D55E00"))
title(main = "Primerset 2: Ecological vs. Individual Variance")

vp_2way_ps2_frac <- vp_2way_ps2$part$indfract
cat("\n\nVariance Components (Adjusted R²):\n")
cat(sprintf("  Pure Ecological: %.4f\n", vp_2way_ps2_frac$Adj.R.square[1]))
cat(sprintf("  Pure Individual: %.4f\n", vp_2way_ps2_frac$Adj.R.square[2]))
cat(sprintf("  Shared: %.4f\n", vp_2way_ps2_frac$Adj.R.square[3]))
cat(sprintf("  Unexplained: %.4f\n", 1 - sum(vp_2way_ps2_frac$Adj.R.square[1:3])))
```

## 3.2 Three-Way Partition: Region vs. Waveform vs. Individual

```{r variance-partition-3way}
cat("\n\n=== THREE-WAY VARIANCE PARTITIONING ===\n\n")
cat("Group 1: Region\n")
cat("Group 2: Waveform_Type\n")
cat("Group 3: Individual (Sex + body_size + Parasite_status)\n\n")

# Primerset 1
cat("PRIMERSET 1:\n")
vp_3way_ps1 <- varpart(
  comp_ps1$distance,
  ~ Region,
  ~ Waveform_Type,
  ~ Sex + body_size + Parasite_status,
  data = comp_ps1$metadata
)
print(vp_3way_ps1)
plot(vp_3way_ps1, digits = 3,
     Xnames = c("Region", "Waveform", "Individual"),
     bg = c("#009E73", "#0072B2", "#D55E00"))
title(main = "Primerset 1: Three-Way Variance Partition")

vp_3way_ps1_frac <- vp_3way_ps1$part$indfract
cat("\n\nVariance Components (Adjusted R²):\n")
cat(sprintf("  Pure Region: %.4f\n", vp_3way_ps1_frac$Adj.R.square[1]))
cat(sprintf("  Pure Waveform: %.4f\n", vp_3way_ps1_frac$Adj.R.square[2]))
cat(sprintf("  Pure Individual: %.4f\n", vp_3way_ps1_frac$Adj.R.square[3]))
cat(sprintf("  Shared Region+Waveform: %.4f\n", vp_3way_ps1_frac$Adj.R.square[4]))
cat(sprintf("  Shared Region+Individual: %.4f\n", vp_3way_ps1_frac$Adj.R.square[5]))
cat(sprintf("  Shared Waveform+Individual: %.4f\n", vp_3way_ps1_frac$Adj.R.square[6]))
cat(sprintf("  Shared All Three: %.4f\n", vp_3way_ps1_frac$Adj.R.square[7]))

# Primerset 2
cat("\n\nPRIMERSET 2:\n")
vp_3way_ps2 <- varpart(
  comp_ps2$distance,
  ~ Region,
  ~ Waveform_Type,
  ~ Sex + body_size + Parasite_status,
  data = comp_ps2$metadata
)
print(vp_3way_ps2)
plot(vp_3way_ps2, digits = 3,
     Xnames = c("Region", "Waveform", "Individual"),
     bg = c("#009E73", "#0072B2", "#D55E00"))
title(main = "Primerset 2: Three-Way Variance Partition")

vp_3way_ps2_frac <- vp_3way_ps2$part$indfract
cat("\n\nVariance Components (Adjusted R²):\n")
cat(sprintf("  Pure Region: %.4f\n", vp_3way_ps2_frac$Adj.R.square[1]))
cat(sprintf("  Pure Waveform: %.4f\n", vp_3way_ps2_frac$Adj.R.square[2]))
cat(sprintf("  Pure Individual: %.4f\n", vp_3way_ps2_frac$Adj.R.square[3]))
cat(sprintf("  Shared Region+Waveform: %.4f\n", vp_3way_ps2_frac$Adj.R.square[4]))
cat(sprintf("  Shared Region+Individual: %.4f\n", vp_3way_ps2_frac$Adj.R.square[5]))
cat(sprintf("  Shared Waveform+Individual: %.4f\n", vp_3way_ps2_frac$Adj.R.square[6]))
cat(sprintf("  Shared All Three: %.4f\n", vp_3way_ps2_frac$Adj.R.square[7]))
```

### Variance Partitioning Visualization

```{r variance-partition-bars, fig.width=10, fig.height=6}
# Create stacked bar chart for variance components

# Primerset 1 - Two-way
vp_2way_ps1_data <- data.frame(
  Component = c("Pure Ecological", "Pure Individual", "Shared", "Unexplained"),
  Variance = c(
    max(0, vp_2way_ps1_frac$Adj.R.square[1]),
    max(0, vp_2way_ps1_frac$Adj.R.square[2]),
    max(0, vp_2way_ps1_frac$Adj.R.square[3]),
    max(0, 1 - sum(vp_2way_ps1_frac$Adj.R.square[1:3]))
  ),
  PrimerSet = "PS1",
  Partition = "Two-Way"
)

# Primerset 2 - Two-way
vp_2way_ps2_data <- data.frame(
  Component = c("Pure Ecological", "Pure Individual", "Shared", "Unexplained"),
  Variance = c(
    max(0, vp_2way_ps2_frac$Adj.R.square[1]),
    max(0, vp_2way_ps2_frac$Adj.R.square[2]),
    max(0, vp_2way_ps2_frac$Adj.R.square[3]),
    max(0, 1 - sum(vp_2way_ps2_frac$Adj.R.square[1:3]))
  ),
  PrimerSet = "PS2",
  Partition = "Two-Way"
)

# Primerset 1 - Three-way (simplified)
vp_3way_ps1_data <- data.frame(
  Component = c("Pure Region", "Pure Waveform", "Pure Individual", "Shared", "Unexplained"),
  Variance = c(
    max(0, vp_3way_ps1_frac$Adj.R.square[1]),
    max(0, vp_3way_ps1_frac$Adj.R.square[2]),
    max(0, vp_3way_ps1_frac$Adj.R.square[3]),
    max(0, sum(vp_3way_ps1_frac$Adj.R.square[4:7])),
    max(0, 1 - sum(vp_3way_ps1_frac$Adj.R.square[1:7]))
  ),
  PrimerSet = "PS1",
  Partition = "Three-Way"
)

# Primerset 2 - Three-way (simplified)
vp_3way_ps2_data <- data.frame(
  Component = c("Pure Region", "Pure Waveform", "Pure Individual", "Shared", "Unexplained"),
  Variance = c(
    max(0, vp_3way_ps2_frac$Adj.R.square[1]),
    max(0, vp_3way_ps2_frac$Adj.R.square[2]),
    max(0, vp_3way_ps2_frac$Adj.R.square[3]),
    max(0, sum(vp_3way_ps2_frac$Adj.R.square[4:7])),
    max(0, 1 - sum(vp_3way_ps2_frac$Adj.R.square[1:7]))
  ),
  PrimerSet = "PS2",
  Partition = "Three-Way"
)

# Combine
vp_combined <- bind_rows(vp_2way_ps1_data, vp_2way_ps2_data, vp_3way_ps1_data, vp_3way_ps2_data)
vp_combined$Component <- factor(vp_combined$Component,
  levels = c("Pure Region", "Pure Waveform", "Pure Ecological", "Pure Individual", "Shared", "Unexplained"))

# Plot
p_varpart <- ggplot(vp_combined, aes(x = PrimerSet, y = Variance, fill = Component)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ Partition, ncol = 2) +
  scale_fill_manual(values = c(
    "Pure Region" = "#009E73",
    "Pure Waveform" = "#0072B2",
    "Pure Ecological" = "#56B4E9",
    "Pure Individual" = "#D55E00",
    "Shared" = "#E69F00",
    "Unexplained" = "#999999"
  )) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Variance Partitioning: Ecological vs. Individual Factors",
    x = "Primer Set",
    y = "Proportion of Variance",
    fill = "Component"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 12)
  )

print(p_varpart)
```

---

# 4. PCoA Visualization

## 4.1 Multi-Panel PCoA: All Factors

```{r pcoa-all-factors, fig.width=12, fig.height=8}
cat("=== PCOA VISUALIZATION ===\n\n")

# Function to create PCoA from distance matrix
create_pcoa_data <- function(dist_mat, metadata) {
  pcoa_result <- cmdscale(dist_mat, eig = TRUE, k = 2)

  pcoa_df <- data.frame(
    PC1 = pcoa_result$points[, 1],
    PC2 = pcoa_result$points[, 2],
    metadata
  )

  # Calculate variance explained
  eig <- pcoa_result$eig
  var_exp <- eig / sum(eig)

  list(
    data = pcoa_df,
    var_pc1 = var_exp[1],
    var_pc2 = var_exp[2]
  )
}

# Create PCoA data
pcoa_ps1 <- create_pcoa_data(comp_ps1$distance, comp_ps1$metadata)
pcoa_ps2 <- create_pcoa_data(comp_ps2$distance, comp_ps2$metadata)

# Primerset 1 - Multi-panel
p1_wave_region <- ggplot(pcoa_ps1$data, aes(x = PC1, y = PC2, color = Waveform_Type, shape = Region)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = c("BP" = bp_color, "TP" = tp_color)) +
  labs(
    title = "A: Waveform & Region",
    x = sprintf("PC1 (%.1f%%)", pcoa_ps1$var_pc1 * 100),
    y = sprintf("PC2 (%.1f%%)", pcoa_ps1$var_pc2 * 100)
  ) +
  theme_minimal()

p1_sex_parasite <- ggplot(pcoa_ps1$data, aes(x = PC1, y = PC2, color = Sex, shape = Parasite_status)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = c("Male" = "#0072B2", "Female" = "#D55E00", "Juvenile" = "#009E73")) +
  labs(
    title = "B: Sex & Parasitism",
    x = sprintf("PC1 (%.1f%%)", pcoa_ps1$var_pc1 * 100),
    y = sprintf("PC2 (%.1f%%)", pcoa_ps1$var_pc2 * 100)
  ) +
  theme_minimal()

p1_size <- ggplot(pcoa_ps1$data, aes(x = PC1, y = PC2, color = body_size)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_viridis_c(name = "Body Size (mm)") +
  labs(
    title = "C: Body Size",
    x = sprintf("PC1 (%.1f%%)", pcoa_ps1$var_pc1 * 100),
    y = sprintf("PC2 (%.1f%%)", pcoa_ps1$var_pc2 * 100)
  ) +
  theme_minimal()

# Combine panels
p_pcoa_ps1 <- (p1_wave_region | p1_sex_parasite | p1_size) +
  plot_annotation(title = "Primerset 1: PCoA of Diet Composition (Bray-Curtis)")

print(p_pcoa_ps1)

# Primerset 2 - Multi-panel
p2_wave_region <- ggplot(pcoa_ps2$data, aes(x = PC1, y = PC2, color = Waveform_Type, shape = Region)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = c("BP" = bp_color, "TP" = tp_color)) +
  labs(
    title = "A: Waveform & Region",
    x = sprintf("PC1 (%.1f%%)", pcoa_ps2$var_pc1 * 100),
    y = sprintf("PC2 (%.1f%%)", pcoa_ps2$var_pc2 * 100)
  ) +
  theme_minimal()

p2_sex_parasite <- ggplot(pcoa_ps2$data, aes(x = PC1, y = PC2, color = Sex, shape = Parasite_status)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = c("Male" = "#0072B2", "Female" = "#D55E00", "Juvenile" = "#009E73")) +
  labs(
    title = "B: Sex & Parasitism",
    x = sprintf("PC1 (%.1f%%)", pcoa_ps2$var_pc1 * 100),
    y = sprintf("PC2 (%.1f%%)", pcoa_ps2$var_pc2 * 100)
  ) +
  theme_minimal()

p2_size <- ggplot(pcoa_ps2$data, aes(x = PC1, y = PC2, color = body_size)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_viridis_c(name = "Body Size (mm)") +
  labs(
    title = "C: Body Size",
    x = sprintf("PC1 (%.1f%%)", pcoa_ps2$var_pc1 * 100),
    y = sprintf("PC2 (%.1f%%)", pcoa_ps2$var_pc2 * 100)
  ) +
  theme_minimal()

p_pcoa_ps2 <- (p2_wave_region | p2_sex_parasite | p2_size) +
  plot_annotation(title = "Primerset 2: PCoA of Diet Composition (Bray-Curtis)")

print(p_pcoa_ps2)
```

---

# 5. Model Diagnostics

## 5.1 Dispersion Tests (betadisper)

PERMANOVA assumes homogeneous dispersion across groups. We test this assumption for each categorical factor.

```{r betadisper-tests, fig.width=12, fig.height=8}
cat("=== DISPERSION TESTS (betadisper) ===\n\n")

# Function to run betadisper and plot
test_dispersion <- function(dist_mat, metadata, factor_var, ps_name) {
  cat(ps_name, "-", factor_var, "\n")

  # Run betadisper
  bd <- betadisper(dist_mat, metadata[[factor_var]])

  # Permutation test
  bd_test <- permutest(bd, permutations = 999)

  cat("  F =", round(bd_test$statistic, 3), ", p =", round(bd_test$tab$`Pr(>F)`[1], 4), "\n")

  if (bd_test$tab$`Pr(>F)`[1] < 0.05) {
    cat("  WARNING: Significant dispersion heterogeneity detected!\n")
  } else {
    cat("  OK: Homogeneous dispersion (assumption met)\n")
  }
  cat("\n")

  # Create plot
  bd_df <- data.frame(
    Distance_to_Centroid = bd$distances,
    Group = bd$group
  )

  p <- ggplot(bd_df, aes(x = Group, y = Distance_to_Centroid, fill = Group)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.5) +
    labs(
      title = paste(ps_name, ":", factor_var),
      subtitle = sprintf("F = %.2f, p = %.4f", bd_test$statistic, bd_test$tab$`Pr(>F)`[1]),
      x = factor_var,
      y = "Distance to Centroid"
    ) +
    theme_minimal() +
    theme(legend.position = "none")

  list(test = bd_test, plot = p)
}

# Primerset 1 - All factors
cat("PRIMERSET 1:\n\n")
bd_ps1_waveform <- test_dispersion(comp_ps1$distance, comp_ps1$metadata, "Waveform_Type", "PS1")
bd_ps1_region <- test_dispersion(comp_ps1$distance, comp_ps1$metadata, "Region", "PS1")
bd_ps1_sex <- test_dispersion(comp_ps1$distance, comp_ps1$metadata, "Sex", "PS1")
bd_ps1_parasite <- test_dispersion(comp_ps1$distance, comp_ps1$metadata, "Parasite_status", "PS1")

# Combine plots
p_disp_ps1 <- (bd_ps1_waveform$plot | bd_ps1_region$plot) /
              (bd_ps1_sex$plot | bd_ps1_parasite$plot) +
  plot_annotation(title = "Primerset 1: Dispersion Tests (betadisper)")

print(p_disp_ps1)

# Primerset 2 - All factors
cat("\nPRIMERSET 2:\n\n")
bd_ps2_waveform <- test_dispersion(comp_ps2$distance, comp_ps2$metadata, "Waveform_Type", "PS2")
bd_ps2_region <- test_dispersion(comp_ps2$distance, comp_ps2$metadata, "Region", "PS2")
bd_ps2_sex <- test_dispersion(comp_ps2$distance, comp_ps2$metadata, "Sex", "PS2")
bd_ps2_parasite <- test_dispersion(comp_ps2$distance, comp_ps2$metadata, "Parasite_status", "PS2")

p_disp_ps2 <- (bd_ps2_waveform$plot | bd_ps2_region$plot) /
              (bd_ps2_sex$plot | bd_ps2_parasite$plot) +
  plot_annotation(title = "Primerset 2: Dispersion Tests (betadisper)")

print(p_disp_ps2)
```

**Interpretation:** If betadisper tests are significant, PERMANOVA results may be driven by differences in dispersion (spread) rather than centroid location. In such cases, interpret PERMANOVA results with caution.

---

# 6. Summary and Interpretation

```{r summary-interpretation}
cat("=== SUMMARY: COMPREHENSIVE PERMANOVA RESULTS ===\n\n")

# Extract key results for summary
summarize_results <- function(perm_df, ps_name) {
  cat(ps_name, "\n")
  cat(strrep("=", nchar(ps_name) + 1), "\n\n")

  # Significant factors (p < 0.05)
  sig_factors <- perm_df %>%
    filter(Term != "Residual", `Pr(>F)` < 0.05) %>%
    arrange(desc(R2))

  if (nrow(sig_factors) > 0) {
    cat("Significant factors (p < 0.05):\n")
    for (i in 1:nrow(sig_factors)) {
      cat(sprintf("  %d. %s: R² = %.4f, p = %.4f\n",
                  i, sig_factors$Term[i], sig_factors$R2[i], sig_factors$`Pr(>F)`[i]))
    }
  } else {
    cat("No significant factors detected.\n")
  }

  # Effect size ranking
  cat("\nEffect Size Ranking (all terms):\n")
  all_terms <- perm_df %>%
    filter(Term != "Residual") %>%
    arrange(desc(R2))
  for (i in 1:nrow(all_terms)) {
    cat(sprintf("  %d. %s: R² = %.4f\n", i, all_terms$Term[i], all_terms$R2[i]))
  }

  cat("\n\n")
}

summarize_results(perm_ps1_df, "PRIMERSET 1")
summarize_results(perm_ps2_df, "PRIMERSET 2")
```

## 6.1 Key Findings

**Question 1: Does Waveform have an independent effect after controlling for confounders?**

Marginal (Type III) tests show Waveform's unique contribution after accounting for:

- Geography (Region)
- Sexual dimorphism (Sex)
- Body size (body_size)
- Health status (Parasite_status)

Primerset 1 Waveform marginal R² = `r sprintf("%.4f", perm_ps1_df[perm_ps1_df$Term == "Waveform_Type", "R2"])`, p = `r sprintf("%.4f", perm_ps1_df[perm_ps1_df$Term == "Waveform_Type", "Pr(>F)"])`

Primerset 2 Waveform marginal R² = `r sprintf("%.4f", perm_ps2_df[perm_ps2_df$Term == "Waveform_Type", "R2"])`, p = `r sprintf("%.4f", perm_ps2_df[perm_ps2_df$Term == "Waveform_Type", "Pr(>F)"])`

**Question 2: How much variance is uniquely explained by Waveform vs. Region vs. Individual factors?**

Three-way variance partitioning (Primerset 1):

- Pure Region: `r sprintf("%.1f%%", max(0, vp_3way_ps1_frac$Adj.R.square[1]) * 100)`
- Pure Waveform: `r sprintf("%.1f%%", max(0, vp_3way_ps1_frac$Adj.R.square[2]) * 100)`
- Pure Individual: `r sprintf("%.1f%%", max(0, vp_3way_ps1_frac$Adj.R.square[3]) * 100)`

Three-way variance partitioning (Primerset 2):

- Pure Region: `r sprintf("%.1f%%", max(0, vp_3way_ps2_frac$Adj.R.square[1]) * 100)`
- Pure Waveform: `r sprintf("%.1f%%", max(0, vp_3way_ps2_frac$Adj.R.square[2]) * 100)`
- Pure Individual: `r sprintf("%.1f%%", max(0, vp_3way_ps2_frac$Adj.R.square[3]) * 100)`

## 6.2 Biological Interpretation

**Waveform-based niche partitioning hypothesis:**

The comprehensive PERMANOVA model allows us to test whether EOD waveform type associates with dietary differences after accounting for potential confounding factors:

1. **Geography:** Different regions may have different prey communities
2. **Sexual dimorphism:** Males and females may have different foraging behaviors
3. **Body size:** Larger fish may consume different prey
4. **Parasitism:** Parasites may affect foraging efficiency or prey choice

**Evidence for waveform effect:**

- If Waveform shows significant marginal effect → independent dietary niche partitioning
- If Waveform effect is non-significant → dietary differences explained by confounders
- Marginal tests assess independent contribution of each factor in the additive model

**Variance partitioning insights:**

- **Two-way partition:** Shows whether dietary variation is primarily driven by ecological context (geography + waveform) vs. individual phenotype (sex, size, health)
- **Three-way partition:** Decomposes ecological variance to distinguish geographic from waveform-specific effects

**Comparison to previous analyses:**

- Level 3 examined Sex, body_size, and Parasitism independently
- Level 4 examined Region and Waveform effects
- This comprehensive model combines all factors in a single additive framework to assess independent contributions while controlling for all confounders

---

# 7. Conclusions

This comprehensive PERMANOVA analysis provides a complete assessment of factors influencing dietary variation in *Paramormyrops kingsleyae*. By including both ecological (Region, Waveform) and individual (Sex, body_size, Parasitism) factors in a single additive model with marginal tests, we can:

1. **Identify significant predictors** of diet composition
2. **Rank effect sizes** to determine relative importance
3. **Test independent contributions** of each factor while controlling for others
4. **Partition variance** to quantify unique vs. shared contributions
5. **Assess robustness** across two independent primer sets

**Key methodological strengths:**

- Marginal (Type III) tests appropriate for confounded factors
- Additive model assesses independent contribution of each factor
- Variance partitioning provides complementary perspective
- betadisper diagnostics validate PERMANOVA assumptions
- Complete case analysis ensures all factors tested on same samples
- Includes all life stages (juveniles, males, females) for maximum power

**Limitations:**

- Complete case filtering may reduce sample size
- Sex and body_size are partially confounded (juveniles smaller)
- Limited geographic replication (2 regions)
- PERMANOVA sensitive to dispersion heterogeneity if detected

---

# Session Info

```{r session-info}
sessionInfo()
```
