---
title: "LEVEL 4 Supplement: SIMPER and Selectivity Analyses"
author: "Gorsuch, Picq, Gallant"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
    keep_tex: false
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
    df_print: paged
knit: (function(inputFile, encoding) { root <- rprojroot::find_root(".git/index"); rmarkdown::render(inputFile, encoding = encoding, output_dir = file.path(root, "output_data/07_Analysis")) })
header-includes:
- \usepackage{longtable}
- \usepackage{booktabs}
- \usepackage{float}
- \usepackage{array}
- \usepackage{pdflscape}
---

```{r setup, include=FALSE}
# Configure knitr
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  dpi = 300
)

# Set root path and source core setup
root <- rprojroot::find_root(".git/index")
source(file.path(root, "code/07_Analysis/core_setup.R"))
```

# Overview

This document contains focused analyses on:
1. **SIMPER analyses (4.3b-g)**: Identifying prey families that drive waveform and regional dietary differences
2. **Prey selectivity analyses (4.6-4.7)**: Testing whether compositional differences reflect active selection or passive availability

These sections are extracted from the comprehensive Level 4 analysis for detailed examination and manuscript preparation.

---

## Data Preparation

This section creates the data objects needed for SIMPER and selectivity analyses.

```{r data-prep}
cat("=== DATA PREPARATION ===\n")

# Add Region variable to phyloseq objects
add_region_variable <- function(ps) {
  meta <- data.frame(sample_data(ps))

  # Define regions based on locality
  # North: Mengono Creek, Iboundji Creek
  # South: Bambomo Creek, Ipassa Creek
  meta$Region <- ifelse(meta$Locality_Name %in% c("Mengono Creek", "Iboundji Creek"),
    "North", "South"
  )

  # Update sample data
  sample_data(ps) <- meta
  return(ps)
}

ps1_pk <- add_region_variable(ps1_pk)
ps2_pk <- add_region_variable(ps2_pk)

cat("Region variable added to phyloseq objects.\n")

# Create distance matrices and metadata for beta diversity
# PS1
otu1_pk <- as.matrix(otu_table(ps1_pk))
if (taxa_are_rows(ps1_pk)) {
  otu1_pk <- t(otu1_pk)
}

meta1_pk_beta <- data.frame(sample_data(ps1_pk), stringsAsFactors = FALSE)
meta1_pk_beta$Region <- factor(
  ifelse(meta1_pk_beta$Locality_Name %in% c("Mengono Creek", "Iboundji Creek"),
    "North", "South"
  ),
  levels = c("North", "South")
)
meta1_pk_beta$Waveform_Type <- factor(meta1_pk_beta$Waveform_Type)
rownames(meta1_pk_beta) <- rownames(otu1_pk)
dist1_pk <- vegdist(otu1_pk, method = "bray")

cat("PS1: Distance matrix created (Bray-Curtis)\n")

# PS2
otu2_pk <- as.matrix(otu_table(ps2_pk))
if (taxa_are_rows(ps2_pk)) {
  otu2_pk <- t(otu2_pk)
}

meta2_pk_beta <- data.frame(sample_data(ps2_pk), stringsAsFactors = FALSE)
meta2_pk_beta$Region <- factor(
  ifelse(meta2_pk_beta$Locality_Name %in% c("Mengono Creek", "Iboundji Creek"),
    "North", "South"
  ),
  levels = c("North", "South")
)
meta2_pk_beta$Waveform_Type <- factor(meta2_pk_beta$Waveform_Type)
rownames(meta2_pk_beta) <- rownames(otu2_pk)
dist2_pk <- vegdist(otu2_pk, method = "bray")

cat("PS2: Distance matrix created (Bray-Curtis)\n")

cat("\nData preparation complete.\n")
cat("  PS1: ", nrow(otu1_pk), " samples, ", ncol(otu1_pk), " ASVs\n", sep = "")
cat("  PS2: ", nrow(otu2_pk), " samples, ", ncol(otu2_pk), " ASVs\n", sep = "")
```

---

## 4.3b SIMPER Analysis: Waveform Differences (BP vs TP)

Identify which prey families contribute most to dietary dissimilarity between BP and TP waveforms.

```{r level4-simper-waveform, fig.width=10, fig.height=8}
cat("=== LEVEL 4.3b: SIMPER ANALYSIS - WAVEFORM DIFFERENCES ===\n")
cat("Identifying prey families driving BP vs. TP dissimilarity\n\n")

run_simper_waveform <- function(ps, primerset_name) {
  cat("\n", primerset_name, ":\n", sep = "")

  # Get OTU table (samples as rows, taxa as columns)
  otu_mat <- as.matrix(otu_table(ps))
  if (taxa_are_rows(ps)) {
    otu_mat <- t(otu_mat)
  }

  # Get metadata
  meta <- data.frame(sample_data(ps))

  # Check if there are at least 2 different waveform groups
  n_groups <- length(unique(meta$Waveform_Type))
  cat("Number of unique waveform groups:", n_groups, "\n")

  if (n_groups < 2) {
    cat("Insufficient waveform groups for SIMPER analysis (need at least 2)\n")
    cat("Waveform groups present:", paste(unique(meta$Waveform_Type), collapse = ", "), "\n\n")
    return(NULL)
  }

  # Run SIMPER
  simper_result <- vegan::simper(otu_mat, meta$Waveform_Type, permutations = 999)

  # Extract first comparison (BP_TP or TP_BP)
  comparison_name <- names(simper_result)[1]
  comp <- simper_result[[comparison_name]]

  cat(
    "Overall dissimilarity between waveforms:",
    round(comp$overall, 3), "\n\n"
  )

  # Build data frame from simper result (ASV level)
  comparison_data <- data.frame(
    species = comp$species,
    average = comp$average,
    sd = comp$sd,
    ratio = comp$ratio,
    ava = comp$ava,
    avb = comp$avb,
    stringsAsFactors = FALSE
  )

  # Calculate contribution percentage
  comparison_data$contr <- comparison_data$average / comp$overall

  # Sort by contribution
  comparison_data <- comparison_data[order(-comparison_data$contr), ]

  # Calculate cumulative sum
  comparison_data$cumsum <- cumsum(comparison_data$contr)

  # Get top contributors (cumulative < 70%)
  top_asvs <- comparison_data[comparison_data$cumsum <= 0.70, ]
  if (nrow(top_asvs) == 0) {
    top_asvs <- head(comparison_data, 15)
  }

  # Get taxonomy
  tax_df <- as.data.frame(tax_table(ps))
  top_asvs$ASV <- top_asvs$species

  # Add Family names
  top_asvs$Family <- sapply(top_asvs$ASV, function(asv) {
    if (asv %in% rownames(tax_df)) {
      fam <- tax_df[asv, "Family"]
      if (is.na(fam) || fam == "") {
        return(tax_df[asv, "Order"])
      }
      return(clean_family_name(fam))
    }
    return("Unknown")
  })

  # Aggregate by Family
  family_aggregated <- top_asvs %>%
    group_by(Family) %>%
    summarise(
      contr_pct = 100 * sum(contr),
      n_asvs = n(),
      .groups = "drop"
    ) %>%
    arrange(desc(contr_pct))

  # Print table
  print(kable(family_aggregated,
    digits = 2,
    caption = paste0(
      "Top prey families driving waveform differences (",
      primerset_name, ")"
    ),
    col.names = c("Family", "Contribution (%)", "N ASVs"),
    booktabs = TRUE,
    row.names = FALSE
  ) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover"),
      full_width = FALSE
    ))

  # Create bar plot
  family_aggregated$Family <- factor(family_aggregated$Family,
    levels = rev(family_aggregated$Family)
  )

  p <- ggplot(family_aggregated, aes(x = contr_pct, y = Family)) +
    geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
    theme_minimal() +
    labs(
      title = paste0(
        "Top Prey Families Driving Waveform Differences (",
        primerset_name, ")"
      ),
      x = "Contribution to Dissimilarity (%)",
      y = "Prey Family"
    )

  return(list(simper = simper_result, plot = p, table = family_aggregated))
}

# Run SIMPER for both primer sets
simper_waveform_ps1 <- run_simper_waveform(ps1_pk, "PRIMERSET 1")
simper_waveform_ps2 <- run_simper_waveform(ps2_pk, "PRIMERSET 2")

# Plot results
if (!is.null(simper_waveform_ps1)) {
  print(simper_waveform_ps1$plot)
}
if (!is.null(simper_waveform_ps2)) {
  print(simper_waveform_ps2$plot)
}
```

## 4.3c SIMPER Analysis: Region Differences (North vs South)

Identify which prey families contribute most to dietary dissimilarity between North and South regions.

```{r level4-simper-region, fig.width=10, fig.height=8}
cat("=== LEVEL 4.3c: SIMPER ANALYSIS - REGION DIFFERENCES ===\n")
cat("Identifying prey families driving North vs. South dissimilarity\n\n")

run_simper_region <- function(ps, primerset_name) {
  cat("\n", primerset_name, ":\n", sep = "")

  # Get OTU table (samples as rows, taxa as columns)
  otu_mat <- as.matrix(otu_table(ps))
  if (taxa_are_rows(ps)) {
    otu_mat <- t(otu_mat)
  }

  # Get metadata
  meta <- data.frame(sample_data(ps))

  # Check if there are at least 2 different region groups
  n_groups <- length(unique(meta$Region))
  cat("Number of unique region groups:", n_groups, "\n")

  if (n_groups < 2) {
    cat("Insufficient region groups for SIMPER analysis (need at least 2)\n")
    cat("Region groups present:", paste(unique(meta$Region), collapse = ", "), "\n\n")
    return(NULL)
  }

  # Run SIMPER
  simper_result <- vegan::simper(otu_mat, meta$Region, permutations = 999)

  # Extract first comparison (North_South or South_North)
  comparison_name <- names(simper_result)[1]
  comp <- simper_result[[comparison_name]]

  cat(
    "Overall dissimilarity between regions:",
    round(comp$overall, 3), "\n\n"
  )

  # Build data frame from simper result (ASV level)
  comparison_data <- data.frame(
    species = comp$species,
    average = comp$average,
    sd = comp$sd,
    ratio = comp$ratio,
    ava = comp$ava,
    avb = comp$avb,
    stringsAsFactors = FALSE
  )

  # Calculate contribution percentage
  comparison_data$contr <- comparison_data$average / comp$overall

  # Sort by contribution
  comparison_data <- comparison_data[order(-comparison_data$contr), ]

  # Calculate cumulative sum
  comparison_data$cumsum <- cumsum(comparison_data$contr)

  # Get top contributors (cumulative < 70%)
  top_asvs <- comparison_data[comparison_data$cumsum <= 0.70, ]
  if (nrow(top_asvs) == 0) {
    top_asvs <- head(comparison_data, 15)
  }

  # Get taxonomy
  tax_df <- as.data.frame(tax_table(ps))
  top_asvs$ASV <- top_asvs$species

  # Add Family names
  top_asvs$Family <- sapply(top_asvs$ASV, function(asv) {
    if (asv %in% rownames(tax_df)) {
      fam <- tax_df[asv, "Family"]
      if (is.na(fam) || fam == "") {
        return(tax_df[asv, "Order"])
      }
      return(clean_family_name(fam))
    }
    return("Unknown")
  })

  # Aggregate by Family
  family_aggregated <- top_asvs %>%
    group_by(Family) %>%
    summarise(
      contr_pct = 100 * sum(contr),
      n_asvs = n(),
      .groups = "drop"
    ) %>%
    arrange(desc(contr_pct))

  # Print table
  print(kable(family_aggregated,
    digits = 2,
    caption = paste0(
      "Top prey families driving region differences (",
      primerset_name, ")"
    ),
    col.names = c("Family", "Contribution (%)", "N ASVs"),
    booktabs = TRUE,
    row.names = FALSE
  ) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover"),
      full_width = FALSE
    ))

  # Create bar plot
  family_aggregated$Family <- factor(family_aggregated$Family,
    levels = rev(family_aggregated$Family)
  )

  p <- ggplot(family_aggregated, aes(x = contr_pct, y = Family)) +
    geom_bar(stat = "identity", fill = "darkgreen", alpha = 0.7) +
    theme_minimal() +
    labs(
      title = paste0(
        "Top Prey Families Driving Region Differences (",
        primerset_name, ")"
      ),
      x = "Contribution to Dissimilarity (%)",
      y = "Prey Family"
    )

  return(list(simper = simper_result, plot = p, table = family_aggregated))
}

# Run SIMPER for both primer sets
simper_region_ps1 <- run_simper_region(ps1_pk, "PRIMERSET 1")
simper_region_ps2 <- run_simper_region(ps2_pk, "PRIMERSET 2")

# Plot results
if (!is.null(simper_region_ps1)) {
  print(simper_region_ps1$plot)
}
if (!is.null(simper_region_ps2)) {
  print(simper_region_ps2$plot)
}
```

## SIMPER Summary Table

Consolidated summary of families contributing > 5% to dissimilarity across both comparisons (BP vs TP, North vs South) and both primersets.

```{r simper-summary-table}
cat("=== SIMPER SUMMARY TABLE ===\n")
cat("Families contributing > 2.5% to dissimilarity\n\n")

# Create a list to store results
summary_list <- list()

# Process PS1 waveform results
if (!is.null(simper_waveform_ps1) && !is.null(simper_waveform_ps1$table)) {
  ps1_wf <- simper_waveform_ps1$table %>%
    filter(contr_pct > 2.5) %>%
    mutate(
      Comparison = "BP vs TP",
      Primerset = "PS1",
      Contribution = contr_pct
    ) %>%
    select(Family, Comparison, Primerset, Contribution)
  summary_list[[1]] <- ps1_wf
}

# Process PS2 waveform results
if (!is.null(simper_waveform_ps2) && !is.null(simper_waveform_ps2$table)) {
  ps2_wf <- simper_waveform_ps2$table %>%
    filter(contr_pct > 2.5) %>%
    mutate(
      Comparison = "BP vs TP",
      Primerset = "PS2",
      Contribution = contr_pct
    ) %>%
    select(Family, Comparison, Primerset, Contribution)
  summary_list[[2]] <- ps2_wf
}

# Process PS1 region results
if (!is.null(simper_region_ps1) && !is.null(simper_region_ps1$table)) {
  ps1_reg <- simper_region_ps1$table %>%
    filter(contr_pct > 2.5) %>%
    mutate(
      Comparison = "North vs South",
      Primerset = "PS1",
      Contribution = contr_pct
    ) %>%
    select(Family, Comparison, Primerset, Contribution)
  summary_list[[3]] <- ps1_reg
}

# Process PS2 region results
if (!is.null(simper_region_ps2) && !is.null(simper_region_ps2$table)) {
  ps2_reg <- simper_region_ps2$table %>%
    filter(contr_pct > 2.5) %>%
    mutate(
      Comparison = "North vs South",
      Primerset = "PS2",
      Contribution = contr_pct
    ) %>%
    select(Family, Comparison, Primerset, Contribution)
  summary_list[[4]] <- ps2_reg
}

# Combine all results
if (length(summary_list) > 0) {
  simper_summary <- bind_rows(summary_list) %>%
    arrange(Comparison, Primerset, desc(Contribution))

  # Print the table
  print(kable(simper_summary,
    digits = 2,
    caption = "Summary of SIMPER Analysis: Families Contributing > 5% to Dissimilarity",
    col.names = c("Family", "Comparison", "Primerset", "Contribution (%)"),
    booktabs = TRUE,
    row.names = FALSE
  ) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover"),
      full_width = FALSE
    ))
} else {
  cat("No SIMPER results available for summary table.\n")
}
```

## 4.3d Mechanistic Drill-Down: Waveform Differences

Partition waveform-related dissimilarity into turnover (species replacement) vs. abundance (differential abundance) components for each contributing family.

```{r level4-waveform-drilldown, fig.width=12, fig.height=10}
cat("=== LEVEL 4.3d: MECHANISTIC DRILL-DOWN - WAVEFORM DIFFERENCES ===\n")
cat("Partitioning dissimilarity into turnover vs. abundance components\n\n")

# Check if SIMPER results exist
if (!is.null(simper_waveform_ps1) && !is.null(simper_waveform_ps1$table)) {
  cat("=== PRIMERSET 1: WAVEFORM DRILL-DOWN ===\n")

  # Extract SIMPER table
  simper_waveform_ps1_table <- simper_waveform_ps1$table

  # Run drill-down
  drilldown_waveform_ps1 <- run_multi_family_drilldown(
    ps = ps1_pk,
    simper_table = simper_waveform_ps1_table,
    group_var = "Waveform_Type",
    group1 = "BP",
    group2 = "TP",
    primerset_name = "PRIMERSET 1",
    contr_threshold = 5,
    max_families = 10
  )

  # Generate summary
  if (length(drilldown_waveform_ps1) > 0) {
    summary_waveform_ps1 <- summarize_drilldown_results(drilldown_waveform_ps1)

    cat("\n\nSummary Table:\n")
    print(kable(summary_waveform_ps1,
      digits = 2,
      caption = "PRIMERSET 1: Waveform Differences - Turnover vs Abundance",
      booktabs = TRUE
    ) %>%
      kable_styling(
        bootstrap_options = c("striped", "hover"),
        full_width = FALSE,
        font_size = 9
      ))

    # Plot summary
    p1 <- plot_drilldown_summary(
      summary_waveform_ps1,
      "PRIMERSET 1: Waveform Differences (BP vs TP)"
    )
    print(p1)
  }
} else {
  cat("PRIMERSET 1: SIMPER results not available for waveform drill-down\n")
}

# Primerset 2
if (!is.null(simper_waveform_ps2) && !is.null(simper_waveform_ps2$table)) {
  cat("\n\n=== PRIMERSET 2: WAVEFORM DRILL-DOWN ===\n")

  simper_waveform_ps2_table <- simper_waveform_ps2$table

  drilldown_waveform_ps2 <- run_multi_family_drilldown(
    ps = ps2_pk,
    simper_table = simper_waveform_ps2_table,
    group_var = "Waveform_Type",
    group1 = "BP",
    group2 = "TP",
    primerset_name = "PRIMERSET 2",
    contr_threshold = 5,
    max_families = 10
  )

  if (length(drilldown_waveform_ps2) > 0) {
    summary_waveform_ps2 <- summarize_drilldown_results(drilldown_waveform_ps2)

    cat("\n\nSummary Table:\n")
    print(kable(summary_waveform_ps2,
      digits = 2,
      caption = "PRIMERSET 2: Waveform Differences - Turnover vs Abundance",
      booktabs = TRUE
    ) %>%
      kable_styling(
        bootstrap_options = c("striped", "hover"),
        full_width = FALSE,
        font_size = 9
      ))

    p2 <- plot_drilldown_summary(
      summary_waveform_ps2,
      "PRIMERSET 2: Waveform Differences (BP vs TP)"
    )
    print(p2)
  }
} else {
  cat("PRIMERSET 2: SIMPER results not available for waveform drill-down\n")
}
```

## 4.3e Mechanistic Drill-Down: Region Differences

Partition region-related dissimilarity into turnover vs. abundance components for each contributing family.

```{r level4-region-drilldown, fig.width=12, fig.height=10}
cat("=== LEVEL 4.3e: MECHANISTIC DRILL-DOWN - REGION DIFFERENCES ===\n")
cat("Partitioning dissimilarity into turnover vs. abundance components\n\n")

# Check if SIMPER results exist
if (!is.null(simper_region_ps1) && !is.null(simper_region_ps1$table)) {
  cat("=== PRIMERSET 1: REGION DRILL-DOWN ===\n")

  # Extract SIMPER table
  simper_region_ps1_table <- simper_region_ps1$table

  # Run drill-down
  drilldown_region_ps1 <- run_multi_family_drilldown(
    ps = ps1_pk,
    simper_table = simper_region_ps1_table,
    group_var = "Region",
    group1 = "North",
    group2 = "South",
    primerset_name = "PRIMERSET 1",
    contr_threshold = 5,
    max_families = 10
  )

  # Generate summary
  if (length(drilldown_region_ps1) > 0) {
    summary_region_ps1 <- summarize_drilldown_results(drilldown_region_ps1)

    cat("\n\nSummary Table:\n")
    print(kable(summary_region_ps1,
      digits = 2,
      caption = "PRIMERSET 1: Region Differences - Turnover vs Abundance",
      booktabs = TRUE
    ) %>%
      kable_styling(
        bootstrap_options = c("striped", "hover"),
        full_width = FALSE,
        font_size = 9
      ))

    # Plot summary
    p1 <- plot_drilldown_summary(
      summary_region_ps1,
      "PRIMERSET 1: Region Differences (North vs South)"
    )
    print(p1)
  }
} else {
  cat("PRIMERSET 1: SIMPER results not available for region drill-down\n")
}

# Primerset 2
if (!is.null(simper_region_ps2) && !is.null(simper_region_ps2$table)) {
  cat("\n\n=== PRIMERSET 2: REGION DRILL-DOWN ===\n")

  simper_region_ps2_table <- simper_region_ps2$table

  drilldown_region_ps2 <- run_multi_family_drilldown(
    ps = ps2_pk,
    simper_table = simper_region_ps2_table,
    group_var = "Region",
    group1 = "North",
    group2 = "South",
    primerset_name = "PRIMERSET 2",
    contr_threshold = 5,
    max_families = 10
  )

  if (length(drilldown_region_ps2) > 0) {
    summary_region_ps2 <- summarize_drilldown_results(drilldown_region_ps2)

    cat("\n\nSummary Table:\n")
    print(kable(summary_region_ps2,
      digits = 2,
      caption = "PRIMERSET 2: Region Differences - Turnover vs Abundance",
      booktabs = TRUE
    ) %>%
      kable_styling(
        bootstrap_options = c("striped", "hover"),
        full_width = FALSE,
        font_size = 9
      ))

    p2 <- plot_drilldown_summary(
      summary_region_ps2,
      "PRIMERSET 2: Region Differences (North vs South)"
    )
    print(p2)
  }
} else {
  cat("PRIMERSET 2: SIMPER results not available for region drill-down\n")
}
```

## 4.3f ASV-Level Visualization: Waveform Differences

Two-panel figure showing ASV-level patterns driving waveform differences.

```{r level4-waveform-asv-viz, fig.width=14, fig.height=7}
cat("=== LEVEL 4.3f: ASV-LEVEL VISUALIZATION - WAVEFORM DIFFERENCES ===\n\n")

# PRIMERSET 1
if (!is.null(simper_waveform_ps1) && !is.null(simper_waveform_ps1$simper)) {
  cat("Creating two-panel ASV figure for PRIMERSET 1...\n")

  comp_name_ps1 <- names(simper_waveform_ps1$simper)[1]

  fig_waveform_ps1 <- create_asv_twopanel_figure(
    ps = ps1_pk,
    simper_result = simper_waveform_ps1$simper,
    comparison_name = comp_name_ps1,
    group_var = "Waveform_Type",
    group1 = "BP",
    group2 = "TP",
    main_title = "PRIMERSET 1: Waveform Differences at ASV Level",
    top_n = 15
  )

  print(fig_waveform_ps1)
} else {
  cat("PRIMERSET 1: SIMPER results not available for ASV visualization\n")
}

# PRIMERSET 2
if (!is.null(simper_waveform_ps2) && !is.null(simper_waveform_ps2$simper)) {
  cat("\n\nCreating two-panel ASV figure for PRIMERSET 2...\n")

  comp_name_ps2 <- names(simper_waveform_ps2$simper)[1]

  fig_waveform_ps2 <- create_asv_twopanel_figure(
    ps = ps2_pk,
    simper_result = simper_waveform_ps2$simper,
    comparison_name = comp_name_ps2,
    group_var = "Waveform_Type",
    group1 = "BP",
    group2 = "TP",
    main_title = "PRIMERSET 2: Waveform Differences at ASV Level",
    top_n = 15
  )

  print(fig_waveform_ps2)
} else {
  cat("PRIMERSET 2: SIMPER results not available for ASV visualization\n")
}
```

## 4.3g ASV-Level Visualization: Region Differences

Two-panel figure showing ASV-level patterns driving region differences.

```{r level4-region-asv-viz, fig.width=14, fig.height=7}
cat("=== LEVEL 4.3g: ASV-LEVEL VISUALIZATION - REGION DIFFERENCES ===\n\n")

# PRIMERSET 1
if (!is.null(simper_region_ps1) && !is.null(simper_region_ps1$simper)) {
  cat("Creating two-panel ASV figure for PRIMERSET 1...\n")

  comp_name_ps1 <- names(simper_region_ps1$simper)[1]

  fig_region_ps1 <- create_asv_twopanel_figure(
    ps = ps1_pk,
    simper_result = simper_region_ps1$simper,
    comparison_name = comp_name_ps1,
    group_var = "Region",
    group1 = "North",
    group2 = "South",
    main_title = "PRIMERSET 1: Region Differences at ASV Level",
    top_n = 15
  )

  print(fig_region_ps1)
} else {
  cat("PRIMERSET 1: SIMPER results not available for ASV visualization\n")
}

# PRIMERSET 2
if (!is.null(simper_region_ps2) && !is.null(simper_region_ps2$simper)) {
  cat("\n\nCreating two-panel ASV figure for PRIMERSET 2...\n")

  comp_name_ps2 <- names(simper_region_ps2$simper)[1]

  fig_region_ps2 <- create_asv_twopanel_figure(
    ps = ps2_pk,
    simper_result = simper_region_ps2$simper,
    comparison_name = comp_name_ps2,
    group_var = "Region",
    group1 = "North",
    group2 = "South",
    main_title = "PRIMERSET 2: Region Differences at ASV Level",
    top_n = 15
  )

  print(fig_region_ps2)
} else {
  cat("PRIMERSET 2: SIMPER results not available for ASV visualization\n")
}
```

---

## 4.6 Prey Selectivity Analysis

Integrate environmental prey availability data to distinguish between:
1. **Compositional differences**: BP and TP eat different prey
2. **Selectivity differences**: BP and TP show different preferences for the same available prey

This requires D-frame net data from the field sites. If available, calculate Ivlev's electivity index (E) and test whether selectivity profiles differ between waveforms.

```{r level4-selectivity, fig.width=12, fig.height=10}
cat("=== LEVEL 4.6: PREY SELECTIVITY ANALYSIS ===\n\n")

# Check if environmental data is available
env_file <- file.path(root, "input_data", "gabon_inverts_compiled_cleaned_2026_final.csv")

if (file.exists(env_file)) {
  cat("Environmental prey availability data found.\n")
  cat("Loading and processing...\n\n")

  # Load environmental data
  env_raw <- read.csv(env_file, stringsAsFactors = FALSE)
  colnames(env_raw) <- trimws(colnames(env_raw))

  # Clean and aggregate environmental data
  env_data <- env_raw %>%
    mutate(Count = as.numeric(Count)) %>%
    mutate(Family_clean = sapply(Family, clean_family_name)) %>%
    filter(!is.na(Family_clean), !is.na(Count), Count > 0) %>%
    select(Site = Name, Family = Family_clean, Count)

  # Map locality names to site names in environmental data
  # This mapping depends on your data - adjust as needed
  locality_to_site <- c(
    "Bambomo" = "Bambomo Creek",
    "Apassa" = "Ipassa Creek",
    "Mengono" = "Mengono Creek",
    "Iboundji" = "Iboundji Creek"
  )

  # Get family-level gut contents for P. kingsleyae
  ps1_pk_family_counts <- extract_family_counts(ps1_pk)
  ps2_pk_family_counts <- extract_family_counts(ps2_pk)

  # Add metadata
  meta1_sel <- data.frame(sample_data(ps1_pk))
  meta1_sel$SampleID <- rownames(meta1_sel)

  meta2_sel <- data.frame(sample_data(ps2_pk))
  meta2_sel$SampleID <- rownames(meta2_sel)

  ps1_pk_family_counts <- ps1_pk_family_counts %>%
    left_join(
      meta1_sel %>%
        select(SampleID, Locality_Name, Waveform_Type, Region),
      by = "SampleID"
    )

  ps2_pk_family_counts <- ps2_pk_family_counts %>%
    left_join(
      meta2_sel %>%
        select(SampleID, Locality_Name, Waveform_Type, Region),
      by = "SampleID"
    )

  # Function to calculate selectivity for an individual fish
  calc_individual_selectivity <- function(sample_id, diet_data, env_data_site, metadata) {
    # Get diet for this individual fish
    diet_individual <- diet_data %>%
      filter(SampleID == sample_id)

    if (nrow(diet_individual) == 0) {
      return(NULL)
    }

    # Calculate selectivity
    selectivity <- calculate_selectivity(
      diet_counts = diet_individual %>% select(Family, Count),
      env_counts = env_data_site %>% select(Family, Count)
    )

    # Add metadata
    meta_row <- metadata[metadata$SampleID == sample_id, ]
    selectivity$SampleID <- sample_id
    selectivity$Locality <- meta_row$Locality_Name
    selectivity$Waveform <- meta_row$Waveform_Type
    selectivity$Region <- meta_row$Region

    return(selectivity)
  }

  # Calculate selectivity for EACH INDIVIDUAL FISH
  selectivity_results_ps1 <- list()
  selectivity_results_ps2 <- list()

  # Get unique samples with locality info
  samples_ps1 <- ps1_pk_family_counts %>%
    distinct(SampleID, Locality_Name, Waveform_Type, Region)

  for (i in 1:nrow(samples_ps1)) {
    sample_id <- samples_ps1$SampleID[i]
    locality <- samples_ps1$Locality_Name[i]

    # Get site name for environmental data
    site_name <- locality_to_site[locality]
    if (is.na(site_name)) next

    # Get environmental data for this site
    env_site <- env_data %>% filter(Site == site_name)
    if (nrow(env_site) == 0) next

    # Calculate selectivity for this individual
    selectivity <- calc_individual_selectivity(
      sample_id = sample_id,
      diet_data = ps1_pk_family_counts,
      env_data_site = env_site,
      metadata = meta1_sel
    )

    if (!is.null(selectivity)) {
      selectivity_results_ps1[[sample_id]] <- selectivity
    }
  }

  # Combine results
  if (length(selectivity_results_ps1) > 0) {
    selectivity_ps1_df <- bind_rows(selectivity_results_ps1)

    cat("PRIMERSET 1 - Selectivity Analysis (Individual-level):\n")
    cat("  Individuals analyzed:", n_distinct(selectivity_ps1_df$SampleID), "\n")
    cat("  Total family-individual combinations:", nrow(selectivity_ps1_df), "\n\n")

    # Summary statistics by Waveform
    cat("Mean Ivlev's E by Waveform:\n")
    selectivity_summary_ps1 <- selectivity_ps1_df %>%
      group_by(Waveform, Region) %>%
      summarise(
        n_individuals = n_distinct(SampleID),
        n_families = n_distinct(Family),
        mean_E = mean(Ivlev_E, na.rm = TRUE),
        mean_abs_E = mean(abs(Ivlev_E), na.rm = TRUE),
        n_preference = sum(Selection_Type == "Preference"),
        n_avoidance = sum(Selection_Type == "Avoidance"),
        .groups = "drop"
      )

    print(selectivity_summary_ps1)

    # Test: Do BP and TP differ in selectivity profiles?
    # Create wide matrix of Ivlev's E values (one row per individual)
    selectivity_wide_ps1 <- selectivity_ps1_df %>%
      select(SampleID, Region, Waveform, Locality, Family, Ivlev_E) %>%
      group_by(SampleID, Region, Waveform, Locality, Family) %>%
      summarise(Ivlev_E = mean(Ivlev_E, na.rm = TRUE), .groups = "drop") %>%
      pivot_wider(names_from = Family, values_from = Ivlev_E, values_fill = 0)

    if (nrow(selectivity_wide_ps1) > 3) { # Need at least 4 samples
      selectivity_matrix_ps1 <- as.matrix(selectivity_wide_ps1[, -(1:4)])
      rownames(selectivity_matrix_ps1) <- selectivity_wide_ps1$SampleID

      selectivity_meta_ps1 <- selectivity_wide_ps1 %>%
        select(SampleID, Region, Waveform, Locality)

      # Distance matrix
      selectivity_dist_ps1 <- vegdist(selectivity_matrix_ps1, method = "euclidean")

      # PERMANOVA: Does waveform predict selectivity?
      cat("\n\nPERMANOVA: Individual Selectivity ~ Waveform + Region\n")
      cat("Sample size: ", nrow(selectivity_meta_ps1), " individuals\n")

      if (n_distinct(selectivity_meta_ps1$Waveform) > 1 &&
        n_distinct(selectivity_meta_ps1$Region) > 0) {
        # Check degrees of freedom
        n_samples <- nrow(selectivity_meta_ps1)
        n_waveforms <- n_distinct(selectivity_meta_ps1$Waveform)
        n_regions <- n_distinct(selectivity_meta_ps1$Region)
        model_df <- (n_waveforms - 1) + (n_regions - 1)
        residual_df <- n_samples - model_df - 1

        cat("Model df:", model_df, ", Residual df:", residual_df, "\n\n")

        if (residual_df > 0) {
          perm_selectivity_ps1 <- adonis2(
            selectivity_dist_ps1 ~ Waveform + Region,
            data = selectivity_meta_ps1,
            permutations = 999,
            by = "margin"
          )
          print(perm_selectivity_ps1)

          cat("\nInterpretation:\n")
          if (!is.na(perm_selectivity_ps1$`Pr(>F)`[1]) &&
            perm_selectivity_ps1$`Pr(>F)`[1] < 0.05) {
            cat("Waveform DOES predict selectivity (p < 0.05)\n")
            cat("→ BP and TP show different prey preferences\n")
          } else {
            cat("Waveform does NOT predict selectivity (p >= 0.05)\n")
            cat("→ BP and TP show similar prey preferences\n")
            cat("   (compositional differences driven by availability, not selection)\n")
          }
        } else {
          cat("No residual component - insufficient samples for this model.\n")
          cat("Try testing Waveform alone:\n\n")

          perm_selectivity_wf_ps1 <- adonis2(
            selectivity_dist_ps1 ~ Waveform,
            data = selectivity_meta_ps1,
            permutations = 999
          )
          print(perm_selectivity_wf_ps1)
        }
      } else {
        cat("Insufficient waveform variation for PERMANOVA.\n")
      }

      # PCoA visualization
      pcoa_sel_ps1 <- cmdscale(selectivity_dist_ps1, eig = TRUE, k = 2)

      pcoa_sel_df_ps1 <- data.frame(
        PCoA1 = pcoa_sel_ps1$points[, 1],
        PCoA2 = pcoa_sel_ps1$points[, 2],
        selectivity_meta_ps1
      )

      p_selectivity_ps1 <- ggplot(
        pcoa_sel_df_ps1,
        aes(
          x = PCoA1, y = PCoA2,
          color = Waveform, shape = Region
        )
      ) +
        geom_point(size = 4, alpha = 0.7) +
        theme_minimal() +
        scale_color_brewer(palette = "Set1") +
        labs(
          title = "PS1: Individual Prey Selectivity Profiles (BP vs. TP)",
          subtitle = "PCoA on Ivlev's E values (each point = one fish)",
          color = "Waveform",
          shape = "Region"
        )

      print(p_selectivity_ps1)
    } else {
      cat("Insufficient samples for PERMANOVA (need at least 4 individuals).\n")
    }
  } else {
    cat("Could not calculate selectivity for PS1 (no matching environmental data)\n")
  }

  # ===== PRIMERSET 2 ANALYSIS =====

  # Get unique samples with locality info for PS2
  samples_ps2 <- ps2_pk_family_counts %>%
    distinct(SampleID, Locality_Name, Waveform_Type, Region)

  for (i in 1:nrow(samples_ps2)) {
    sample_id <- samples_ps2$SampleID[i]
    locality <- samples_ps2$Locality_Name[i]

    # Get site name for environmental data
    site_name <- locality_to_site[locality]
    if (is.na(site_name)) next

    # Get environmental data for this site
    env_site <- env_data %>% filter(Site == site_name)
    if (nrow(env_site) == 0) next

    # Calculate selectivity for this individual
    selectivity <- calc_individual_selectivity(
      sample_id = sample_id,
      diet_data = ps2_pk_family_counts,
      env_data_site = env_site,
      metadata = meta2_sel
    )

    if (!is.null(selectivity)) {
      selectivity_results_ps2[[sample_id]] <- selectivity
    }
  }

  # Combine and analyze PS2 results
  if (length(selectivity_results_ps2) > 0) {
    selectivity_ps2_df <- bind_rows(selectivity_results_ps2)

    cat("\n\n==============================================\n")
    cat("PRIMERSET 2 - Selectivity Analysis (Individual-level):\n")
    cat("  Individuals analyzed:", n_distinct(selectivity_ps2_df$SampleID), "\n")
    cat(
      "  Total family-individual combinations:",
      nrow(selectivity_ps2_df), "\n\n"
    )

    # Summary statistics by Waveform
    cat("Mean Ivlev's E by Waveform:\n")
    selectivity_summary_ps2 <- selectivity_ps2_df %>%
      group_by(Waveform, Region) %>%
      summarise(
        n_individuals = n_distinct(SampleID),
        n_families = n_distinct(Family),
        mean_E = mean(Ivlev_E, na.rm = TRUE),
        mean_abs_E = mean(abs(Ivlev_E), na.rm = TRUE),
        n_preference = sum(Selection_Type == "Preference"),
        n_avoidance = sum(Selection_Type == "Avoidance"),
        .groups = "drop"
      )

    print(selectivity_summary_ps2)

    # Create wide matrix for PERMANOVA
    selectivity_wide_ps2 <- selectivity_ps2_df %>%
      select(SampleID, Region, Waveform, Locality, Family, Ivlev_E) %>%
      group_by(SampleID, Region, Waveform, Locality, Family) %>%
      summarise(Ivlev_E = mean(Ivlev_E, na.rm = TRUE), .groups = "drop") %>%
      pivot_wider(names_from = Family, values_from = Ivlev_E, values_fill = 0)

    if (nrow(selectivity_wide_ps2) > 3) {
      selectivity_matrix_ps2 <- as.matrix(selectivity_wide_ps2[, -(1:4)])
      rownames(selectivity_matrix_ps2) <- selectivity_wide_ps2$SampleID

      selectivity_meta_ps2 <- selectivity_wide_ps2 %>%
        select(SampleID, Region, Waveform, Locality)

      # Distance matrix
      selectivity_dist_ps2 <- vegdist(selectivity_matrix_ps2, method = "euclidean")

      # PERMANOVA
      cat("\n\nPERMANOVA: Individual Selectivity ~ Waveform + Region\n")
      cat("Sample size: ", nrow(selectivity_meta_ps2), " individuals\n")

      if (n_distinct(selectivity_meta_ps2$Waveform) > 1 &&
        n_distinct(selectivity_meta_ps2$Region) > 0) {
        n_samples <- nrow(selectivity_meta_ps2)
        n_waveforms <- n_distinct(selectivity_meta_ps2$Waveform)
        n_regions <- n_distinct(selectivity_meta_ps2$Region)
        model_df <- (n_waveforms - 1) + (n_regions - 1)
        residual_df <- n_samples - model_df - 1

        cat("Model df:", model_df, ", Residual df:", residual_df, "\n\n")

        if (residual_df > 0) {
          perm_selectivity_ps2 <- adonis2(
            selectivity_dist_ps2 ~ Waveform + Region,
            data = selectivity_meta_ps2,
            permutations = 999,
            by = "margin"
          )
          print(perm_selectivity_ps2)

          cat("\nInterpretation:\n")
          if (!is.na(perm_selectivity_ps2$`Pr(>F)`[1]) &&
            perm_selectivity_ps2$`Pr(>F)`[1] < 0.05) {
            cat("Waveform DOES predict selectivity (p < 0.05)\n")
            cat("→ BP and TP show different prey preferences\n")
          } else {
            cat("Waveform does NOT predict selectivity (p >= 0.05)\n")
            cat("→ BP and TP show similar prey preferences\n")
          }
        } else {
          cat("No residual component - insufficient samples for this model.\n")
          cat("Try testing Waveform alone:\n\n")

          perm_selectivity_wf_ps2 <- adonis2(
            selectivity_dist_ps2 ~ Waveform,
            data = selectivity_meta_ps2,
            permutations = 999
          )
          print(perm_selectivity_wf_ps2)
        }
      } else {
        cat("Insufficient waveform variation for PERMANOVA.\n")
      }

      # PCoA visualization
      pcoa_sel_ps2 <- cmdscale(selectivity_dist_ps2, eig = TRUE, k = 2)

      pcoa_sel_df_ps2 <- data.frame(
        PCoA1 = pcoa_sel_ps2$points[, 1],
        PCoA2 = pcoa_sel_ps2$points[, 2],
        selectivity_meta_ps2
      )

      p_selectivity_ps2 <- ggplot(
        pcoa_sel_df_ps2,
        aes(
          x = PCoA1, y = PCoA2,
          color = Waveform, shape = Region
        )
      ) +
        geom_point(size = 4, alpha = 0.7) +
        theme_minimal() +
        scale_color_brewer(palette = "Set1") +
        labs(
          title = "PS2: Individual Prey Selectivity Profiles (BP vs. TP)",
          subtitle = "PCoA on Ivlev's E values (each point = one fish)",
          color = "Waveform",
          shape = "Region"
        )

      print(p_selectivity_ps2)
    } else {
      cat("Insufficient samples for PERMANOVA (need at least 4 individuals).\n")
    }
  } else {
    cat("Could not calculate selectivity for PS2 (no matching environmental data)\n")
  }
} else {
  cat("Environmental prey availability data not found at:\n")
  cat(env_file, "\n\n")
  cat("Skipping prey selectivity analysis.\n")
  cat("This analysis requires D-frame net collection data.\n")
}
```

## 4.7 Mean Selectivity Comparison

Calculate mean absolute selectivity (|E|) per fish to measure overall "choosiness", then compare BP vs. TP.

```{r level4-mean-selectivity, fig.width=10, fig.height=6}
cat("=== LEVEL 4.7: MEAN SELECTIVITY COMPARISON ===\n\n")

if (exists("selectivity_ps1_df") && nrow(selectivity_ps1_df) > 0) {
  # Calculate mean absolute selectivity per individual fish
  mean_selectivity <- selectivity_ps1_df %>%
    group_by(SampleID, Region, Waveform, Locality) %>%
    summarise(
      mean_abs_E = mean(abs(Ivlev_E), na.rm = TRUE),
      n_families = n(),
      .groups = "drop"
    )

  cat("Mean Absolute Selectivity by Individual:\n")
  cat("(Higher |E| = more selective/choosy)\n\n")
  print(head(mean_selectivity, 10))

  cat("\n\nSummary by Waveform:\n")
  print(mean_selectivity %>%
    group_by(Waveform) %>%
    summarise(
      n_fish = n(),
      mean_selectivity = mean(mean_abs_E),
      sd_selectivity = sd(mean_abs_E),
      .groups = "drop"
    ))

  # Test: Do BP and TP differ in overall selectivity?
  if (n_distinct(mean_selectivity$Waveform) > 1) {
    cat("\n\nWilcoxon Test: Mean |E| ~ Waveform\n")
    wt_mean_sel <- wilcox.test(mean_abs_E ~ Waveform, data = mean_selectivity)
    print(wt_mean_sel)

    cat("\n\nInterpretation:\n")
    if (wt_mean_sel$p.value < 0.05) {
      cat("BP and TP differ significantly in overall selectivity (p < 0.05)\n")

      bp_mean <- mean(mean_selectivity$mean_abs_E[mean_selectivity$Waveform == "BP"])
      tp_mean <- mean(mean_selectivity$mean_abs_E[mean_selectivity$Waveform == "TP"])

      if (bp_mean > tp_mean) {
        cat("→ BP fish are MORE selective (choosier) than TP fish\n")
      } else {
        cat("→ TP fish are MORE selective (choosier) than BP fish\n")
      }
    } else {
      cat("No significant difference in overall selectivity (p >= 0.05)\n")
      cat("→ BP and TP are equally selective\n")
    }

    # Visualization
    p_mean_sel <- ggplot(
      mean_selectivity,
      aes(x = Waveform, y = mean_abs_E, fill = Waveform)
    ) +
      geom_boxplot(alpha = 0.7) +
      geom_jitter(width = 0.1, size = 3, alpha = 0.6) +
      theme_minimal() +
      scale_fill_brewer(palette = "Set1") +
      labs(
        title = "Mean Absolute Selectivity: BP vs. TP",
        subtitle = "Higher values = more selective (choosier)",
        x = "Waveform",
        y = "Mean |Ivlev's E|"
      ) +
      theme(legend.position = "none")

    print(p_mean_sel)
  }
} else {
  cat("PS1: Selectivity data not available.\n")
}

# ===== PRIMERSET 2 MEAN SELECTIVITY =====

if (exists("selectivity_ps2_df") && nrow(selectivity_ps2_df) > 0) {
  cat("\n\n==============================================\n")
  cat("PRIMERSET 2 - Mean Selectivity\n")
  cat("==============================================\n\n")

  # Calculate mean absolute selectivity per individual fish
  mean_selectivity_ps2 <- selectivity_ps2_df %>%
    group_by(SampleID, Region, Waveform, Locality) %>%
    summarise(
      mean_abs_E = mean(abs(Ivlev_E), na.rm = TRUE),
      n_families = n(),
      .groups = "drop"
    )

  cat("Mean Absolute Selectivity by Individual:\n")
  cat("(Higher |E| = more selective/choosy)\n\n")
  print(head(mean_selectivity_ps2, 10))

  cat("\n\nSummary by Waveform:\n")
  print(mean_selectivity_ps2 %>%
    group_by(Waveform) %>%
    summarise(
      n_fish = n(),
      mean_selectivity = mean(mean_abs_E),
      sd_selectivity = sd(mean_abs_E),
      .groups = "drop"
    ))

  # Test: Do BP and TP differ in overall selectivity?
  if (n_distinct(mean_selectivity_ps2$Waveform) > 1) {
    cat("\n\nWilcoxon Test: Mean |E| ~ Waveform\n")
    wt_mean_sel_ps2 <- wilcox.test(mean_abs_E ~ Waveform,
      data = mean_selectivity_ps2
    )
    print(wt_mean_sel_ps2)

    cat("\n\nInterpretation:\n")
    if (wt_mean_sel_ps2$p.value < 0.05) {
      cat("BP and TP differ significantly in overall selectivity (p < 0.05)\n")

      bp_mean <- mean(mean_selectivity_ps2$mean_abs_E[
        mean_selectivity_ps2$Waveform == "BP"
      ])
      tp_mean <- mean(mean_selectivity_ps2$mean_abs_E[
        mean_selectivity_ps2$Waveform == "TP"
      ])

      if (bp_mean > tp_mean) {
        cat("→ BP fish are MORE selective (choosier) than TP fish\n")
      } else {
        cat("→ TP fish are MORE selective (choosier) than BP fish\n")
      }
    } else {
      cat("No significant difference in overall selectivity (p >= 0.05)\n")
      cat("→ BP and TP are equally selective\n")
    }

    # Visualization
    p_mean_sel_ps2 <- ggplot(
      mean_selectivity_ps2,
      aes(
        x = Waveform, y = mean_abs_E,
        fill = Waveform
      )
    ) +
      geom_boxplot(alpha = 0.7) +
      geom_jitter(width = 0.1, size = 3, alpha = 0.6) +
      theme_minimal() +
      scale_fill_brewer(palette = "Set1") +
      labs(
        title = "PS2: Mean Absolute Selectivity: BP vs. TP",
        subtitle = "Higher values = more selective (choosier)",
        x = "Waveform",
        y = "Mean |Ivlev's E|"
      ) +
      theme(legend.position = "none")

    print(p_mean_sel_ps2)
  }
} else {
  cat("PS2: Selectivity data not available.\n")
}
```

---

# Summary

This document contains:

- **Section 4.3b**: SIMPER analysis identifying prey families driving BP vs. TP dietary differences
- **Section 4.3c**: SIMPER analysis identifying prey families driving North vs. South dietary differences
- **Section 4.3d**: Mechanistic drill-down partitioning waveform differences into turnover vs. abundance
- **Section 4.3e**: Mechanistic drill-down partitioning region differences into turnover vs. abundance
- **Section 4.3f**: ASV-level visualization of waveform differences
- **Section 4.3g**: ASV-level visualization of region differences
- **Section 4.6**: Prey selectivity analysis using environmental D-frame net data
- **Section 4.7**: Mean selectivity comparison between BP and TP waveforms

All analyses use the phyloseq objects (`ps1_pk`, `ps2_pk`) and helper functions defined in `core_setup.R` and `gut_analysis_2026_helpers.R`.

---

# 4.8 Comprehensive Publication Figure

**Objective**: Synthesize waveform-based dietary niche partitioning findings into a 4-panel publication figure.

**The Story**: BP and TP fish eat different prey species (turnover, not abundance) → this reflects genuine preference differences → but they're equally selective overall.

**Panels**:
- **A**: ASV read composition (shared vs BP-only vs TP-only)
- **B**: Key ASVs driving differences (top 10 per primerset, faceted)
- **C**: Selectivity profiles (preference patterns via PCoA)
- **D**: Overall selectivity levels (equal choosiness via mean |E|)

---

```{r publication-figure, fig.width=14, fig.height=10, warning=FALSE}
cat("=== CREATING PUBLICATION FIGURE ===\n\n")

# ============================================================================
# PANEL A: Calculate ASV overlap and read proportions
# ============================================================================

cat("Panel A: Calculating ASV overlap and read proportions...\n")

overlap_ps1 <- calculate_asv_overlap_jaccard(ps1_pk, "Waveform_Type", "BP", "TP")
overlap_ps2 <- calculate_asv_overlap_jaccard(ps2_pk, "Waveform_Type", "BP", "TP")

props_ps1 <- calculate_asv_category_proportions(ps1_pk, overlap_ps1,
                                                  "Waveform_Type", "BP", "TP")
props_ps1$PrimerSet <- "PS1"

props_ps2 <- calculate_asv_category_proportions(ps2_pk, overlap_ps2,
                                                  "Waveform_Type", "BP", "TP")
props_ps2$PrimerSet <- "PS2"

props_combined <- rbind(props_ps1, props_ps2)
props_combined$Group_PS <- paste(props_combined$Group, props_combined$PrimerSet)
props_combined$Group_PS <- factor(props_combined$Group_PS,
                                   levels = c("BP PS1", "TP PS1", "BP PS2", "TP PS2"))

# ============================================================================
# PANEL B: Extract top 10 SIMPER ASVs from each primerset separately
# ============================================================================

cat("Panel B: Extracting top SIMPER ASVs...\n")

comp_name_ps1 <- names(simper_waveform_ps1$simper)[1]
comp_name_ps2 <- names(simper_waveform_ps2$simper)[1]

# Get top 10 from each primerset
top_asvs_ps1 <- extract_top_simper_asvs(
  simper_result = simper_waveform_ps1$simper,
  comparison_name = comp_name_ps1,
  ps = ps1_pk,
  overlap_result = overlap_ps1,
  group1 = "BP",
  group2 = "TP",
  top_n = 10
)
top_asvs_ps1$PrimerSet <- "PS1"

top_asvs_ps2 <- extract_top_simper_asvs(
  simper_result = simper_waveform_ps2$simper,
  comparison_name = comp_name_ps2,
  ps = ps2_pk,
  overlap_result = overlap_ps2,
  group1 = "BP",
  group2 = "TP",
  top_n = 10
)
top_asvs_ps2$PrimerSet <- "PS2"

# Combine but keep separate for faceting
top_asvs_combined <- rbind(top_asvs_ps1, top_asvs_ps2)

# Factor levels for proper ordering within each facet
top_asvs_combined <- top_asvs_combined %>%
  group_by(PrimerSet) %>%
  mutate(Display_Label = factor(Display_Label,
                                 levels = rev(Display_Label))) %>%
  ungroup()

top_asvs_combined$PrimerSet <- factor(top_asvs_combined$PrimerSet,
                                       levels = c("PS1", "PS2"))

cat("  Top 10 ASVs identified for each primerset\n")

# ============================================================================
# PANEL C: PCoA variance explained (already exists in pcoa_sel_ps1$eig)
# ============================================================================

cat("Panel C: Calculating variance explained...\n")

eig <- pcoa_sel_ps1$eig
var_exp_ps1 <- eig / sum(eig) * 100

cat(sprintf("  PCoA1: %.1f%%, PCoA2: %.1f%%\n", var_exp_ps1[1], var_exp_ps1[2]))

# ============================================================================
# PANEL D: Combine mean selectivity from both primersets
# ============================================================================

cat("Panel D: Combining mean selectivity data...\n")

mean_selectivity$PrimerSet <- "PS1"
mean_selectivity_ps2$PrimerSet <- "PS2"
mean_sel_combined <- rbind(mean_selectivity, mean_selectivity_ps2)
mean_sel_combined$Group <- paste(mean_sel_combined$Waveform, mean_sel_combined$PrimerSet)
mean_sel_combined$Group <- factor(mean_sel_combined$Group,
                                   levels = c("BP PS1", "TP PS1", "BP PS2", "TP PS2"))

# ============================================================================
# CREATE PANELS
# ============================================================================

cat("\nCreating panels...\n")

# Color scheme (consistent across all panels)
asv_colors <- c("Shared" = "gray60",
                "BP-only" = "#E07020",   # orange
                "TP-only" = "#2090C0")   # blue
wf_colors <- c("BP" = "#E07020", "TP" = "#2090C0")

# Panel A: Stacked bar chart (read proportions)
panel_a <- ggplot(props_combined, aes(x = Group_PS, y = Proportion, fill = Category)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = sprintf("%.1f%%", Proportion)),
            position = position_stack(vjust = 0.5),
            color = "white", fontface = "bold", size = 3.5) +
  scale_fill_manual(values = asv_colors, name = "ASV Category") +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 100)) +
  labs(title = "A. Read Proportion from Shared & Unique ASVs",
       x = NULL, y = "Proportion of Reads (%)") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 13, face = "bold"),
    axis.title.y = element_text(face = "bold"),
    panel.grid.major.x = element_blank()
  )

# Panel B: Horizontal bars faceted by primerset
panel_b <- ggplot(top_asvs_combined, aes(x = Contribution_Pct, y = Display_Label, fill = Category)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = asv_colors, name = "ASV Category") +
  scale_x_continuous(expand = c(0, 0)) +
  facet_wrap(~ PrimerSet, ncol = 1, scales = "free_y") +
  labs(title = "B. Top 10 ASVs Driving Waveform Differences",
       subtitle = "Ranked by contribution to dissimilarity within each primerset",
       x = "Contribution to Dissimilarity (%)",
       y = "Taxon (Genus + Family)") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 8),
    plot.title = element_text(size = 13, face = "bold"),
    plot.subtitle = element_text(size = 10),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    strip.text = element_text(face = "bold", size = 11),
    strip.background = element_rect(fill = "gray90", color = NA),
    legend.position = "none",
    panel.grid.major.y = element_blank()
  )

# Panel C: PCoA scatter
panel_c <- ggplot(pcoa_sel_df_ps1, aes(x = PCoA1, y = PCoA2,
                                        color = Waveform, shape = Region)) +
  geom_point(size = 3.5, alpha = 0.8) +
  stat_ellipse(aes(group = Waveform), level = 0.95, linetype = 2) +
  scale_color_manual(values = wf_colors, name = "Waveform") +
  scale_shape_manual(values = c("North" = 16, "South" = 17), name = "Region") +
  labs(title = "C. Prey Selectivity Profiles (Ivlev's E)",
       subtitle = "PS1 only; PS2 shows concordant pattern",
       x = sprintf("PCoA1 (%.1f%%)", var_exp_ps1[1]),
       y = sprintf("PCoA2 (%.1f%%)", var_exp_ps1[2])) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 13, face = "bold"),
    plot.subtitle = element_text(size = 10),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    legend.title = element_text(face = "bold")
  )

# Panel D: Boxplots with annotations
# Determine appropriate y-axis range
y_max <- max(mean_sel_combined$mean_abs_E, na.rm = TRUE)
y_range <- y_max - 0.75

panel_d <- ggplot(mean_sel_combined, aes(x = Group, y = mean_abs_E, fill = Waveform)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  scale_fill_manual(values = wf_colors, name = "Waveform") +
  scale_y_continuous(limits = c(0.75, y_max + 0.02), expand = c(0, 0)) +
  labs(title = "D. Mean Absolute Selectivity by Waveform",
       x = NULL, y = "Mean |Ivlev's E|") +
  # Add n.s. brackets (positioned relative to top of range)
  annotate("segment", x = 0.7, xend = 2.3, y = y_max - 0.01, yend = y_max - 0.01, linewidth = 0.5) +
  annotate("text", x = 1.5, y = y_max + 0.005, label = "n.s.", size = 3.5) +
  annotate("segment", x = 2.7, xend = 4.3, y = y_max - 0.01, yend = y_max - 0.01, linewidth = 0.5) +
  annotate("text", x = 3.5, y = y_max + 0.005, label = "n.s.", size = 3.5) +
  # Add italic annotation (positioned near bottom)
  annotate("text", x = 2.5, y = 0.76,
           label = "Equally selective, different preferences",
           fontface = "italic", size = 3.5, color = "gray30") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 13, face = "bold"),
    axis.title.y = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# ============================================================================
# COMBINE PANELS
# ============================================================================

cat("Combining panels...\n")

combined_fig <- (panel_a | panel_b) / (panel_c | panel_d) +
  plot_annotation(
    title = "Waveform-Based Dietary Niche Partitioning in Paramormyrops kingsleyae",
    subtitle = "Compositional differences driven by species turnover, not differential selectivity",
    theme = theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
                  plot.subtitle = element_text(size = 12, hjust = 0.5))
  ) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

print(combined_fig)

cat("\n=== FIGURE CREATION COMPLETE ===\n")
```

```{r publication-figure-export, include=FALSE}
cat("\n=== EXPORTING FIGURE ===\n\n")

output_dir <- file.path(root, "output_data/07_Analysis")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

pdf_path <- file.path(output_dir, "pk_waveform_comprehensive_figure.pdf")
png_path <- file.path(output_dir, "pk_waveform_comprehensive_figure.png")

ggsave(pdf_path, combined_fig, width = 14, height = 10, units = "in")
ggsave(png_path, combined_fig, width = 14, height = 10, units = "in", dpi = 300)

cat("Figure exported successfully:\n")
cat("  PDF:", pdf_path, "\n")
cat("  PNG:", png_path, "\n")
```
