---
title: "Mormyrid Gut Content Analysis: Dietary Niche Partitioning and EOD Waveform Polymorphism"
author: "Gorsuch, Picq, Gallant"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
    keep_tex: false
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
    df_print: paged
header-includes:
- \usepackage{longtable}
- \usepackage{booktabs}
- \usepackage{float}
- \usepackage{array}
- \usepackage{pdflscape}
---

# Overview

This analysis examines dietary niche partitioning in mormyrid electric fishes across four hierarchical biological levels, with a focus on the relationship between EOD waveform polymorphism and diet in *Paramormyrops kingsleyae*.

**Analytical Framework:**

1. **LEVEL 1:** Mormyrid Community at Balé Creek (9 species, 35 individuals, single site)
2. **LEVEL 2:** BP vs. TP Mormyrids Globally (~97-99 individuals across all sites)
3. **LEVEL 3:** *P. kingsleyae* General Population Factors (~47 individuals from 4 rivers)
4. **LEVEL 4:** BP vs. TP Within *P. kingsleyae* (The Core Question)
5. **LEVEL 5:** Cross-Validation Across Primer Sets

**Primer Sets:**

| Set | Forward | Reverse | Name |
|-----|---------|---------|------|
| PS1 | mlCOIintF | BR2 | Primerset 1 |
| PS2 | BF2 | BR1 | Primerset 2 |

All analyses are conducted on both primer sets in parallel to ensure biological robustness.

---

```{r setup, include=FALSE}
# Configure knitr
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,   # or 8                                                                                                                                                   
  fig.height = 4.5 ,# or 5   
  dpi = 300
)

# Load required libraries
library(tidyverse)      # Data manipulation and visualization
library(qiime2R)        # QIIME2 data import
library(phyloseq)       # Microbiome analysis framework
library(microbiome)     # Microbiome utilities
library(vegan)          # Ecological statistics
library(pairwiseAdonis) # Pairwise PERMANOVA tests
library(spaa)           # Diet overlap indices
library(patchwork)      # Multi-panel figures
library(knitr)          # Table formatting
library(kableExtra)     # Enhanced tables
library(ape)            # Phylogenetic analysis
library(effsize)        # Effect size calculations
library(stringdist)     # String distance calculations for fuzzy matching

# Set seed for reproducibility
set.seed(123)

# Set root path
root <- rprojroot::find_root(".git/index")

# Source helper functions
source(file.path(root, "code/07_Analysis/gut_analysis_2026_helpers.R"))

# Color palettes (colorblind-friendly)
# Waveform colors
bp_color <- "#0072B2"      # Blue for BP
tp_color <- "#D55E00"      # Orange for TP

# Region colors
north_color <- "#009E73"   # Green for North
south_color <- "#CC79A7"   # Purple for South

# Primer set colors
ps1_color <- "#E69F00"     # Orange
ps2_color <- "#56B4E9"     # Sky blue

# Selectivity colors
preference_color <- "#D55E00"   # Orange
neutral_color <- "#F0E442"      # Yellow
avoidance_color <- "#0072B2"    # Blue

# Set global ggplot2 theme with sans-serif font and publication-quality sizes
theme_set(theme_minimal(base_size = 12, base_family = "sans"))
theme_update(
  axis.title = element_text(size = 14, face = "bold"),
  axis.text = element_text(size = 12),
  legend.title = element_text(size = 12, face = "bold"),
  legend.text = element_text(size = 11),
  strip.text = element_text(size = 12, face = "bold"),
  plot.title = element_text(size = 14, face = "bold")
)
```

---

# Data Loading and Preparation

Load phyloseq objects for both primer sets, filter host sequences, and rarefy to common depth.

```{r load-data}
cat("=== LOADING PRIMER SET DATA ===\n\n")

# Load both primer sets with host filtering
ps1 <- load_primerset_data(1, use_no_host = TRUE)
ps2 <- load_primerset_data(2, use_no_host = TRUE)

cat("\n=== INITIAL DATASET SUMMARY ===\n")
cat("Primer Set 1:", nsamples(ps1), "samples,", ntaxa(ps1), "ASVs\n")
cat("Primer Set 2:", nsamples(ps2), "samples,", ntaxa(ps2), "ASVs\n")

# Check rarefaction depths
cat("\nRarefaction depths:\n")
cat("  PS1:", unique(sample_sums(ps1)), "\n")
cat("  PS2:", unique(sample_sums(ps2)), "\n")

# Initialize global color palette for consistent family colors across all plots
cat("\n=== CREATING GLOBAL COLOR PALETTE ===\n")
GLOBAL_FAMILY_COLORS <- create_global_family_palette(ps1, ps2, min_families = 50)
cat("Created color palette for", length(GLOBAL_FAMILY_COLORS) - 1, "families + Other\n")
```

## Sample Matching and Metadata Preparation

```{r sample-matching}
# Get sample names for both primer sets
ps1_samples <- sample_data(ps1)$Sample_Name
ps2_samples <- sample_data(ps2)$Sample_Name

# Find shared samples
shared_samples <- intersect(ps1_samples, ps2_samples)

cat("=== SAMPLE OVERLAP ===\n")
cat("PS1 total samples:", length(ps1_samples), "\n")
cat("PS2 total samples:", length(ps2_samples), "\n")
cat("Shared samples:", length(shared_samples), "\n")
cat("Overlap proportion:",
    round(length(shared_samples) / length(union(ps1_samples, ps2_samples)), 3), "\n")

# Extract metadata (will be same for both after matching)
meta_all <- data.frame(sample_data(ps1), stringsAsFactors = FALSE)

# Create species-level scientific names
meta_all$Species <- clean_species_name(meta_all$Species)

# Display sample sizes by species and river
sample_summary <- meta_all %>%
  group_by(Species, Locality_Name) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = Locality_Name, values_from = n, values_fill = list(n = 0)) %>%
  mutate(Total = rowSums(select(., -Species))) %>%
  bind_rows(
    summarise(., across(where(is.numeric), sum), Species = "Total")
  )

kable(sample_summary,
      caption = "Sample sizes by species and river",
      align = c("l", rep("r", ncol(sample_summary) - 1)),
      booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE) %>%
  row_spec(nrow(sample_summary), bold = TRUE)
```

## Create Subsets for Different Analytical Levels

```{r create-subsets}
cat("\n=== CREATING ANALYTICAL SUBSETS ===\n\n")

# LEVEL 1: Balé Creek only (9 species)
ps1_bale <- subset_samples(ps1, Locality_Name == "Bale Creek")
ps2_bale <- subset_samples(ps2, Locality_Name == "Bale Creek")

# Remove rare taxa (present in <2 samples or <10 total reads)
ps1_bale <- filter_taxa(ps1_bale, function(x) sum(x > 0) >= 2 & sum(x) >= 10, TRUE)
ps2_bale <- filter_taxa(ps2_bale, function(x) sum(x > 0) >= 2 & sum(x) >= 10, TRUE)

cat("LEVEL 1 - Balé Creek:\n")
cat("  PS1:", nsamples(ps1_bale), "samples,", ntaxa(ps1_bale), "ASVs\n")
cat("  PS2:", nsamples(ps2_bale), "samples,", ntaxa(ps2_bale), "ASVs\n")

# LEVEL 2: All mormyrids (BP vs TP globally)
ps1_all_mormyrids <- ps1
ps2_all_mormyrids <- ps2

# Add waveform type metadata (BP or TP)
# Note: This assumes Waveform_Type column exists in metadata
cat("\nLEVEL 2 - All Mormyrids (BP vs TP):\n")
cat("  PS1:", nsamples(ps1_all_mormyrids), "samples\n")
cat("  PS2:", nsamples(ps2_all_mormyrids), "samples\n")

waveform_counts <- table(sample_data(ps1)$Waveform_Type)
cat("  BP samples:", waveform_counts["BP"], "\n")
cat("  TP samples:", waveform_counts["TP"], "\n")

# LEVEL 3 & 4: P. kingsleyae only
ps1_pk <- subset_samples(ps1, Species == "Paramormyrops kingsleyae")
ps2_pk <- subset_samples(ps2, Species == "Paramormyrops kingsleyae")

# Remove rare taxa
ps1_pk <- filter_taxa(ps1_pk, function(x) sum(x > 0) >= 2 & sum(x) >= 10, TRUE)
ps2_pk <- filter_taxa(ps2_pk, function(x) sum(x > 0) >= 2 & sum(x) >= 10, TRUE)

cat("\nLEVEL 3/4 - P. kingsleyae:\n")
cat("  PS1:", nsamples(ps1_pk), "samples,", ntaxa(ps1_pk), "ASVs\n")
cat("  PS2:", nsamples(ps2_pk), "samples,", ntaxa(ps2_pk), "ASVs\n")

# Check P. kingsleyae sampling design
pk_meta <- data.frame(sample_data(ps1_pk))
pk_design <- pk_meta %>%
  group_by(Locality_Name, Waveform_Type) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = Waveform_Type, values_from = n, values_fill = list(n = 0))

cat("\nP. kingsleyae sampling design:\n")
print(pk_design)
```

## Load Phylogenetic Data

```{r load-phylogeny}
cat("\n=== LOADING PHYLOGENETIC DATA ===\n\n")

# Load phylogenetic tree
tree_file <- file.path(root, "input_data/ssAll_iqtree_timtree.tree")

if (file.exists(tree_file)) {
  # Read tree - NEXUS format may create multiple trees, extract the real one
  phylo_tree_raw <- read.tree(tree_file)

  # If multiPhylo object, extract the second tree (the actual phylogeny)
  if (inherits(phylo_tree_raw, "multiPhylo")) {
    phylo_tree <- phylo_tree_raw[[2]]
    cat("Note: NEXUS file contained", length(phylo_tree_raw),
        "trees, extracted tree 2\n\n")
  } else {
    phylo_tree <- phylo_tree_raw
  }

  cat("Phylogenetic tree loaded:\n")
  cat("  Tips:", length(phylo_tree$tip.label), "\n")
  cat("  Tree type:", ifelse(is.ultrametric(phylo_tree), "Ultrametric",
                              "Non-ultrametric"), "\n")

  # Create cophenetic distance matrix
  phylo_dist <- cophenetic(phylo_tree)

  cat("\nPhylogenetic distances computed\n")
  cat("  Dimensions:", dim(phylo_dist)[1], "x", dim(phylo_dist)[2], "\n")

  # Show relevant tree tips for debugging
  cat("\nRelevant tree tips by genus:\n")
  param_tips <- grep("Paramormyrops", phylo_tree$tip.label, value = TRUE)
  morm_tips <- grep("^Mormyrops", phylo_tree$tip.label, value = TRUE)
  marc_tips <- grep("Marcusenius", phylo_tree$tip.label, value = TRUE)
  petro_tips <- grep("Petrocephalus", phylo_tree$tip.label, value = TRUE)

  cat("\nParamormyrops (", length(param_tips), " tips):\n", sep = "")
  cat(paste(param_tips, collapse = "\n"), "\n")

  cat("\nMormyrops (", length(morm_tips), " tips):\n", sep = "")
  cat(paste(morm_tips, collapse = "\n"), "\n")

  cat("\nMarcusenius (", length(marc_tips), " tips):\n", sep = "")
  cat(paste(marc_tips, collapse = "\n"), "\n")

  cat("\nPetrocephalus (", length(petro_tips), " tips, showing first 5):\n", sep = "")
  cat(paste(head(petro_tips, 5), collapse = "\n"), "\n")

  # Fuzzy match Balé species to phylogeny
  match_results <- match_bale_species_to_phylogeny(ps1_bale, phylo_tree)

  cat("\nBalé Creek species matching:\n")
  cat("  Successfully matched:", match_results$n_matched, "/", match_results$n_species, "\n\n")

  # Print detailed results
  print(match_results$match_table)

  # Extract matched tips for downstream use
  matched_tips <- match_results$matched_species
  if (length(matched_tips) > 0) {
    phylo_dist_bale <- phylo_dist[matched_tips, matched_tips]
  }

} else {
  cat("WARNING: Phylogenetic tree file not found at:\n")
  cat("  ", tree_file, "\n")
  phylo_tree <- NULL
  phylo_dist <- NULL
}
```

## Helper Functions

Additional helper functions specific to this analysis.

```{r additional-helpers}
#' Aggregate phyloseq object to family level
#'
#' @param ps phyloseq object
#' @return phyloseq object aggregated to family level
aggregate_to_family <- function(ps) {
  ps_family <- tax_glom(ps, taxrank = "Family", NArm = FALSE)
  return(ps_family)
}


#' Create stacked bar chart of prey composition at family level
#'
#' @param ps phyloseq object (should be aggregated to family level)
#' @param group_var Character, grouping variable for x-axis
#' @param top_n Integer, number of top families to show (rest as "Other")
#' @param title Character, plot title
#' @return ggplot object
plot_composition_bars <- function(ps, group_var, top_n = 10, title = "") {

  # Melt to long format (before normalization)
  ps_melt <- psmelt(ps)

  # Clean family names
  ps_melt$Family_clean <- sapply(ps_melt$Family, clean_family_name)

  # Aggregate by group_var and Family, summing counts
  ps_agg <- ps_melt %>%
    group_by(.data[[group_var]], Family_clean) %>%
    summarise(Abundance = sum(Abundance), .groups = "drop")

  # Normalize within each group to get relative abundance
  ps_agg <- ps_agg %>%
    group_by(.data[[group_var]]) %>%
    mutate(RelAbund = Abundance / sum(Abundance)) %>%
    ungroup()

  # Identify top N families by mean relative abundance across groups
  top_families <- ps_agg %>%
    group_by(Family_clean) %>%
    summarise(mean_abund = mean(RelAbund), .groups = "drop") %>%
    arrange(desc(mean_abund)) %>%
    slice_head(n = top_n) %>%
    pull(Family_clean)

  # Categorize families and aggregate "Other"
  ps_final <- ps_agg %>%
    mutate(Family_display = ifelse(Family_clean %in% top_families,
                                   Family_clean, "Other")) %>%
    group_by(.data[[group_var]], Family_display) %>%
    summarise(RelAbund = sum(RelAbund), .groups = "drop")

  # Reorder for plotting (top families first, then Other)
  ps_final$Family_display <- factor(ps_final$Family_display,
                                     levels = c(top_families, "Other"))

  # Create plot with consistent color scale
  p <- ggplot(ps_final, aes(x = .data[[group_var]], y = RelAbund,
                            fill = Family_display)) +
    geom_bar(stat = "identity", position = "stack") +
    scale_fill_manual(values = GLOBAL_FAMILY_COLORS,
                      name = "Prey Family",
                      na.value = "#808080",
                      drop = FALSE) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom",
          legend.key.size = unit(0.35, "cm"),
          legend.box = "vertical") +
    guides(fill = guide_legend(ncol = 4, title.position = "top",
                               byrow = TRUE, nrow = 3)) +
    labs(title = title, x = group_var, y = "Relative Abundance") +
    scale_y_continuous(labels = scales::percent)

  return(p)
}


#' Variance partitioning for PERMANOVA
#'
#' @param beta_result Beta diversity result from run_beta_diversity_analysis()
#' @param var1_formula Character, formula for first variable (e.g., "~ Region")
#' @param var2_formula Character, formula for second variable (e.g., "~ Waveform_Type")
#' @return List with varpart result and sequential PERMANOVA results
variance_partition_permanova <- function(beta_result, var1_formula, var2_formula) {

  dist_mat <- beta_result$distance
  meta <- beta_result$metadata

  # Method 1: varpart
  varpart_result <- varpart(dist_mat,
                            as.formula(var1_formula),
                            as.formula(var2_formula),
                            data = meta)

  # Method 2: Sequential PERMANOVA (variable 1 first)
  formula_seq1 <- paste("dist_mat", var1_formula, "+",
                       gsub("~\\s*", "", var2_formula))
  perm_seq1 <- adonis2(as.formula(formula_seq1),
                      data = meta, permutations = 999, by = "terms")

  # Method 3: Sequential PERMANOVA (variable 2 first)
  formula_seq2 <- paste("dist_mat", var2_formula, "+",
                       gsub("~\\s*", "", var1_formula))
  perm_seq2 <- adonis2(as.formula(formula_seq2),
                      data = meta, permutations = 999, by = "terms")

  # Method 4: Marginal effects
  formula_marginal <- paste("dist_mat", var1_formula, "+",
                           gsub("~\\s*", "", var2_formula))
  perm_marginal <- adonis2(as.formula(formula_marginal),
                          data = meta, permutations = 999, by = "margin")

  return(list(
    varpart = varpart_result,
    sequential_var1_first = perm_seq1,
    sequential_var2_first = perm_seq2,
    marginal = perm_marginal
  ))
}


#' Calculate diet overlap using Schoener and Pianka indices
#'
#' @param ps phyloseq object
#' @param group_var Character, grouping variable (e.g., "Species")
#' @return List with Schoener and Pianka matrices
calculate_diet_overlap <- function(ps, group_var) {

  # Convert to relative abundance
  ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))

  # Get OTU table (taxa as rows, samples as columns)
  otu_mat <- as.matrix(otu_table(ps_rel))
  if (!taxa_are_rows(ps_rel)) {
    otu_mat <- t(otu_mat)
  }

  # Get metadata
  meta <- data.frame(sample_data(ps_rel))

  # Calculate mean abundance per group
  groups <- unique(meta[[group_var]])
  mean_abundances <- matrix(0, nrow = ntaxa(ps_rel), ncol = length(groups))
  rownames(mean_abundances) <- taxa_names(ps_rel)
  colnames(mean_abundances) <- groups

  for (g in groups) {
    samples_in_group <- rownames(meta)[meta[[group_var]] == g]
    if (length(samples_in_group) == 1) {
      mean_abundances[, g] <- otu_mat[, samples_in_group]
    } else {
      mean_abundances[, g] <- rowMeans(otu_mat[, samples_in_group, drop = FALSE])
    }
  }

  # Transpose for spaa package (expects species as columns, sites/groups as rows)
  mean_abundances_t <- t(mean_abundances)

  # Calculate overlap indices
  schoener <- niche.overlap(mean_abundances_t, method = "schoener")
  pianka <- niche.overlap(mean_abundances_t, method = "pianka")

  return(list(
    schoener = schoener,
    pianka = pianka,
    mean_abundances = mean_abundances
  ))
}
```

---

# PRIMER SET CONCORDANCE ANALYSIS

## Overview

This section quantifies concordance between PS1 and PS2 on shared samples to address concerns about primer effects in metabarcoding. Esnaola et al. (2018) found primer choice explained R² = 0.353 of variation in fish diet - nearly twice the study site effect. We test whether our primers show similar effects when applied to identical biological samples.

```{r primer-concordance-setup}
cat("\n=== PRIMER SET CONCORDANCE ANALYSIS ===\n\n")

# Subset to shared samples only (fair comparison)
ps1_shared <- subset_samples(ps1, Sample_Name %in% shared_samples)
ps2_shared <- subset_samples(ps2, Sample_Name %in% shared_samples)

# Remove absent taxa
ps1_shared <- prune_taxa(taxa_sums(ps1_shared) > 0, ps1_shared)
ps2_shared <- prune_taxa(taxa_sums(ps2_shared) > 0, ps2_shared)

cat(sprintf("Shared samples: n = %d\n", length(shared_samples)))
cat(sprintf("PS1 taxa after pruning: %d\n", ntaxa(ps1_shared)))
cat(sprintf("PS2 taxa after pruning: %d\n", ntaxa(ps2_shared)))
```

```{r primer-beta-diversity}
cat("\n=== Beta Diversity Analysis (Species Effect) ===\n\n")

# PS1
beta1_shared <- run_beta_diversity_analysis(ps1_shared, "~ Species")
species_r2_ps1 <- beta1_shared$permanova$R2[1]
species_p_ps1 <- beta1_shared$permanova$`Pr(>F)`[1]

cat(sprintf("PS1 - Species R²: %.3f (p = %s)\n",
            species_r2_ps1, format.pval(species_p_ps1, digits = 3)))

# PS2
beta2_shared <- run_beta_diversity_analysis(ps2_shared, "~ Species")
species_r2_ps2 <- beta2_shared$permanova$R2[1]
species_p_ps2 <- beta2_shared$permanova$`Pr(>F)`[1]

cat(sprintf("PS2 - Species R²: %.3f (p = %s)\n",
            species_r2_ps2, format.pval(species_p_ps2, digits = 3)))

mean_r2 <- mean(c(species_r2_ps1, species_r2_ps2))
cat(sprintf("\nMean Species R²: %.3f\n", mean_r2))
cat(sprintf("Esnaola primer effect threshold: 0.353\n"))
```

```{r primer-concordance-tests}
cat("\n=== Concordance Tests ===\n\n")

# Ensure samples match by Sample_Name (not SampleID)
# Convert distance matrices to matrices with Sample_Name labels
dist1_mat <- as.matrix(beta1_shared$distance)
dist2_mat <- as.matrix(beta2_shared$distance)

# Relabel with Sample_Names from metadata
rownames(dist1_mat) <- beta1_shared$metadata$Sample_Name
colnames(dist1_mat) <- beta1_shared$metadata$Sample_Name
rownames(dist2_mat) <- beta2_shared$metadata$Sample_Name
colnames(dist2_mat) <- beta2_shared$metadata$Sample_Name

# Find truly shared samples and subset (automatically aligned by name)
shared_check <- intersect(beta1_shared$metadata$Sample_Name,
                          beta2_shared$metadata$Sample_Name)
cat(sprintf("Samples matched by Sample_Name: %d\n", length(shared_check)))

dist1_aligned <- as.dist(dist1_mat[shared_check, shared_check])
dist2_aligned <- as.dist(dist2_mat[shared_check, shared_check])

# Mantel test - correlation between distance matrices
mantel_result <- run_mantel_test(dist1_aligned, dist2_aligned)
cat(sprintf("Mantel correlation: r = %.3f (p = %s)\n",
            mantel_result$statistic,
            format.pval(mantel_result$signif, digits = 3)))

# For Procrustes, also need to align PCoA coordinates by Sample_Name
pcoa1_mat <- beta1_shared$pcoa$points
pcoa2_mat <- beta2_shared$pcoa$points
rownames(pcoa1_mat) <- beta1_shared$metadata$Sample_Name
rownames(pcoa2_mat) <- beta2_shared$metadata$Sample_Name

# Subset and reorder to match
pcoa1_aligned <- pcoa1_mat[shared_check, ]
pcoa2_aligned <- pcoa2_mat[shared_check, ]

# Procrustes analysis - ordination congruence
library(vegan)
pro <- procrustes(pcoa1_aligned, pcoa2_aligned, symmetric = TRUE)
pro_test <- protest(pcoa1_aligned, pcoa2_aligned, permutations = 999)

# pro_test$t0 IS the correlation, not m²
pro_cor <- pro_test$t0  # correlation directly from protest()
pro_m2 <- 1 - pro_test$t0^2  # calculate m² from correlation

cat(sprintf("Procrustes correlation: %.3f (higher = better concordance, p = %s)\n",
            pro_cor, format.pval(pro_test$signif, digits = 3)))
cat(sprintf("Procrustes m²: %.4f (lower = better fit)\n",
            pro_m2))

cat("\n=== Primer Set Effect Size (PERMANOVA) ===\n\n")

# Create merged dataset to test Primer_Set effect
# Merge OTU tables (union of all taxa, filling zeros where absent)
otu1 <- as.matrix(otu_table(ps1_shared))
otu2 <- as.matrix(otu_table(ps2_shared))

# Get union of all taxa
all_taxa <- union(rownames(otu1), rownames(otu2))

# Create full matrices with zeros for missing taxa
otu1_full <- matrix(0, nrow = length(all_taxa), ncol = ncol(otu1))
rownames(otu1_full) <- all_taxa
colnames(otu1_full) <- colnames(otu1)
otu1_full[rownames(otu1), ] <- otu1

otu2_full <- matrix(0, nrow = length(all_taxa), ncol = ncol(otu2))
rownames(otu2_full) <- all_taxa
colnames(otu2_full) <- colnames(otu2)
otu2_full[rownames(otu2), ] <- otu2

# Relabel samples with Sample_Name and add PS suffix for uniqueness
meta1 <- beta1_shared$metadata[match(colnames(otu1), rownames(beta1_shared$metadata)), ]
meta2 <- beta2_shared$metadata[match(colnames(otu2), rownames(beta2_shared$metadata)), ]

colnames(otu1_full) <- paste0(meta1$Sample_Name, "_PS1")
colnames(otu2_full) <- paste0(meta2$Sample_Name, "_PS2")

# Combine OTU tables
otu_combined <- cbind(otu1_full, otu2_full)

# Create combined metadata
meta1$SampleID_new <- paste0(meta1$Sample_Name, "_PS1")
meta1$Primer_Set <- "PS1"
meta2$SampleID_new <- paste0(meta2$Sample_Name, "_PS2")
meta2$Primer_Set <- "PS2"

meta_combined <- rbind(meta1, meta2)
rownames(meta_combined) <- meta_combined$SampleID_new

# Transpose for vegan (samples as rows)
otu_combined_t <- t(otu_combined)

# Calculate distance and run PERMANOVA
dist_combined <- vegdist(otu_combined_t, method = "bray")

# Test Primer_Set effect (controlling for Species)
perm_primer <- adonis2(dist_combined ~ Primer_Set + Species,
                       data = meta_combined,
                       permutations = 999)

primer_r2 <- perm_primer$R2[1]
primer_p <- perm_primer$`Pr(>F)`[1]

cat(sprintf("Unstratified PERMANOVA:\n"))
cat(sprintf("Primer_Set R²: %.3f (p = %s)\n",
            primer_r2, format.pval(primer_p, digits = 3)))
cat(sprintf("Esnaola et al. primer effect: R² = 0.353\n"))
cat(sprintf("Our primer effect: %.1f%% of Esnaola's\n\n",
            (primer_r2 / 0.353) * 100))

# Stratified PERMANOVA - accounts for paired samples (same fish, both primers)
perm_primer_strata <- adonis2(dist_combined ~ Primer_Set,
                               data = meta_combined,
                               strata = meta_combined$Sample_Name,
                               permutations = 999)

primer_r2_strata <- perm_primer_strata$R2[1]
primer_p_strata <- perm_primer_strata$`Pr(>F)`[1]

cat(sprintf("Stratified by Sample (paired design):\n"))
cat(sprintf("Primer_Set R²: %.3f (p = %s)\n",
            primer_r2_strata, format.pval(primer_p_strata, digits = 3)))
cat(sprintf("Esnaola comparison: %.1f%% of their primer effect\n",
            (primer_r2_strata / 0.353) * 100))
```

```{r primer-pcoa-plots, fig.width=12, fig.height=6}
# Use aligned coordinates that match by Sample_Name
pcoa1_df <- data.frame(
  PC1 = pcoa1_aligned[, 1],
  PC2 = pcoa1_aligned[, 2],
  Sample_Name = rownames(pcoa1_aligned),
  Primer_Set = "PS1"
)

pcoa2_df <- data.frame(
  PC1 = pcoa2_aligned[, 1],
  PC2 = pcoa2_aligned[, 2],
  Sample_Name = rownames(pcoa2_aligned),
  Primer_Set = "PS2"
)

# Add Species labels from metadata
meta_shared <- beta1_shared$metadata %>%
  filter(Sample_Name %in% shared_check) %>%
  select(Sample_Name, Species)

pcoa1_df <- merge(pcoa1_df, meta_shared, by = "Sample_Name")
pcoa2_df <- merge(pcoa2_df, meta_shared, by = "Sample_Name")

pcoa_combined <- rbind(pcoa1_df, pcoa2_df)

# Create faceted plot
ggplot(pcoa_combined, aes(x = PC1, y = PC2, color = Species)) +
  geom_point(size = 3, alpha = 0.7) +
  facet_wrap(~ Primer_Set) +
  theme_minimal() +
  labs(title = "Beta Diversity: Comparing Primer Sets on Shared Samples",
       subtitle = sprintf("Mantel r = %.3f | PS1 R² = %.3f | PS2 R² = %.3f",
                          mantel_result$statistic, species_r2_ps1, species_r2_ps2)) +
  coord_fixed()
```

```{r primer-procrustes-plot, fig.width=8, fig.height=8}
# Extract transformed coordinates (pro object already uses aligned matrices)
coords_ps1 <- data.frame(pro$X)
coords_ps2 <- data.frame(pro$Yrot)

# Create plot dataframe with proper column names
plot_df <- data.frame(
  x1 = coords_ps1[, 1],
  y1 = coords_ps1[, 2],
  x2 = coords_ps2[, 1],
  y2 = coords_ps2[, 2]
)

ggplot(plot_df) +
  geom_segment(aes(x = x1, y = y1, xend = x2, yend = y2),
               arrow = arrow(length = unit(0.1, "cm")),
               alpha = 0.3, color = "gray50") +
  geom_point(aes(x = x1, y = y1),
             color = "#2C7BB6", size = 3, alpha = 0.6, shape = 16) +
  geom_point(aes(x = x2, y = y2),
             color = "#D7191C", size = 3, alpha = 0.6, shape = 17) +
  theme_minimal() +
  labs(title = "Procrustes Superimposition: PS1 (circles) vs PS2 (triangles)",
       subtitle = sprintf(
         "Correlation = %.3f (higher = better) | m² = %.4f (lower = better) | p = %.3f",
         pro_cor, pro_m2, pro_test$signif)) +
  coord_fixed()
```

```{r primer-effect-comparison, fig.width=8, fig.height=5}
r2_comparison <- data.frame(
  Factor = c("Species (PS1)", "Species (PS2)",
             "This Study: Primer Effect (unstratified)",
             "This Study: Primer Effect (stratified)",
             "Esnaola: Primer Effect", "Esnaola: Study Site"),
  R2 = c(species_r2_ps1, species_r2_ps2, primer_r2, primer_r2_strata,
         0.353, 0.184),
  Study = c("This Study", "This Study", "This Study", "This Study",
            "Esnaola 2018", "Esnaola 2018"),
  Type = c("Biological", "Biological", "Technical", "Technical",
           "Technical", "Biological")
)

ggplot(r2_comparison, aes(x = reorder(Factor, R2), y = R2, fill = Type)) +
  geom_col() +
  geom_hline(yintercept = 0.353, linetype = "dashed", color = "red",
             linewidth = 1) +
  annotate("text", x = 2.5, y = 0.37, label = "Esnaola primer threshold",
           size = 3, color = "red") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Effect Sizes: Biological vs Technical Variation",
       subtitle = sprintf("Our primer effect (R²=%.3f) is %.0f%% of Esnaola's",
                          primer_r2, (primer_r2/0.353)*100),
       x = "", y = "R² (Proportion of Variation Explained)") +
  scale_fill_manual(values = c("Biological" = "#2C7BB6",
                                "Technical" = "#D7191C"))
```

```{r primer-stats-table}
primer_stats <- data.frame(
  Metric = c(
    "Shared Samples (n)",
    "PS1: Species R²",
    "PS1: Species p-value",
    "PS2: Species R²",
    "PS2: Species p-value",
    "Mean Species R²",
    "Primer Effect R² (unstratified)",
    "Primer Effect p-value (unstratified)",
    "Primer Effect R² (stratified)",
    "Primer Effect p-value (stratified)",
    "Mantel correlation (r)",
    "Mantel p-value",
    "Procrustes correlation (high = good)",
    "Procrustes m² (low = good)",
    "Procrustes p-value"
  ),
  Value = c(
    length(shared_samples),
    round(species_r2_ps1, 3),
    format.pval(species_p_ps1, digits = 3),
    round(species_r2_ps2, 3),
    format.pval(species_p_ps2, digits = 3),
    round(mean(c(species_r2_ps1, species_r2_ps2)), 3),
    round(primer_r2, 3),
    format.pval(primer_p, digits = 3),
    round(primer_r2_strata, 3),
    format.pval(primer_p_strata, digits = 3),
    round(mantel_result$statistic, 3),
    format.pval(mantel_result$signif, digits = 3),
    round(pro_cor, 3),
    round(pro_m2, 4),
    format.pval(pro_test$signif, digits = 3)
  )
)

kable(primer_stats, caption = "Primer Set Concordance Statistics") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

## Compositional Differences vs. Variation Explained

### Conceptual Framework

Before interpreting the results, it's critical to distinguish between two related but
distinct concepts:

1. **Compositional differences (univariate)**: Taxon-specific absolute shifts in
   relative abundance between primer sets (e.g., "Hydroptilidae is 321× more abundant
   in PS1 than PS2"). Measured as fold-change ratios for individual taxa.

2. **Compositional variation explained (multivariate)**: The proportion of
   sample-to-sample variation in community composition attributable to primer choice
   (R² from PERMANOVA). Measures how much primers *differentially* affect samples.

**Why these can diverge:** If all samples experience the same absolute shift
(e.g., all samples show 10× more Hydroptilidae in PS1), the fold-change is huge
(10×) but the R² is low because the primer doesn't change the *relative differences*
between samples—the ecological structure (which samples are similar/different) is
preserved. The primers shift all samples in parallel, maintaining beta diversity
patterns.

Conversely, high R² with moderate fold-changes indicates that primers *differentially*
affect samples (Sample A shifts one way, Sample B shifts another), disrupting the
ecological signal.

```{r taxon-fold-changes}
cat("\n=== Taxon-Specific Compositional Differences ===\n\n")

# Calculate mean relative abundance by family for each primer set
# Transform to relative abundance
ps1_rel <- transform_sample_counts(ps1_shared, function(x) x / sum(x) * 100)
ps2_rel <- transform_sample_counts(ps2_shared, function(x) x / sum(x) * 100)

# Aggregate by family and calculate mean across all samples
ps1_family <- tax_glom(ps1_rel, taxrank = "Family", NArm = FALSE)
ps2_family <- tax_glom(ps2_rel, taxrank = "Family", NArm = FALSE)

# Get mean abundance per family
ps1_means <- apply(otu_table(ps1_family), 1, mean)
ps2_means <- apply(otu_table(ps2_family), 1, mean)

# Get family names
ps1_tax <- tax_table(ps1_family)[, "Family"]
ps2_tax <- tax_table(ps2_family)[, "Family"]

# Create dataframes
ps1_df <- data.frame(
  Family = as.character(ps1_tax),
  PS1_mean_pct = ps1_means,
  stringsAsFactors = FALSE
)

ps2_df <- data.frame(
  Family = as.character(ps2_tax),
  PS2_mean_pct = ps2_means,
  stringsAsFactors = FALSE
)

# Merge and calculate fold-changes
fc_data <- merge(ps1_df, ps2_df, by = "Family", all = TRUE)
fc_data$PS1_mean_pct[is.na(fc_data$PS1_mean_pct)] <- 0.001  # pseudocount
fc_data$PS2_mean_pct[is.na(fc_data$PS2_mean_pct)] <- 0.001  # pseudocount

fc_data$fold_change <- fc_data$PS1_mean_pct / fc_data$PS2_mean_pct
fc_data$log2_fc <- log2(fc_data$fold_change)
fc_data$abs_log2_fc <- abs(fc_data$log2_fc)

# Sort by absolute fold-change
fc_data <- fc_data[order(-fc_data$abs_log2_fc), ]

# Get top 10 families by abundance
top_families <- fc_data %>%
  mutate(max_abundance = pmax(PS1_mean_pct, PS2_mean_pct)) %>%
  arrange(desc(max_abundance)) %>%
  head(10)

cat(sprintf("Maximum fold-change observed: %.1f× (%s)\n",
            fc_data$fold_change[1], fc_data$Family[1]))
cat(sprintf("Number of families with >10× difference: %d\n",
            sum(fc_data$fold_change > 10 | fc_data$fold_change < 0.1)))
cat(sprintf("\n#### Top 10 most abundant families (fold-change = PS1/PS2):\n\n"))
kable(top_families[, c("Family", "PS1_mean_pct", "PS2_mean_pct",
                             "fold_change")],
            digits = c(0, 2, 2, 1),
            col.names = c("Family", "PS1 Mean %", "PS2 Mean %",
                          "Fold-Change"),
            caption = "Compositional Differences: Top 10 Families",
            booktabs = TRUE) %>%
      kable_styling(latex_options = c("striped", "HOLD_position"),
                    bootstrap_options = c("striped", "condensed"),
                    full_width = FALSE, font_size = 9)
```

```{r fold-change-plot, fig.width=10, fig.height=6}
# Create visualization showing fold-changes for top families
top_families_plot <- top_families %>%
  mutate(
    Direction = ifelse(fold_change > 1, "Higher in PS1", "Higher in PS2"),
    Family_label = paste0(Family, " (", round(fold_change, 1), "×)")
  )

ggplot(top_families_plot,
       aes(x = reorder(Family, abs_log2_fc), y = log2_fc, fill = Direction)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "solid", color = "black") +
  geom_hline(yintercept = c(-1, 1), linetype = "dashed", color = "gray50",
             alpha = 0.5) +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Taxon-Specific Compositional Differences Between Primer Sets",
    subtitle = sprintf(
      "Top 10 families by abundance | Max fold-change: %.1f× | Primer R² = %.3f",
      fc_data$fold_change[1], primer_r2_strata
    ),
    x = "Family",
    y = "Log2 Fold-Change (PS1/PS2)",
    caption = "Dashed lines: 2× fold-change threshold"
  ) +
  scale_fill_manual(values = c("Higher in PS1" = "#2C7BB6",
                               "Higher in PS2" = "#D7191C")) +
  theme(legend.position = "bottom")
```

```{r compositional-reconciliation, results='asis'}
cat("\n### Reconciling Large Fold-Changes with Low R²\n\n")

max_fc <- fc_data$fold_change[1]
max_fc_family <- fc_data$Family[1]

cat(sprintf("We observe substantial taxon-specific compositional differences between
primer sets, with a maximum fold-change of %.1f× for %s. Yet our stratified
PERMANOVA shows primer choice explains only R² = %.3f (%.1f%%) of sample-to-sample
variation.\n\n", max_fc, max_fc_family, primer_r2_strata,
primer_r2_strata * 100))

if (primer_r2_strata < 0.15) {
  cat("This apparent paradox—large univariate shifts but low multivariate R²—reveals
that the primers are shifting community composition in a **consistent, parallel
manner across all samples**. When all fish show similar directional shifts
(e.g., all samples gain Hydroptilidae and lose Chironomidae in the same proportions),
the absolute compositional changes are large, but the *relative ecological
relationships between samples* are preserved.\n\n")

  cat("This is actually **good news for cross-primer comparisons**: it means that
while the primers detect different absolute abundances, they maintain the
ecological structure (which species are similar vs. different). The biological
signal—species-level dietary differences—is robust to primer choice because the
primer bias affects all samples uniformly.\n\n")

  cat(sprintf("In contrast, Esnaola et al. (2018) found R² = 0.353 for primer
effects, suggesting their primers *differentially* affected samples in ways that
disrupted ecological patterns. Our low R² (%.3f) with high concordance
(Mantel r = %.3f, Procrustes r = %.3f) demonstrates primer compatibility despite
absolute compositional shifts.\n\n", primer_r2_strata, mantel_result$statistic,
pro_cor))
} else {
  cat("The combination of large fold-changes with moderate-to-high R² suggests that
primers are *differentially* affecting samples (some samples shift more than others),
which can disrupt ecological patterns. This indicates primer choice is a more
critical consideration for interpreting community structure.\n\n")
}
```

## Interpretation

```{r primer-interpretation, results='asis'}
# Generate interpretation based on actual results
cat("\n### Biological Signal Strength\n\n")

if (species_r2_ps1 > 0.353 && species_r2_ps2 > 0.353) {
  cat(sprintf("Both primer sets demonstrate strong biological signal: PS1 (R² = %.3f) and PS2 (R² = %.3f) both exceed the primer effect threshold (R² = 0.353) documented by Esnaola et al. (2018). This indicates that species-level dietary variation dominates over any potential technical variation introduced by primer choice.\n\n", species_r2_ps1, species_r2_ps2))
} else if (mean_r2 > 0.353) {
  cat(sprintf("On average, biological variation (mean R² = %.3f) exceeds the technical variation threshold from Esnaola et al. (R² = 0.353), though individual primer sets show variable strength (PS1: %.3f, PS2: %.3f).\n\n", mean_r2, species_r2_ps1, species_r2_ps2))
} else {
  cat(sprintf("Both primer sets show biological signal below Esnaola's primer effect threshold (PS1: %.3f, PS2: %.3f), suggesting moderate species-level dietary differentiation at this site.\n\n", species_r2_ps1, species_r2_ps2))
}

cat("### Primer Concordance\n\n")

# Determine concordance levels
mantel_level <- ifelse(mantel_result$statistic > 0.6, "strong",
                      ifelse(mantel_result$statistic > 0.4, "moderate", "weak"))
# For Procrustes: LOW m² = GOOD fit, HIGH correlation = GOOD concordance
procrustes_level <- ifelse(pro_cor > 0.8, "strong",
                          ifelse(pro_cor > 0.6, "moderate", "weak"))

cat(sprintf("The two primer sets show %s concordance in both distance-based (Mantel r = %.3f, p %s) and ordination-based (Procrustes correlation = %.3f, m² = %.4f, p %s) tests. ",
            mantel_level, mantel_result$statistic,
            format.pval(mantel_result$signif, digits = 3),
            pro_cor, pro_m2,
            format.pval(pro_test$signif, digits = 3)))

# Note: LOW m² means GOOD fit (0 = perfect superimposition)
if (mantel_result$statistic > 0.6 && pro_cor > 0.8) {
  cat("Both metrics indicate highly consistent dietary patterns despite sequencing different molecular markers, demonstrating that our results are robust to primer choice.\n\n")
} else if (mantel_result$statistic > 0.4 || pro_cor > 0.6) {
  cat("The primers detect reasonably consistent patterns, though some technical variation is evident.\n\n")
} else {
  cat("Concordance is limited, suggesting potential primer-specific biases that should be considered when interpreting results.\n\n")
}

cat("### Primer Effect Size\n\n")

cat(sprintf("Direct PERMANOVA testing shows that primer choice explains R² = %.3f (%.1f%%) of community variation (unstratified) and R² = %.3f (%.1f%%) when accounting for the paired design (stratified by sample). ",
            primer_r2, primer_r2 * 100, primer_r2_strata, primer_r2_strata * 100))

# Use stratified result for primary interpretation (accounts for paired design)
primer_effect_primary <- primer_r2_strata

if (primer_effect_primary < 0.353) {
  cat(sprintf("The stratified analysis (R² = %.3f) represents %.0f%% of the primer effect observed by Esnaola et al. (R² = 0.353), suggesting our primers show substantially less bias than the universal COI primers tested in their Mediterranean fish diet study. ",
              primer_effect_primary, (primer_effect_primary / 0.353) * 100))

  if (primer_effect_primary < mean_r2) {
    cat(sprintf("Importantly, our primer effect (R² = %.3f) is smaller than the species effect in both primer sets (mean R² = %.3f), demonstrating that biological signal dominates over technical variation.\n\n",
                primer_effect_primary, mean_r2))
  } else {
    cat(sprintf("However, this primer effect (R² = %.3f) is comparable to or exceeds the species effect (mean R² = %.3f), suggesting substantial technical variation that should be considered in interpretation.\n\n",
                primer_effect_primary, mean_r2))
  }
} else {
  cat(sprintf("This matches or exceeds Esnaola's primer effect (R² = 0.353), indicating substantial technical variation comparable to their study.\n\n"))
}

cat("### Contrast to Esnaola et al. (2018)\n\n")
cat(sprintf("Esnaola et al. reported that primer choice explained 35.3%% of variation (nearly twice their study site effect of 18.4%%). Our paired-sample design, where both primers are applied to identical biological material, yields a primer effect of R² = %.3f (%.1f%%) in the stratified analysis that accounts for paired samples. ",
            primer_r2_strata, primer_r2_strata * 100))

if (primer_r2_strata < 0.353 && mantel_result$statistic > 0.5) {
  cat(sprintf("The combination of lower primer effect size (%.1f%% of Esnaola's) and strong concordance (Mantel r = %.3f) suggests that our primers are more compatible than those tested by Esnaola, possibly due to better taxonomic coverage in our study system or reduced primer bias. ",
      (primer_r2_strata / 0.353) * 100, mantel_result$statistic))
  cat("Additionally, their large primer effect may have been partially confounded by uncontrolled differences in sample composition or environmental conditions between primer sets, which our paired design explicitly controls for.\n\n")
} else if (primer_r2_strata >= 0.353) {
  cat(sprintf("Like Esnaola et al., we observe substantial primer-related variation, reinforcing the importance of primer choice in metabarcoding studies. This primer effect (R² = %.3f) is comparable to Esnaola's finding, suggesting it may be an inherent challenge in multi-primer COI metabarcoding rather than a study-specific artifact.\n\n",
              primer_r2_strata))
} else {
  cat("While our primer effect is reduced compared to Esnaola's, the moderate concordance suggests that primer choice remains an important consideration in metabarcoding study design and interpretation.\n\n")
}
```

---

# LEVEL 1: Mormyrid Community at Balé Creek

**Biological context:** 9 mormyrid species, 35 individuals, single site. This level controls for geography and examines dietary differences among sympatric species.

## 1.1 Prey Composition Summary

Family-level prey composition across the 9 Balé Creek mormyrid species.

```{r level1-prey-composition}
cat("=== LEVEL 1.1: PREY COMPOSITION (FAMILY LEVEL) ===\n\n")

# Aggregate to family level
ps1_bale_family <- aggregate_to_family(ps1_bale)
ps2_bale_family <- aggregate_to_family(ps2_bale)

cat("Family-level aggregation:\n")
cat("  PS1:", ntaxa(ps1_bale_family), "families\n")
cat("  PS2:", ntaxa(ps2_bale_family), "families\n")

# Calculate summary statistics (top families)
# Convert to relative abundance and get mean ± SD per species
ps1_bale_rel <- transform_sample_counts(ps1_bale_family, function(x) x / sum(x) * 100)
ps2_bale_rel <- transform_sample_counts(ps2_bale_family, function(x) x / sum(x) * 100)

# Melt and summarize
calc_family_summary <- function(ps_rel) {
  ps_melt <- psmelt(ps_rel)
  ps_melt$Family_clean <- sapply(ps_melt$Family, clean_family_name)

  summary_table <- ps_melt %>%
    group_by(Family_clean) %>%
    summarise(
      mean_pct = mean(Abundance),
      sd_pct = sd(Abundance),
      n_samples = n(),
      .groups = "drop"
    ) %>%
    arrange(desc(mean_pct)) %>%
    mutate(mean_sd = paste0(round(mean_pct, 2), " ± ", round(sd_pct, 2)))

  return(summary_table)
}

summary_ps1 <- calc_family_summary(ps1_bale_rel)
summary_ps2 <- calc_family_summary(ps2_bale_rel)

cat("\nTop 10 prey families (Primer Set 1):\n")
print(head(summary_ps1, 10))

cat("\nTop 10 prey families (Primer Set 2):\n")
print(head(summary_ps2, 10))

# Identify top families across BOTH datasets for unified legend
combined_summary <- bind_rows(
  summary_ps1 %>% mutate(Dataset = "PS1"),
  summary_ps2 %>% mutate(Dataset = "PS2")
) %>%
  group_by(Family_clean) %>%
  summarise(Mean_Prop = mean(mean_pct), .groups = "drop") %>%
  arrange(desc(Mean_Prop)) %>%
  slice_head(n = 10)

top_families_unified <- combined_summary$Family_clean

# Ensure all top families have colors assigned in global palette
missing_families <- setdiff(top_families_unified, names(GLOBAL_FAMILY_COLORS))
if (length(missing_families) > 0) {
  cat("Adding colors for", length(missing_families), "families not in global palette:\n")
  cat("  ", paste(missing_families, collapse = ", "), "\n")

  # Get additional colors from the same palette used in create_global_family_palette
  distinct_colors <- c(
    "#4477AA", "#EE6677", "#228833", "#CCBB44", "#66CCEE", "#AA3377", "#BBBBBB",
    "#EE3377", "#009988", "#CC3311", "#EE7733", "#0077BB", "#33BBEE", "#EE3377",
    "#CC6677", "#882255", "#AA4499", "#117733", "#332288", "#44AA99", "#999933",
    "#DDCC77", "#661100", "#6699CC", "#AA4466", "#888888", "#771122", "#DD7788"
  )

  # Find colors not yet used
  used_colors <- GLOBAL_FAMILY_COLORS[names(GLOBAL_FAMILY_COLORS) != "Other"]
  available_colors <- setdiff(distinct_colors, used_colors)

  # Add colors for missing families
  new_colors <- setNames(
    available_colors[1:length(missing_families)],
    missing_families
  )
  GLOBAL_FAMILY_COLORS <- c(GLOBAL_FAMILY_COLORS, new_colors)
}

# Create modified plot function that uses pre-specified families
plot_composition_unified <- function(ps, group_var, families_to_show, title = "") {
  ps_melt <- psmelt(ps)
  ps_melt$Family_clean <- sapply(ps_melt$Family, clean_family_name)

  ps_agg <- ps_melt %>%
    group_by(.data[[group_var]], Family_clean) %>%
    summarise(Abundance = sum(Abundance), .groups = "drop")

  ps_agg <- ps_agg %>%
    group_by(.data[[group_var]]) %>%
    mutate(RelAbund = Abundance / sum(Abundance)) %>%
    ungroup()

  # Use provided families
  ps_final <- ps_agg %>%
    mutate(Family_display = ifelse(Family_clean %in% families_to_show,
                                   Family_clean, "Other")) %>%
    group_by(.data[[group_var]], Family_display) %>%
    summarise(RelAbund = sum(RelAbund), .groups = "drop")

  # Consistent factor levels across all plots
  all_levels <- c(families_to_show, "Other")
  ps_final$Family_display <- factor(ps_final$Family_display,
                                     levels = all_levels)

  p <- ggplot(ps_final, aes(x = .data[[group_var]], y = RelAbund,
                            fill = Family_display)) +
    geom_bar(stat = "identity", position = "stack") +
    scale_fill_manual(values = GLOBAL_FAMILY_COLORS,
                      breaks = all_levels,
                      name = "Prey Family",
                      na.value = "#808080",
                      drop = FALSE) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = title, x = group_var, y = "Relative Abundance") +
    scale_y_continuous(labels = scales::percent)

  return(p)
}

# Create stacked bar plots with unified family list
p1 <- plot_composition_unified(ps1_bale_family, group_var = "sample_Species",
                               families_to_show = top_families_unified,
                               title = "Prey Composition - Balé Creek (PS1)")

p2 <- plot_composition_unified(ps2_bale_family, group_var = "sample_Species",
                               families_to_show = top_families_unified,
                               title = "Prey Composition - Balé Creek (PS2)")

# Display side by side with unified legend
# Manually create unified legend since patchwork guides="collect" isn't working
library(cowplot)

# Remove legends from both plots
p1_no_legend <- p1 + theme(legend.position = "none")
p2_no_legend <- p2 + theme(legend.position = "none")

# Create a dummy plot with ALL families for legend extraction
# Combine data from both plots to ensure all families appear
dummy_data <- data.frame(
  x = rep("A", length(top_families_unified) + 1),
  y = rep(1, length(top_families_unified) + 1),
  Family = factor(c(top_families_unified, "Other"),
                  levels = c(top_families_unified, "Other"))
)

p_dummy <- ggplot(dummy_data, aes(x = x, y = y, fill = Family)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = GLOBAL_FAMILY_COLORS,
                    breaks = c(top_families_unified, "Other"),
                    name = "Prey Family",
                    drop = FALSE) +
  theme(legend.position = "bottom",
        legend.key.size = unit(0.4, "cm")) +
  guides(fill = guide_legend(ncol = 5, byrow = TRUE, title = "Prey Family"))

# Extract legend from dummy plot (has all families)
legend <- get_legend(p_dummy)

# Combine plots and legend
plot_grid(
  plot_grid(p1_no_legend, p2_no_legend, ncol = 2),
  legend,
  ncol = 1,
  rel_heights = c(1, 0.15)
)
```

Top prey families table:

```{r level1-top-families-table}
# Combine top families from both primer sets
top_families_combined <- full_join(
  summary_ps1 %>% select(Family_clean, mean_pct, sd_pct) %>% slice_head(n = 15),
  summary_ps2 %>% select(Family_clean, mean_pct, sd_pct) %>% slice_head(n = 15),
  by = "Family_clean",
  suffix = c("_PS1", "_PS2")
) %>%
  arrange(desc(pmax(mean_pct_PS1, mean_pct_PS2, na.rm = TRUE)))

kable(top_families_combined,
      caption = "Top 15 prey families at Balé Creek (mean $\\pm$ SD relative abundance \\%)",
      digits = 2,
      col.names = c("Family", "Mean \\% (PS1)", "SD (PS1)", "Mean \\% (PS2)", "SD (PS2)"),
      booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 1.2 Alpha Diversity Among Species

Shannon diversity analysis across Balé Creek species with cross-primer validation.

```{r level1-alpha-diversity, fig.width=12, fig.height=6}
cat("=== LEVEL 1.2: ALPHA DIVERSITY AMONG SPECIES ===\n\n")

# Analyze alpha diversity for both primer sets
alpha1_bale <- analyze_alpha_diversity(ps1_bale, "Species")
alpha2_bale <- analyze_alpha_diversity(ps2_bale, "Species")

cat("\n=== PRIMERSET 1 - Balé Creek Shannon Diversity ===\n")
cat("Test type:", alpha1_bale$test_type, "\n")
cat("Shapiro-Wilk p-value:", round(alpha1_bale$shapiro_p, 4), "\n")
if (alpha1_bale$test_type == "Kruskal-Wallis") {
  print(alpha1_bale$test)
} else {
  print(summary(alpha1_bale$test))
}

cat("\n=== PRIMERSET 2 - Balé Creek Shannon Diversity ===\n")
cat("Test type:", alpha2_bale$test_type, "\n")
cat("Shapiro-Wilk p-value:", round(alpha2_bale$shapiro_p, 4), "\n")
if (alpha2_bale$test_type == "Kruskal-Wallis") {
  print(alpha2_bale$test)
} else {
  print(summary(alpha2_bale$test))
}

# Comparative plot
p_bale_alpha_comp <- plot_comparative_shannon(alpha1_bale$diversity,
                                              alpha2_bale$diversity,
                                              "Species",
                                              "Balé Creek Species Diet Diversity Comparison")
print(p_bale_alpha_comp)
```

### Cross-Primer Correlation for Alpha Diversity

```{r level1-alpha-correlation, fig.width=8, fig.height=6}
# Correlate Shannon values for shared samples
cor_bale_alpha <- correlate_shared_samples(alpha1_bale$diversity,
                                          alpha2_bale$diversity)

cat("\n=== BALÉ CREEK ALPHA DIVERSITY CORRELATION ===\n")
cat("Shared samples:", cor_bale_alpha$n_shared, "\n")

if (!is.null(cor_bale_alpha$cor_test)) {
  cat("Pearson r:", round(cor_bale_alpha$cor_test$estimate, 3), "\n")
  cat("95% CI: [", round(cor_bale_alpha$cor_test$conf.int[1], 3), ",",
      round(cor_bale_alpha$cor_test$conf.int[2], 3), "]\n")
  cat("P-value:", format.pval(cor_bale_alpha$cor_test$p.value, digits = 3), "\n")

  # Scatterplot - get shared samples
  shared_ids <- intersect(alpha1_bale$diversity$Sample_Name,
                          alpha2_bale$diversity$Sample_Name)

  # Filter to shared samples and match order
  div1_shared <- alpha1_bale$diversity %>%
    filter(Sample_Name %in% shared_ids) %>%
    arrange(Sample_Name)
  div2_shared <- alpha2_bale$diversity %>%
    filter(Sample_Name %in% shared_ids) %>%
    arrange(Sample_Name)

  cor_df <- data.frame(
    PS1_Shannon = div1_shared$Shannon,
    PS2_Shannon = div2_shared$Shannon
  )

  p_alpha_cor <- ggplot(cor_df, aes(x = PS1_Shannon, y = PS2_Shannon)) +
    geom_point(size = 3, alpha = 0.6, color = ps1_color) +
    geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
    geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "gray50") +
    theme_minimal() +
    labs(title = "Balé Creek: Shannon Diversity Correlation",
         subtitle = paste0("r = ", round(cor_bale_alpha$cor_test$estimate, 3),
                          ", p ", format.pval(cor_bale_alpha$cor_test$p.value, digits = 2)),
         x = "Primerset 1 Shannon", y = "Primerset 2 Shannon") +
    coord_fixed()

  print(p_alpha_cor)
} else {
  cat(cor_bale_alpha$message, "\n")
}
```

## 1.3 Beta Diversity Among Species

Bray-Curtis dissimilarity analysis with PERMANOVA testing for dietary differences among species.

```{r level1-beta-diversity, fig.width=14, fig.height=6}
cat("\n=== LEVEL 1.3: BETA DIVERSITY AMONG SPECIES ===\n\n")

# Filter species with < 3 samples for stable PERMANOVA
cat("=== PRIMERSET 1 FILTERING ===\n")
ps1_bale_filt <- filter_low_n_species(ps1_bale, min_n = 3)

cat("\n=== PRIMERSET 2 FILTERING ===\n")
ps2_bale_filt <- filter_low_n_species(ps2_bale, min_n = 3)

# Run beta diversity analyses
beta1_bale <- run_beta_diversity_analysis(ps1_bale_filt, "~ Species")
beta2_bale <- run_beta_diversity_analysis(ps2_bale_filt, "~ Species")

cat("\n=== PRIMERSET 1 - Balé Creek PERMANOVA ===\n")
print(beta1_bale$permanova)

cat("\n=== PRIMERSET 2 - Balé Creek PERMANOVA ===\n")
print(beta2_bale$permanova)

# Faceted PCoA plot
p_bale_pcoa <- plot_faceted_pcoa(beta1_bale, beta2_bale,
                                 color_var = "Species",
                                 title = "Balé Creek Diet Composition by Species")
print(p_bale_pcoa)
```

### Pairwise PERMANOVA Tests

```{r level1-beta-pairwise}
# Pairwise tests if more than 2 species
n_species_ps1 <- length(unique(sample_data(ps1_bale_filt)$Species))
n_species_ps2 <- length(unique(sample_data(ps2_bale_filt)$Species))

cat("\n=== PAIRWISE PERMANOVA TESTS ===\n")

if (n_species_ps1 > 2) {
  cat("\n#### PRIMERSET 1:\n\n")
  pairwise1_bale <- pairwise.adonis2(beta1_bale$distance ~ Species,
                                     data = beta1_bale$metadata)
  # Sanitize row names for LaTeX (replace _ with space)
  pairwise1_df <- as.data.frame(pairwise1_bale)
  rownames(pairwise1_df) <- gsub("_", " vs ", rownames(pairwise1_df))
  # Select only key columns to fit on page
  key_cols <- intersect(colnames(pairwise1_df), c("R2", "F.Model", "Pr(>F)"))
  if (length(key_cols) == 0) {
    # If those columns don't exist, take first 3 numeric columns plus p-value
    key_cols <- colnames(pairwise1_df)[1:min(4, ncol(pairwise1_df))]
  }
  kable(pairwise1_df[, key_cols, drop = FALSE],
              digits = 3,
              booktabs = TRUE,
              caption = "Pairwise PERMANOVA: Primerset 1") %>%
        kable_styling(latex_options = c("striped", "HOLD_position"),
                      font_size = 9,
                      full_width = FALSE)
}

if (n_species_ps2 > 2) {
  cat("\n#### PRIMERSET 2:\n\n")
  pairwise2_bale <- pairwise.adonis2(beta2_bale$distance ~ Species,
                                     data = beta2_bale$metadata)
  # Sanitize row names for LaTeX (replace _ with space)
  pairwise2_df <- as.data.frame(pairwise2_bale)
  rownames(pairwise2_df) <- gsub("_", " vs ", rownames(pairwise2_df))
  # Select only key columns to fit on page
  key_cols <- intersect(colnames(pairwise2_df), c("R2", "F.Model", "Pr(>F)"))
  if (length(key_cols) == 0) {
    # If those columns don't exist, take first 3 numeric columns plus p-value
    key_cols <- colnames(pairwise2_df)[1:min(4, ncol(pairwise2_df))]
  }
  kable(pairwise2_df[, key_cols, drop = FALSE],
              digits = 3,
              booktabs = TRUE,
              caption = "Pairwise PERMANOVA: Primerset 2") %>%
        kable_styling(latex_options = c("striped", "HOLD_position"),
                      font_size = 9,
                      full_width = FALSE)
}
```

### Beta Diversity Cross-Validation (Mantel Test)

```{r level1-beta-mantel}
# Find shared Sample_Names between datasets
samples1_bale <- beta1_bale$metadata$Sample_Name
samples2_bale <- beta2_bale$metadata$Sample_Name
shared_bale <- intersect(samples1_bale, samples2_bale)

cat("\n=== BALÉ CREEK BETA DIVERSITY MANTEL TEST ===\n")
cat("Shared Sample_Names:", length(shared_bale), "\n")

if (length(shared_bale) >= 3) {
  # Convert distance matrices to matrices with Sample_Name labels
  dist1_mat <- as.matrix(beta1_bale$distance)
  dist2_mat <- as.matrix(beta2_bale$distance)

  # Relabel with Sample_Names
  rownames(dist1_mat) <- beta1_bale$metadata$Sample_Name
  colnames(dist1_mat) <- beta1_bale$metadata$Sample_Name
  rownames(dist2_mat) <- beta2_bale$metadata$Sample_Name
  colnames(dist2_mat) <- beta2_bale$metadata$Sample_Name

  # Subset to shared Sample_Names (automatically aligned)
  dist1_shared <- as.dist(dist1_mat[shared_bale, shared_bale])
  dist2_shared <- as.dist(dist2_mat[shared_bale, shared_bale])

  # Mantel test
  mantel_bale <- run_mantel_test(dist1_shared, dist2_shared)

  cat("\nMantel statistic (r):", round(mantel_bale$statistic, 3), "\n")
  cat("P-value:", format.pval(mantel_bale$signif, digits = 3), "\n")
  cat("Interpretation:",
      ifelse(mantel_bale$statistic > 0.5, "Strong concordance",
             ifelse(mantel_bale$statistic > 0.3, "Moderate concordance",
                    "Weak concordance")), "\n")
} else {
  cat("Insufficient shared samples for Mantel test.\n")
  mantel_bale <- NULL
}
```

## 1.4 Diet Overlap Indices

Schoener and Pianka diet overlap indices quantify dietary similarity among species. We test whether overlap is higher within *Paramormyrops* species than between genera.

```{r level1-diet-overlap}
cat("\n=== LEVEL 1.4: DIET OVERLAP INDICES ===\n\n")

# Function to calculate diet overlap for a phyloseq object
calculate_diet_overlap_bale <- function(ps_bale, primerset_name) {

  cat("\n=== ", primerset_name, " - Diet Overlap ===\n", sep = "")

  # Get OTU and taxonomy
  otu_mat <- t(as.matrix(otu_table(ps_bale)))
  tax_df <- as.data.frame(tax_table(ps_bale))
  meta <- data.frame(sample_data(ps_bale), stringsAsFactors = FALSE)

  # Create taxonomic label (Order level)
  tax_df$TaxLabel <- ifelse(!is.na(tax_df$Order) & tax_df$Order != "",
                           tax_df$Order,
                           ifelse(!is.na(tax_df$Family) & tax_df$Family != "",
                                 tax_df$Family,
                                 ifelse(!is.na(tax_df$Class) & tax_df$Class != "",
                                       tax_df$Class, "Unclassified")))

  # Aggregate by taxonomic group
  otu_df <- as.data.frame(otu_mat)
  otu_tax <- data.frame(row.names = rownames(otu_df))

  for (taxon in unique(tax_df$TaxLabel)) {
    asvs_in_taxon <- rownames(tax_df)[tax_df$TaxLabel == taxon]
    asvs_present <- asvs_in_taxon[asvs_in_taxon %in% colnames(otu_df)]

    if (length(asvs_present) > 0) {
      if (length(asvs_present) == 1) {
        otu_tax[[taxon]] <- otu_df[[asvs_present]]
      } else {
        otu_tax[[taxon]] <- rowSums(otu_df[, asvs_present, drop = FALSE])
      }
    }
  }

  # Aggregate by species
  otu_agg <- aggregate(otu_tax, by = list(Species = meta$Species), FUN = sum)
  rownames(otu_agg) <- otu_agg$Species
  otu_agg <- otu_agg[, -1]

  # Check number of species
  n_species <- nrow(otu_agg)
  cat("Number of species:", n_species, "\n")

  # Need at least 2 species for pairwise overlap
  if (n_species < 2) {
    cat("Warning: Only", n_species, "species found. Cannot calculate pairwise overlap.\n")
    return(list(schoener = NA, pianka = NA, taxa = unique(tax_df$TaxLabel),
                species_list = rownames(otu_agg)))
  }

  # Convert to matrix for proper handling
  otu_mat_agg <- as.matrix(otu_agg)

  # Check for empty rows
  row_sums <- rowSums(otu_mat_agg)
  cat("Row sums range:", min(row_sums), "-", max(row_sums), "\n")

  if (any(row_sums == 0)) {
    cat("Warning: Found", sum(row_sums == 0), "species with zero reads\n")
    otu_mat_agg <- otu_mat_agg[row_sums > 0, , drop = FALSE]
    cat("Removing empty species. New count:", nrow(otu_mat_agg), "\n")

    if (nrow(otu_mat_agg) < 2) {
      cat("Warning: Less than 2 species with data after filtering.\n")
      return(list(schoener = NA, pianka = NA, taxa = unique(tax_df$TaxLabel),
                  species_list = rownames(otu_mat_agg)))
    }
  }

  # Check for empty columns (taxonomic groups with no reads)
  col_sums <- colSums(otu_mat_agg)
  if (any(col_sums == 0)) {
    cat("Warning: Found", sum(col_sums == 0), "taxa with zero reads\n")
    otu_mat_agg <- otu_mat_agg[, col_sums > 0, drop = FALSE]
    cat("Removing empty taxa. New count:", ncol(otu_mat_agg), "\n")
  }

  # Convert to proportions
  otu_prop <- sweep(otu_mat_agg, 1, rowSums(otu_mat_agg), "/")

  # Check for NaN in proportions
  if (any(is.nan(otu_prop))) {
    cat("Warning: NaN values in proportions matrix\n")
  }

  cat("Proportions matrix dimensions:", nrow(otu_prop), "x", ncol(otu_prop), "\n")

  # Calculate overlap indices
  # NOTE: niche.overlap expects species in COLUMNS, so transpose
  schoener <- niche.overlap(t(otu_prop), method = "schoener")
  pianka <- niche.overlap(t(otu_prop), method = "pianka")

  # CRITICAL: niche.overlap returns a dist-like object, convert to matrix
  species_names <- rownames(otu_prop)
  schoener <- as.matrix(schoener)
  pianka <- as.matrix(pianka)

  # Now assign species names as dimnames
  rownames(schoener) <- species_names
  colnames(schoener) <- species_names
  rownames(pianka) <- species_names
  colnames(pianka) <- species_names

  # Check if matrices have valid values
  cat("Schoener matrix dimensions:", nrow(schoener), "x", ncol(schoener), "\n")
  cat("Schoener matrix range:", min(schoener, na.rm = TRUE), "-",
      max(schoener, na.rm = TRUE), "\n")

  # Extract lower triangle for pairwise comparisons
  schoener_vals <- schoener[lower.tri(schoener)]
  pianka_vals <- pianka[lower.tri(pianka)]

  cat("Schoener overlap (mean ± SD):",
      round(mean(schoener_vals, na.rm = TRUE), 3), "±",
      round(sd(schoener_vals, na.rm = TRUE), 3), "\n")

  cat("Pianka overlap (mean ± SD):",
      round(mean(pianka_vals, na.rm = TRUE), 3), "±",
      round(sd(pianka_vals, na.rm = TRUE), 3), "\n")

  return(list(schoener = schoener,
              pianka = pianka,
              taxa = unique(tax_df$TaxLabel),
              species_list = species_names))
}

# Calculate for both primer sets
overlap_ps1 <- calculate_diet_overlap_bale(ps1_bale, "PRIMERSET 1")
overlap_ps2 <- calculate_diet_overlap_bale(ps2_bale, "PRIMERSET 2")

# Taxonomic overlap
shared_taxa <- intersect(overlap_ps1$taxa, overlap_ps2$taxa)
cat("\n=== TAXONOMIC OVERLAP ===\n")
cat("PS1 unique Orders:", length(overlap_ps1$taxa), "\n")
cat("PS2 unique Orders:", length(overlap_ps2$taxa), "\n")
cat("Shared Orders:", length(shared_taxa), "\n")
cat("Jaccard similarity:",
    round(length(shared_taxa) /
          length(union(overlap_ps1$taxa, overlap_ps2$taxa)), 3), "\n")
```

### Within-Paramormyrops vs Between-Genus Overlap Test

Test whether *Paramormyrops* species show higher dietary overlap with each other than with other genera.

```{r level1-overlap-test}
cat("\n=== WITHIN-PARAMORMYROPS VS. BETWEEN-GENUS OVERLAP TEST ===\n\n")

# Function to categorize overlap comparisons
test_paramormyrops_overlap <- function(overlap_matrix, species_list, primerset_name) {

  cat("\n", primerset_name, ":\n", sep = "")

  if (is.na(overlap_matrix[1]) || is.null(overlap_matrix)) {
    cat("No valid overlap data\n")
    return(NULL)
  }

  # Identify Paramormyrops species
  is_paramormyrops <- grepl("Paramormyrops", species_list)
  param_species <- species_list[is_paramormyrops]

  cat("Paramormyrops species:", length(param_species), "\n")
  cat("Other species:", sum(!is_paramormyrops), "\n")

  # Extract overlap values
  within_param <- c()
  between_genus <- c()

  for (i in 1:(length(species_list)-1)) {
    for (j in (i+1):length(species_list)) {
      sp1 <- species_list[i]
      sp2 <- species_list[j]
      overlap_val <- overlap_matrix[i, j]

      # Both are Paramormyrops
      if (is_paramormyrops[i] && is_paramormyrops[j]) {
        within_param <- c(within_param, overlap_val)
      }
      # One is Paramormyrops, one is not
      else if (is_paramormyrops[i] != is_paramormyrops[j]) {
        between_genus <- c(between_genus, overlap_val)
      }
      # Neither is Paramormyrops - skip
    }
  }

  cat("Within-Paramormyrops comparisons:", length(within_param), "\n")
  cat("Between-genus comparisons:", length(between_genus), "\n")

  if (length(within_param) > 0 && length(between_genus) > 0) {
    cat("\nWithin-Paramormyrops overlap:", round(mean(within_param), 3),
        "±", round(sd(within_param), 3), "\n")
    cat("Between-genus overlap:", round(mean(between_genus), 3),
        "±", round(sd(between_genus), 3), "\n")

    # Wilcoxon test
    wilcox_result <- wilcox.test(within_param, between_genus,
                                  alternative = "greater")

    cat("\nWilcoxon rank-sum test (one-sided):\n")
    cat("W =", wilcox_result$statistic, "\n")
    cat("p-value =", format.pval(wilcox_result$p.value, digits = 3), "\n")
    cat("Interpretation:",
        ifelse(wilcox_result$p.value < 0.05,
               "Within-Paramormyrops overlap is significantly higher",
               "No significant difference in overlap"), "\n")

    return(list(within = within_param,
                between = between_genus,
                test = wilcox_result))
  } else {
    cat("Insufficient comparisons for statistical test\n")
    return(NULL)
  }
}

# Test for both primer sets (using Schoener index)
test_ps1 <- test_paramormyrops_overlap(overlap_ps1$schoener,
                                       overlap_ps1$species_list,
                                       "PRIMERSET 1")
test_ps2 <- test_paramormyrops_overlap(overlap_ps2$schoener,
                                       overlap_ps2$species_list,
                                       "PRIMERSET 2")
```

### Diet Overlap Heatmaps

```{r level1-overlap-heatmaps, fig.width=10, fig.height=8}
library(pheatmap)

# Function to create heatmap for diet overlap
plot_overlap_heatmap <- function(overlap_matrix, title, primerset) {

  if (is.na(overlap_matrix[1]) || is.null(overlap_matrix)) {
    cat("No valid data for", title, "-", primerset, "\n")
    return(NULL)
  }

  if (is.null(rownames(overlap_matrix)) || is.null(colnames(overlap_matrix))) {
    cat("Warning: Missing dimnames for", title, "-", primerset, "\n")
    return(NULL)
  }

  p <- pheatmap(overlap_matrix,
                cluster_rows = TRUE,
                cluster_cols = TRUE,
                display_numbers = TRUE,
                number_format = "%.2f",
                color = colorRampPalette(c("white", "yellow", "orange", "red"))(100),
                main = paste0(title, " (", primerset, ")"),
                fontsize_number = 8,
                fontsize_row = 10,
                fontsize_col = 10,
                angle_col = 45,
                cellwidth = 35,
                cellheight = 35,
                breaks = seq(0, 1, length.out = 101))

  return(p)
}

# Create heatmaps for Primerset 1
if (!is.na(overlap_ps1$schoener[1])) {
  plot_overlap_heatmap(overlap_ps1$schoener,
                       "Schoener Diet Overlap",
                       "Primerset 1")
  plot_overlap_heatmap(overlap_ps1$pianka,
                       "Pianka Diet Overlap",
                       "Primerset 1")
}

# Create heatmaps for Primerset 2
if (!is.na(overlap_ps2$schoener[1])) {
  plot_overlap_heatmap(overlap_ps2$schoener,
                       "Schoener Diet Overlap",
                       "Primerset 2")
  plot_overlap_heatmap(overlap_ps2$pianka,
                       "Pianka Diet Overlap",
                       "Primerset 2")
}
```

## 1.5 Phylogenetic Correlation Test

Test whether dietary dissimilarity is correlated with phylogenetic distance (Mantel test).

```{r level1-phylo-correlation}
cat("\n=== LEVEL 1.5: PHYLOGENETIC CORRELATION TEST ===\n\n")

# Function to perform phylogenetic-dietary correlation
test_phylo_diet_correlation <- function(beta_result, phylo_dist_matrix,
                                        phylo_tree_obj, primerset_name) {

  cat("\n", primerset_name, ":\n", sep = "")

  # Get species from beta diversity metadata
  meta <- beta_result$metadata
  species_in_data <- unique(meta$Species)

  cat("Species in dietary data:", length(species_in_data), "\n")

  # Create a temporary phyloseq object to use the matching function
  # We just need the metadata, so create a minimal phyloseq object
  temp_otu <- otu_table(matrix(1, nrow = 1, ncol = nrow(meta)),
                        taxa_are_rows = TRUE)
  colnames(temp_otu) <- rownames(meta)
  temp_meta <- sample_data(meta)
  sample_names(temp_meta) <- rownames(meta)
  temp_ps <- phyloseq(temp_otu, temp_meta)

  # Use the fuzzy matching function to map species to tree tips
  match_results <- match_bale_species_to_phylogeny(temp_ps, phylo_tree_obj)

  cat("Species matched to phylogeny:", match_results$n_matched, "/",
      match_results$n_species, "\n")

  if (match_results$n_matched < 3) {
    cat("Insufficient species overlap with phylogeny (need >= 3)\n")
    return(NULL)
  }

  # Create species-to-tip mapping for matched species
  match_table <- match_results$match_table
  match_table <- match_table[!is.na(match_table$tree_tip), ]

  # Get species that have matches
  shared_species <- match_table$bale_species
  cat("Shared species:", length(shared_species), "\n")
  cat("  ", paste(shared_species, collapse = ", "), "\n")

  # Create species-level dietary distance matrix
  # Average dietary distance between all samples of each species pair
  diet_dist_species <- matrix(0, nrow = length(shared_species),
                              ncol = length(shared_species))
  rownames(diet_dist_species) <- shared_species
  colnames(diet_dist_species) <- shared_species

  for (i in 1:length(shared_species)) {
    for (j in 1:length(shared_species)) {
      if (i != j) {
        # Get samples for species i and j
        samples_i <- rownames(meta)[meta$Species == shared_species[i]]
        samples_j <- rownames(meta)[meta$Species == shared_species[j]]

        # Extract pairwise distances from distance matrix
        dist_mat <- as.matrix(beta_result$distance)
        pairwise_dists <- dist_mat[samples_i, samples_j]

        # Average distance between species
        diet_dist_species[i, j] <- mean(pairwise_dists, na.rm = TRUE)
      }
    }
  }

  # Convert to dist object
  diet_dist_species <- as.dist(diet_dist_species)

  # Create phylogenetic distance matrix using tree tips
  tree_tips <- match_table$tree_tip
  names(tree_tips) <- match_table$bale_species

  # Extract phylogenetic distances for matched tips
  phylo_dist_tips <- phylo_dist_matrix[tree_tips, tree_tips]

  # Rename to species names for consistency
  rownames(phylo_dist_tips) <- names(tree_tips)
  colnames(phylo_dist_tips) <- names(tree_tips)

  # Ensure same order as diet_dist_species
  phylo_dist_tips <- phylo_dist_tips[shared_species, shared_species]
  phylo_dist_shared <- as.dist(phylo_dist_tips)

  # Mantel test
  mantel_result <- vegan::mantel(diet_dist_species, phylo_dist_shared,
                                 method = "pearson", permutations = 999)

  cat("\nMantel Test Results:\n")
  cat("Mantel statistic (r):", round(mantel_result$statistic, 3), "\n")
  cat("P-value:", format.pval(mantel_result$signif, digits = 3), "\n")
  cat("Interpretation:",
      ifelse(mantel_result$signif < 0.05,
             "Significant phylogenetic signal in diet",
             "No significant phylogenetic signal in diet"), "\n")

  return(list(
    mantel = mantel_result,
    diet_dist = diet_dist_species,
    phylo_dist = phylo_dist_shared,
    shared_species = shared_species,
    match_table = match_table
  ))
}

# Test for both primer sets (if phylogeny is available)
if (!is.null(phylo_dist) && !is.null(phylo_tree)) {
  phylo_cor_ps1 <- test_phylo_diet_correlation(beta1_bale, phylo_dist,
                                                phylo_tree, "PRIMERSET 1")
  phylo_cor_ps2 <- test_phylo_diet_correlation(beta2_bale, phylo_dist,
                                                phylo_tree, "PRIMERSET 2")
} else {
  cat("Phylogenetic data not available. Skipping analysis.\n")
  phylo_cor_ps1 <- NULL
  phylo_cor_ps2 <- NULL
}
```

## 1.6 SIMPER Analysis - Species

Similarity Percentage (SIMPER) analysis identifies which prey taxa contribute most to dietary differences among species.

```{r level1-simper, results='asis',out.width="80%"}
cat("\n=== LEVEL 1.6: SIMPER ANALYSIS - SPECIES ===\n\n")

# Function to run and summarize SIMPER
run_simper_analysis <- function(ps, primerset_name, top_n = 10,
                                detailed_output = FALSE) {

  cat("\n", primerset_name, ":\n", sep = "")

  # Get OTU table (samples as rows, taxa as columns)
  otu_mat <- as.matrix(otu_table(ps))
  if (taxa_are_rows(ps)) {
    otu_mat <- t(otu_mat)
  }

  # Get metadata
  meta <- data.frame(sample_data(ps))

  # Run SIMPER
  simper_result <- vegan::simper(otu_mat, meta$Species, permutations = 999)

  cat("\nSIMPER analysis complete\n")
  cat("Number of species comparisons:", length(simper_result), "\n")

  # Extract top contributors for each comparison
  simper_summary <- summary(simper_result)

  # Create consolidated heatmap or detailed tables based on parameter
  if (!detailed_output) {
    cat("\nCreating consolidated heatmap of top prey families...\n")
    heatmap_plot <- plot_simper_heatmap(simper_result, ps,
                                        top_n = 15,
                                        primerset_name = primerset_name)
    print(heatmap_plot)
    cat("\nDetailed tables suppressed. Set detailed_output = TRUE to view.\n\n")
  } else {
    cat("\nTop", top_n, "prey taxa contributing to species differences:\n\n")

    # Print summary for each pairwise comparison
    for (comparison_name in names(simper_summary)) {
    # Sanitize comparison name for LaTeX
    comparison_name_clean <- gsub("_", " vs ", comparison_name)
    cat("###", comparison_name_clean, "\n\n")

    comparison_data <- simper_summary[[comparison_name]]

    # Check structure and convert if needed
    if (!is.data.frame(comparison_data)) {
      comparison_data <- as.data.frame(comparison_data)
    }

    # Skip if no data
    if (nrow(comparison_data) == 0) {
      cat("Warning: No data for this comparison. Skipping.\n\n")
      next
    }

    # Add contr column (use 'average' which is the contribution to dissimilarity)
    # The cumsum column already exists from vegan's summary()
    comparison_data$contr <- comparison_data$average

    # Filter to cumulative contribution < 70%
    # Note: cumsum is already calculated by vegan, but it's as a proportion
    # We need to recalculate based on percentage contribution
    total_dissim <- sum(comparison_data$average, na.rm = TRUE)
    comparison_data$pct_contrib <- (comparison_data$average / total_dissim) * 100
    comparison_data$cumsum_pct <- cumsum(comparison_data$pct_contrib)

    top_taxa <- comparison_data[comparison_data$cumsum_pct <= 70, ]

    # If nothing meets 70%, take top 10
    if (nrow(top_taxa) == 0) {
      top_taxa <- head(comparison_data, top_n)
    }

    # Get taxonomy for these ASVs
    tax_df <- as.data.frame(tax_table(ps))
    top_taxa$ASV <- rownames(top_taxa)

    # Add taxonomic names
    top_taxa$Family <- sapply(top_taxa$ASV, function(asv) {
      if (asv %in% rownames(tax_df)) {
        fam <- tax_df[asv, "Family"]
        if (is.na(fam) || fam == "") {
          return(tax_df[asv, "Order"])
        }
        return(clean_family_name(fam))
      }
      return("Unknown")
    })

    # Aggregate ASVs by Family for plotting
    family_aggregated <- top_taxa %>%
      group_by(Family) %>%
      summarise(
        pct_contrib = sum(pct_contrib),
        n_asvs = n(),
        .groups = "drop"
      ) %>%
      arrange(desc(pct_contrib))

    # Select columns for display (ASV-level detail)
    display_table <- top_taxa[, c("Family", "average", "sd", "ratio",
                                  "ava", "avb", "pct_contrib",
                                  "cumsum_pct")]

    # Rename for clarity
    colnames(display_table)[colnames(display_table) == "pct_contrib"] <- "% Contrib"
    colnames(display_table)[colnames(display_table) == "cumsum_pct"] <- "Cum %"

    # Round numeric columns
    display_table[, c("average", "sd", "ratio", "ava", "avb",
                      "% Contrib", "Cum %")] <-
      round(display_table[, c("average", "sd", "ratio", "ava", "avb",
                              "% Contrib", "Cum %")], 3)

    cat(paste0("\n#### SIMPER: ", comparison_name_clean, "\n\n"))
    # Limit to top 10 rows for space
    display_table_top <- head(display_table, 10)
    kable(display_table_top,
                caption = paste("SIMPER:", comparison_name_clean, "(Top 10)"),
                row.names = FALSE,
                booktabs = TRUE) %>%
            kable_styling(latex_options = c("striped", "HOLD_position"),
                          bootstrap_options = c("striped", "hover"),
                          full_width = FALSE, font_size = 8)

    # Create family-aggregated plot
    plot_data <- head(family_aggregated, 15)
    plot_data$Family <- factor(plot_data$Family,
                               levels = rev(plot_data$Family))

    p <- ggplot(plot_data, aes(x = pct_contrib, y = Family)) +
      geom_bar(stat = "identity", fill = "#0072B2", alpha = 0.7) +
      theme_minimal() +
      labs(title = paste0("Top Prey Families: ", comparison_name_clean,
                          " (", primerset_name, ")"),
           x = "% Contribution to Dissimilarity",
           y = "Prey Family") +
      theme()

    print(p)

    cat("\n\n")
    }
  }

  return(simper_result)
}

# Run SIMPER for both primer sets (heatmap mode by default)
simper_ps1 <- run_simper_analysis(ps1_bale_filt, "PRIMERSET 1",
                                  detailed_output = FALSE)
simper_ps2 <- run_simper_analysis(ps2_bale_filt, "PRIMERSET 2",
                                  detailed_output = FALSE)
```

---

# LEVEL 2: BP vs. TP Mormyrids Globally

**Biological context:** All mormyrid species across all sites (97-99 individuals). This level examines the broadest dietary differences associated with EOD waveform type (biphasic vs. triphasic) across the entire mormyrid community.

## 2.1 Prey Composition by Waveform

Family-level prey composition comparing BP and TP waveform mormyrids globally.

```{r level2-prey-composition}
cat("=== LEVEL 2.1: PREY COMPOSITION BY WAVEFORM (FAMILY LEVEL) ===\n\n")

# Aggregate to family level
ps1_all_family <- aggregate_to_family(ps1_all_mormyrids)
ps2_all_family <- aggregate_to_family(ps2_all_mormyrids)

cat("Family-level aggregation:\n")
cat("  PS1:", ntaxa(ps1_all_family), "families\n")
cat("  PS2:", ntaxa(ps2_all_family), "families\n")

# Calculate summary statistics by waveform type
ps1_all_rel <- transform_sample_counts(ps1_all_family, function(x) x / sum(x) * 100)
ps2_all_rel <- transform_sample_counts(ps2_all_family, function(x) x / sum(x) * 100)

# Function to calculate family summary by waveform
calc_family_summary_waveform <- function(ps_rel) {
  ps_melt <- psmelt(ps_rel)
  ps_melt$Family_clean <- sapply(ps_melt$Family, clean_family_name)

  summary_table <- ps_melt %>%
    group_by(Family_clean, Waveform_Type) %>%
    summarise(
      mean_pct = mean(Abundance),
      se_pct = sd(Abundance) / sqrt(n()),
      n_samples = n(),
      .groups = "drop"
    ) %>%
    arrange(desc(mean_pct)) %>%
    mutate(mean_se = paste0(round(mean_pct, 2), " ± ", round(se_pct, 2)))

  return(summary_table)
}

summary_wave_ps1 <- calc_family_summary_waveform(ps1_all_rel)
summary_wave_ps2 <- calc_family_summary_waveform(ps2_all_rel)

cat("\nTop 15 prey families by waveform (Primer Set 1):\n")
print(head(summary_wave_ps1, 30))

cat("\nTop 15 prey families by waveform (Primer Set 2):\n")
print(head(summary_wave_ps2, 30))

# Identify top families across both primer sets for unified coloring
combined_summary_wave <- bind_rows(
  summary_wave_ps1 %>% mutate(Dataset = "PS1"),
  summary_wave_ps2 %>% mutate(Dataset = "PS2")
) %>%
  group_by(Family_clean) %>%
  summarise(Mean_Prop = mean(mean_pct), .groups = "drop") %>%
  arrange(desc(Mean_Prop)) %>%
  slice_head(n = 15)

top_families_wave_unified <- combined_summary_wave$Family_clean

cat("\nTop families (unified across both primer sets):\n")
print(top_families_wave_unified)

# Ensure all top families have colors assigned in global palette
missing_families_wave <- setdiff(top_families_wave_unified, names(GLOBAL_FAMILY_COLORS))
if (length(missing_families_wave) > 0) {
  cat("Adding colors for", length(missing_families_wave), "families not in global palette:\n")
  print(missing_families_wave)

  # Find colors not yet used
  used_colors <- GLOBAL_FAMILY_COLORS[names(GLOBAL_FAMILY_COLORS) != "Other"]
  available_colors <- setdiff(distinct_colors, used_colors)

  # Add colors for missing families
  new_colors <- setNames(
    available_colors[1:length(missing_families_wave)],
    missing_families_wave
  )
  GLOBAL_FAMILY_COLORS <- c(GLOBAL_FAMILY_COLORS, new_colors)
}

# Create stacked bar plots with unified family list
p1_wave <- plot_composition_unified(ps1_all_family,
                                    group_var = "Waveform_Type",
                                    families_to_show = top_families_wave_unified,
                                    title = "Prey Composition by Waveform (PS1)")

p2_wave <- plot_composition_unified(ps2_all_family,
                                    group_var = "Waveform_Type",
                                    families_to_show = top_families_wave_unified,
                                    title = "Prey Composition by Waveform (PS2)")

# Display side by side with unified legend using cowplot
p1_wave_no_legend <- p1_wave + theme(legend.position = "none")
p2_wave_no_legend <- p2_wave + theme(legend.position = "none")

# Create dummy plot for unified legend
dummy_data_wave <- data.frame(
  x = rep("A", length(top_families_wave_unified) + 1),
  y = rep(1, length(top_families_wave_unified) + 1),
  Family = factor(c(top_families_wave_unified, "Other"),
                  levels = c(top_families_wave_unified, "Other"))
)

p_dummy_wave <- ggplot(dummy_data_wave, aes(x = x, y = y, fill = Family)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = GLOBAL_FAMILY_COLORS,
                    breaks = c(top_families_wave_unified, "Other"),
                    name = "Prey Family",
                    drop = FALSE) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(ncol = 5, byrow = TRUE, title = "Prey Family"))

legend_wave <- get_legend(p_dummy_wave)

# Combine plots and legend
plot_grid(
  plot_grid(p1_wave_no_legend, p2_wave_no_legend, ncol = 2),
  legend_wave,
  ncol = 1,
  rel_heights = c(1, 0.15)
)
```

Summary table of top prey families by waveform:

```{r level2-prey-table}
# Create combined summary table
top_families_wave <- summary_wave_ps1 %>%
  pivot_wider(id_cols = Family_clean,
              names_from = Waveform_Type,
              values_from = c(mean_pct, se_pct),
              values_fill = list(mean_pct = 0, se_pct = 0)) %>%
  arrange(desc(pmax(mean_pct_BP, mean_pct_TP, na.rm = TRUE))) %>%
  slice_head(n = 15)

kable(top_families_wave,
      caption = "Top 15 prey families by waveform type (mean $\\pm$ SE relative abundance \\%)",
      digits = 2,
      col.names = c("Family", "Mean \\% BP", "Mean \\% TP", "SE BP", "SE TP"),
      booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 2.2 Alpha Diversity: BP vs. TP

Shannon diversity analysis comparing BP and TP waveform mormyrids globally.

```{r level2-alpha-diversity, fig.width=12, fig.height=6}
cat("=== LEVEL 2.2: ALPHA DIVERSITY - BP vs TP ===\n\n")

# Analyze alpha diversity for both primer sets
alpha1_wave <- analyze_alpha_diversity(ps1_all_mormyrids, "Waveform_Type")
alpha2_wave <- analyze_alpha_diversity(ps2_all_mormyrids, "Waveform_Type")

# Calculate summary statistics
cat("\n=== PRIMERSET 1 - BP vs TP Shannon Diversity ===\n")
bp_mean_ps1 <- mean(alpha1_wave$diversity$Shannon[alpha1_wave$diversity$Waveform_Type == "BP"], na.rm = TRUE)
bp_sd_ps1 <- sd(alpha1_wave$diversity$Shannon[alpha1_wave$diversity$Waveform_Type == "BP"], na.rm = TRUE)
tp_mean_ps1 <- mean(alpha1_wave$diversity$Shannon[alpha1_wave$diversity$Waveform_Type == "TP"], na.rm = TRUE)
tp_sd_ps1 <- sd(alpha1_wave$diversity$Shannon[alpha1_wave$diversity$Waveform_Type == "TP"], na.rm = TRUE)

cat("BP Shannon (mean ± SD):", round(bp_mean_ps1, 3), "±", round(bp_sd_ps1, 3), "\n")
cat("TP Shannon (mean ± SD):", round(tp_mean_ps1, 3), "±", round(tp_sd_ps1, 3), "\n")

cat("\nTest type:", alpha1_wave$test_type, "\n")
cat("Shapiro-Wilk p-value:", round(alpha1_wave$shapiro_p, 4), "\n")
if (alpha1_wave$test_type == "Kruskal-Wallis") {
  print(alpha1_wave$test)
} else {
  print(summary(alpha1_wave$test))
}

# Cohen's d effect size
cat("\nEffect size (Cohen's d):\n")
bp_vals_ps1 <- alpha1_wave$diversity$Shannon[alpha1_wave$diversity$Waveform_Type == "BP"]
tp_vals_ps1 <- alpha1_wave$diversity$Shannon[alpha1_wave$diversity$Waveform_Type == "TP"]
cohen_ps1 <- effsize::cohen.d(bp_vals_ps1, tp_vals_ps1, na.rm = TRUE)
print(cohen_ps1)

cat("\n=== PRIMERSET 2 - BP vs TP Shannon Diversity ===\n")
bp_mean_ps2 <- mean(alpha2_wave$diversity$Shannon[alpha2_wave$diversity$Waveform_Type == "BP"], na.rm = TRUE)
bp_sd_ps2 <- sd(alpha2_wave$diversity$Shannon[alpha2_wave$diversity$Waveform_Type == "BP"], na.rm = TRUE)
tp_mean_ps2 <- mean(alpha2_wave$diversity$Shannon[alpha2_wave$diversity$Waveform_Type == "TP"], na.rm = TRUE)
tp_sd_ps2 <- sd(alpha2_wave$diversity$Shannon[alpha2_wave$diversity$Waveform_Type == "TP"], na.rm = TRUE)

cat("BP Shannon (mean ± SD):", round(bp_mean_ps2, 3), "±", round(bp_sd_ps2, 3), "\n")
cat("TP Shannon (mean ± SD):", round(tp_mean_ps2, 3), "±", round(tp_sd_ps2, 3), "\n")

cat("\nTest type:", alpha2_wave$test_type, "\n")
cat("Shapiro-Wilk p-value:", round(alpha2_wave$shapiro_p, 4), "\n")
if (alpha2_wave$test_type == "Kruskal-Wallis") {
  print(alpha2_wave$test)
} else {
  print(summary(alpha2_wave$test))
}

# Cohen's d effect size
cat("\nEffect size (Cohen's d):\n")
bp_vals_ps2 <- alpha2_wave$diversity$Shannon[alpha2_wave$diversity$Waveform_Type == "BP"]
tp_vals_ps2 <- alpha2_wave$diversity$Shannon[alpha2_wave$diversity$Waveform_Type == "TP"]
cohen_ps2 <- effsize::cohen.d(bp_vals_ps2, tp_vals_ps2, na.rm = TRUE)
print(cohen_ps2)

# Comparative plot
p_wave_alpha_comp <- plot_comparative_shannon(alpha1_wave$diversity,
                                             alpha2_wave$diversity,
                                             "Waveform_Type",
                                             "Global BP vs TP Diet Diversity Comparison")
print(p_wave_alpha_comp)
```

### Cross-Primer Correlation for Alpha Diversity

```{r level2-alpha-correlation, fig.width=8, fig.height=6}
# Correlate Shannon values for shared samples
cor_wave_alpha <- correlate_shared_samples(alpha1_wave$diversity,
                                          alpha2_wave$diversity)

cat("\n=== GLOBAL WAVEFORM ALPHA DIVERSITY CORRELATION ===\n")
cat("Shared samples:", cor_wave_alpha$n_shared, "\n")

if (!is.null(cor_wave_alpha$cor_test)) {
  cat("Pearson r:", round(cor_wave_alpha$cor_test$estimate, 3), "\n")
  cat("95% CI: [", round(cor_wave_alpha$cor_test$conf.int[1], 3), ",",
      round(cor_wave_alpha$cor_test$conf.int[2], 3), "]\n")
  cat("P-value:", format.pval(cor_wave_alpha$cor_test$p.value, digits = 3), "\n")

  # Scatterplot - get shared samples
  shared_ids <- intersect(alpha1_wave$diversity$Sample_Name,
                          alpha2_wave$diversity$Sample_Name)

  # Filter to shared samples and match order
  div1_shared <- alpha1_wave$diversity %>%
    filter(Sample_Name %in% shared_ids) %>%
    arrange(Sample_Name)
  div2_shared <- alpha2_wave$diversity %>%
    filter(Sample_Name %in% shared_ids) %>%
    arrange(Sample_Name)

  cor_df <- data.frame(
    PS1_Shannon = div1_shared$Shannon,
    PS2_Shannon = div2_shared$Shannon
  )

  p_alpha_cor <- ggplot(cor_df, aes(x = PS1_Shannon, y = PS2_Shannon)) +
    geom_point(size = 3, alpha = 0.6, color = ps1_color) +
    geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
    geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "gray50") +
    theme_minimal() +
    labs(title = "Global Waveform: Shannon Diversity Correlation",
         subtitle = paste0("r = ", round(cor_wave_alpha$cor_test$estimate, 3),
                          ", p ", format.pval(cor_wave_alpha$cor_test$p.value, digits = 2)),
         x = "Primerset 1 Shannon", y = "Primerset 2 Shannon") +
    coord_fixed()

  print(p_alpha_cor)
} else {
  cat(cor_wave_alpha$message, "\n")
}
```

## 2.3 Beta Diversity: BP vs. TP

Bray-Curtis dissimilarity analysis with PERMANOVA testing for dietary differences between BP and TP waveform types globally.

```{r level2-beta-diversity, fig.width=14, fig.height=6}
cat("\n=== LEVEL 2.3: BETA DIVERSITY - BP vs TP ===\n\n")

# Run beta diversity analyses
beta1_wave <- run_beta_diversity_analysis(ps1_all_mormyrids, "~ Waveform_Type")
beta2_wave <- run_beta_diversity_analysis(ps2_all_mormyrids, "~ Waveform_Type")

cat("\n=== PRIMERSET 1 - Global Waveform PERMANOVA ===\n")
print(beta1_wave$permanova)

# Extract R² for reporting
r2_ps1 <- beta1_wave$permanova$R2[1]
p_ps1 <- beta1_wave$permanova$`Pr(>F)`[1]
cat("\nR² =", round(r2_ps1, 4), ", p =", format.pval(p_ps1, digits = 3), "\n")

cat("\n=== PRIMERSET 2 - Global Waveform PERMANOVA ===\n")
print(beta2_wave$permanova)

# Extract R² for reporting
r2_ps2 <- beta2_wave$permanova$R2[1]
p_ps2 <- beta2_wave$permanova$`Pr(>F)`[1]
cat("\nR² =", round(r2_ps2, 4), ", p =", format.pval(p_ps2, digits = 3), "\n")

# Faceted PCoA plot
p_wave_pcoa <- plot_faceted_pcoa(beta1_wave, beta2_wave,
                                color_var = "Waveform_Type",
                                title = "Global Diet Composition: BP vs TP Waveform")
print(p_wave_pcoa)
```

### Beta Diversity Cross-Validation (Mantel Test)

```{r level2-beta-mantel}
# Find shared Sample_Names between datasets
samples1_wave <- beta1_wave$metadata$Sample_Name
samples2_wave <- beta2_wave$metadata$Sample_Name
shared_wave <- intersect(samples1_wave, samples2_wave)

cat("\n=== GLOBAL WAVEFORM BETA DIVERSITY MANTEL TEST ===\n")
cat("Shared Sample_Names:", length(shared_wave), "\n")

if (length(shared_wave) >= 3) {
  # Convert distance matrices to matrices with Sample_Name labels
  dist1_mat <- as.matrix(beta1_wave$distance)
  dist2_mat <- as.matrix(beta2_wave$distance)

  # Relabel with Sample_Names
  rownames(dist1_mat) <- beta1_wave$metadata$Sample_Name
  colnames(dist1_mat) <- beta1_wave$metadata$Sample_Name
  rownames(dist2_mat) <- beta2_wave$metadata$Sample_Name
  colnames(dist2_mat) <- beta2_wave$metadata$Sample_Name

  # Subset to shared Sample_Names (automatically aligned)
  dist1_shared <- as.dist(dist1_mat[shared_wave, shared_wave])
  dist2_shared <- as.dist(dist2_mat[shared_wave, shared_wave])

  # Mantel test
  mantel_wave <- run_mantel_test(dist1_shared, dist2_shared)

  cat("\nMantel statistic (r):", round(mantel_wave$statistic, 3), "\n")
  cat("P-value:", format.pval(mantel_wave$signif, digits = 3), "\n")
  cat("Interpretation:",
      ifelse(mantel_wave$statistic > 0.5, "Strong concordance",
             ifelse(mantel_wave$statistic > 0.3, "Moderate concordance",
                    "Weak concordance")), "\n")
} else {
  cat("Insufficient shared samples for Mantel test.\n")
  mantel_wave <- NULL
}
```

## 2.4 SIMPER Analysis - Waveform

Similarity Percentage (SIMPER) analysis identifies which prey families contribute most to dietary differences between BP and TP waveform types.

```{r level2-simper, fig.width=10, fig.height=8}
cat("\n=== LEVEL 2.4: SIMPER ANALYSIS - WAVEFORM ===\n\n")

# Function to run SIMPER and create visualization
run_simper_waveform <- function(ps, primerset_name) {

  cat("\n", primerset_name, ":\n", sep = "")

  # Get OTU table (samples as rows, taxa as columns)
  otu_mat <- as.matrix(otu_table(ps))
  if (taxa_are_rows(ps)) {
    otu_mat <- t(otu_mat)
  }

  # Get metadata
  meta <- data.frame(sample_data(ps))

  # Run SIMPER
  simper_result <- vegan::simper(otu_mat, meta$Waveform_Type, permutations = 999)

  cat("\nSIMPER analysis complete\n")

  # Extract results for BP vs TP comparison
  # There should be only one comparison (BP_TP or TP_BP)
  comparison_name <- names(simper_result)[1]
  comparison_name_clean <- gsub("_", " vs ", comparison_name)
  comp <- simper_result[[comparison_name]]

  cat("\nComparison:", comparison_name_clean, "\n")

  # Extract overall dissimilarity
  overall_diss <- comp$overall
  cat("Overall average dissimilarity:", round(overall_diss, 3), "\n")

  # Build data frame from simper result
  comparison_data <- data.frame(
    species = comp$species,
    average = comp$average,
    sd = comp$sd,
    ratio = comp$ratio,
    ava = comp$ava,
    avb = comp$avb,
    p = comp$p,
    stringsAsFactors = FALSE
  )

  # Calculate contribution percentage (average contribution / overall dissimilarity)
  comparison_data$contr <- comparison_data$average / overall_diss

  # Sort by contribution (descending)
  comparison_data <- comparison_data[order(-comparison_data$contr), ]
  rownames(comparison_data) <- comparison_data$species

  # Calculate cumulative sum
  comparison_data$cumsum <- cumsum(comparison_data$contr)

  # Get top contributors (cumulative contribution up to 70%)
  top_taxa <- comparison_data[comparison_data$cumsum <= 0.70, ]

  # If nothing meets 70%, take top 15
  if (nrow(top_taxa) == 0) {
    top_taxa <- head(comparison_data, 15)
  }

  # Get taxonomy for these ASVs
  tax_df <- as.data.frame(tax_table(ps))
  top_taxa$ASV <- rownames(top_taxa)

  # Add taxonomic names (Family level)
  top_taxa$Family <- sapply(top_taxa$ASV, function(asv) {
    if (asv %in% rownames(tax_df)) {
      fam <- tax_df[asv, "Family"]
      if (is.na(fam) || fam == "") {
        return(tax_df[asv, "Order"])
      }
      return(clean_family_name(fam))
    }
    return("Unknown")
  })

  # Aggregate ASVs by Family for plotting
  family_aggregated <- top_taxa %>%
    group_by(Family) %>%
    summarise(
      contr = sum(contr),
      average = sum(average),
      n_asvs = n(),
      .groups = "drop"
    ) %>%
    arrange(desc(contr))

  # Print table (ASV-level detail)
  display_table <- top_taxa[, c("Family", "average", "sd", "ratio",
                                "ava", "avb", "contr", "cumsum")]

  # Round numeric columns
  display_table[, c("average", "sd", "ratio", "ava", "avb",
                    "contr", "cumsum")] <-
    round(display_table[, c("average", "sd", "ratio", "ava", "avb",
                            "contr", "cumsum")], 3)

  cat(paste0("\n#### SIMPER: ", comparison_name_clean, " (", primerset_name, ")\n\n"))
  # Limit to top 10 rows for space
  display_table_top <- head(display_table, 10)
  kable(display_table_top,
              caption = paste0("SIMPER: ", comparison_name_clean, " (", primerset_name, ") - Top 10"),
              row.names = FALSE,
              booktabs = TRUE) %>%
          kable_styling(latex_options = c("striped", "HOLD_position"),
                        bootstrap_options = c("striped", "hover"),
                        full_width = FALSE, font_size = 8)

  # Create bar plot of contribution percentage - FAMILY LEVEL
  plot_data <- head(family_aggregated, 15)

  # Create factor for ordering (highest contribution at top)
  plot_data$Family <- factor(plot_data$Family,
                             levels = rev(plot_data$Family))

  p <- ggplot(plot_data, aes(x = contr * 100, y = Family)) +
    geom_bar(stat = "identity", fill = "#0072B2", alpha = 0.7) +
    theme_minimal() +
    labs(title = paste0("Top Prey Families Contributing to BP vs TP Dissimilarity (", primerset_name, ")"),
         x = "Contribution to Dissimilarity (%)",
         y = "Prey Family")

  return(list(simper = simper_result, plot = p))
}

# Run SIMPER for both primer sets
simper_wave_ps1 <- run_simper_waveform(ps1_all_mormyrids, "PRIMERSET 1")
simper_wave_ps2 <- run_simper_waveform(ps2_all_mormyrids, "PRIMERSET 2")

# Display plots
print(simper_wave_ps1$plot)
print(simper_wave_ps2$plot)
```

---

# LEVEL 3: Paramormyrops kingsleyae - General Population Factors

**Biological context:** ~47 *P. kingsleyae* individuals from 4 rivers (2 regions × 2 waveform types). This level examines how sex, body size, parasitism, and geography affect diet before testing the core BP vs TP question in Level 4.

## 3.1 Prey Composition by Population

Family-level prey composition across the four river populations, organized by region and waveform type to facilitate comparisons.

```{r level3-prey-composition, fig.width=14, fig.height=10}
cat("=== LEVEL 3.1: PREY COMPOSITION BY POPULATION ===\n\n")

# Aggregate to family level
ps1_pk_family <- aggregate_to_family(ps1_pk)
ps2_pk_family <- aggregate_to_family(ps2_pk)

cat("Family-level aggregation (P. kingsleyae):\n")
cat("  PS1:", ntaxa(ps1_pk_family), "families\n")
cat("  PS2:", ntaxa(ps2_pk_family), "families\n")

# Add Region variable for faceting
add_region_variable <- function(ps) {
  meta <- data.frame(sample_data(ps))

  # Define regions based on locality
  meta$Region <- ifelse(meta$Locality_Name %in% c("Bambomo", "Apassa"),
                        "South", "North")

  # Update sample data
  sample_data(ps) <- meta
  return(ps)
}

ps1_pk_family <- add_region_variable(ps1_pk_family)
ps2_pk_family <- add_region_variable(ps2_pk_family)

# Create 4-panel stacked bar chart function
create_4panel_composition <- function(ps_family, primerset_name, top_n = 10,
                                      families_to_show = NULL,
                                      use_global_colors = TRUE) {

  # Convert to relative abundance
  ps_rel <- transform_sample_counts(ps_family, function(x) x / sum(x))

  # Melt to long format
  ps_melt <- psmelt(ps_rel)

  # Clean family names
  ps_melt$Family_clean <- sapply(ps_melt$Family, clean_family_name)

  # Identify top N families by mean abundance
  if (is.null(families_to_show)) {
    top_families <- ps_melt %>%
      group_by(Family_clean) %>%
      summarise(mean_abund = mean(Abundance), .groups = "drop") %>%
      arrange(desc(mean_abund)) %>%
      slice_head(n = top_n) %>%
      pull(Family_clean)
  } else {
    top_families <- families_to_show
  }

  # Aggregate by Locality_Name and Family_clean (sum across samples)
  ps_agg <- ps_melt %>%
    group_by(Locality_Name, Region, Waveform_Type, Family_clean) %>%
    summarise(Abundance = sum(Abundance), .groups = "drop")

  # Normalize within each Locality_Name so each bar sums to 1 (100%)
  ps_agg <- ps_agg %>%
    group_by(Locality_Name) %>%
    mutate(RelAbund = Abundance / sum(Abundance)) %>%
    ungroup()

  # Categorize families into top families vs "Other"
  ps_final <- ps_agg %>%
    mutate(Family_display = ifelse(Family_clean %in% top_families,
                                   Family_clean, "Other")) %>%
    group_by(Locality_Name, Region, Waveform_Type, Family_display) %>%
    summarise(RelAbund = sum(RelAbund), .groups = "drop")

  # Reorder for plotting
  ps_final$Family_display <- factor(ps_final$Family_display,
                                     levels = c(top_families, "Other"))

  # Create plot with facet_grid(Region ~ Waveform_Type)
  p <- ggplot(ps_final, aes(x = Locality_Name, y = RelAbund,
                            fill = Family_display)) +
    geom_bar(stat = "identity", position = "stack") +
    facet_grid(Region ~ Waveform_Type, scales = "free_x", space = "free_x") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right"
    ) +
    labs(
      title = paste("P. kingsleyae Prey Composition by Population -", primerset_name),
      subtitle = "Organized by Region (North/South) and Waveform (BP/TP)",
      x = "River",
      y = "Relative Abundance"
    ) +
    scale_y_continuous(labels = scales::percent)

  # Add color scale
  if (use_global_colors) {
    p <- p + scale_fill_manual(values = GLOBAL_FAMILY_COLORS,
                               name = "Prey Family",
                               breaks = c(top_families, "Other"),
                               drop = FALSE)
  } else {
    p <- p + scale_fill_viridis_d(option = "turbo", name = "Prey Family")
  }

  return(p)
}

# Identify top families across both primer sets for unified coloring
# Calculate summaries for both datasets
ps1_pk_rel <- transform_sample_counts(ps1_pk_family, function(x) x / sum(x) * 100)
ps2_pk_rel <- transform_sample_counts(ps2_pk_family, function(x) x / sum(x) * 100)

# Get family summaries
get_family_summary_pk <- function(ps_rel) {
  ps_melt <- psmelt(ps_rel)
  ps_melt$Family_clean <- sapply(ps_melt$Family, clean_family_name)

  summary_table <- ps_melt %>%
    group_by(Family_clean) %>%
    summarise(
      mean_pct = mean(Abundance),
      sd_pct = sd(Abundance),
      n_samples = n(),
      .groups = "drop"
    ) %>%
    arrange(desc(mean_pct)) %>%
    mutate(mean_sd = paste0(round(mean_pct, 2), " ± ", round(sd_pct, 2)))

  return(summary_table)
}

summary_pk_ps1 <- get_family_summary_pk(ps1_pk_rel)
summary_pk_ps2 <- get_family_summary_pk(ps2_pk_rel)

# Combine and get top families
combined_summary_pk <- bind_rows(
  summary_pk_ps1 %>% mutate(Dataset = "PS1"),
  summary_pk_ps2 %>% mutate(Dataset = "PS2")
) %>%
  group_by(Family_clean) %>%
  summarise(Mean_Prop = mean(mean_pct), .groups = "drop") %>%
  arrange(desc(Mean_Prop)) %>%
  slice_head(n = 10)

top_families_pk_unified <- combined_summary_pk$Family_clean

cat("\nTop families for P. kingsleyae (unified across both primer sets):\n")
print(top_families_pk_unified)

# Ensure all top families have colors assigned in global palette
missing_families_pk <- setdiff(top_families_pk_unified, names(GLOBAL_FAMILY_COLORS))
if (length(missing_families_pk) > 0) {
  cat("Adding colors for", length(missing_families_pk), "families not in global palette:\n")
  print(missing_families_pk)

  # Find colors not yet used
  used_colors <- GLOBAL_FAMILY_COLORS[names(GLOBAL_FAMILY_COLORS) != "Other"]
  available_colors <- setdiff(distinct_colors, used_colors)

  # Add colors for missing families
  new_colors <- setNames(
    available_colors[1:length(missing_families_pk)],
    missing_families_pk
  )
  GLOBAL_FAMILY_COLORS <- c(GLOBAL_FAMILY_COLORS, new_colors)
}

# Create plots for both primer sets with unified families
p_comp_ps1 <- create_4panel_composition(ps1_pk_family, "Primerset 1",
                                        families_to_show = top_families_pk_unified,
                                        use_global_colors = TRUE)
p_comp_ps2 <- create_4panel_composition(ps2_pk_family, "Primerset 2",
                                        families_to_show = top_families_pk_unified,
                                        use_global_colors = TRUE)

print(p_comp_ps1)
cat("\n\n")
print(p_comp_ps2)

# Summary table by river
cat("\n=== SUMMARY BY RIVER ===\n")
pk_meta <- data.frame(sample_data(ps1_pk))
pk_summary <- pk_meta %>%
  group_by(Locality_Name, Region, Waveform_Type) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(Region, Waveform_Type)

kable(pk_summary,
      caption = "P. kingsleyae sampling by river",
      col.names = c("River", "Region", "Waveform", "N"),
      booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

## 3.2 Sex Effects on Shannon Diversity

Test for sex differences in diet diversity using ANCOVA with body size as a covariate to account for sexual size dimorphism.

```{r level3-sex-effects, fig.width=12, fig.height=10}
cat("\n=== LEVEL 3.2: SEX EFFECTS ON SHANNON DIVERSITY ===\n\n")

# Prepare data with sex and size
prepare_sex_size_data <- function(ps, size_col = "SL") {
  meta <- as.data.frame(sample_data(ps))
  div <- estimate_richness(ps, measures = c("Shannon"))

  div$Sex <- meta$Sex
  div$body_size <- as.numeric(meta[[size_col]])
  div$Sample_Name <- meta$Sample_Name
  div$Locality_Name <- meta$Locality_Name
  div$Waveform_Type <- meta$Waveform_Type

  # Remove missing data
  div <- div[!is.na(div$Sex) & !is.na(div$body_size), ]

  return(div)
}

div_ps1_pk <- prepare_sex_size_data(ps1_pk, "SL")
div_ps2_pk <- prepare_sex_size_data(ps2_pk, "SL")

cat("Sample sizes (P. kingsleyae):\n")
cat("  PS1:", nrow(div_ps1_pk), "samples\n")
if (nrow(div_ps1_pk) > 0) {
  cat("  PS1 sex distribution:\n")
  print(table(div_ps1_pk$Sex))
}

cat("\n  PS2:", nrow(div_ps2_pk), "samples\n")
if (nrow(div_ps2_pk) > 0) {
  cat("  PS2 sex distribution:\n")
  print(table(div_ps2_pk$Sex))
}

# Filter to adults only for sex comparison (exclude juveniles)
div_ps1_pk_adults <- div_ps1_pk[div_ps1_pk$Sex %in% c("Male", "Female"), ]
div_ps2_pk_adults <- div_ps2_pk[div_ps2_pk$Sex %in% c("Male", "Female"), ]

cat("\nAdult samples (excluding juveniles):\n")
cat("  PS1:", nrow(div_ps1_pk_adults), "adults\n")
cat("  PS2:", nrow(div_ps2_pk_adults), "adults\n")

# Check correlation between sex and size in adults
if (nrow(div_ps1_pk_adults) > 5 && length(unique(div_ps1_pk_adults$Sex)) == 2) {
  cat("\n=== SEX-SIZE CORRELATION (P. kingsleyae ADULTS) ===\n")

  # PS1
  cat("\nPRIMERSET 1:\n")
  cat("Body size by sex (mean ± SD):\n")
  size_by_sex_ps1 <- aggregate(body_size ~ Sex, data = div_ps1_pk_adults,
                                FUN = function(x) {
                                  paste0(round(mean(x), 1), " ± ",
                                         round(sd(x), 1))
                                })
  print(size_by_sex_ps1)

  size_test_ps1 <- t.test(body_size ~ Sex, data = div_ps1_pk_adults)
  cat("t-test p-value:", format.pval(size_test_ps1$p.value, digits = 3), "\n")

  # PS2
  if (nrow(div_ps2_pk_adults) > 5 && length(unique(div_ps2_pk_adults$Sex)) == 2) {
    cat("\nPRIMERSET 2:\n")
    cat("Body size by sex (mean ± SD):\n")
    size_by_sex_ps2 <- aggregate(body_size ~ Sex, data = div_ps2_pk_adults,
                                  FUN = function(x) {
                                    paste0(round(mean(x), 1), " ± ",
                                           round(sd(x), 1))
                                  })
    print(size_by_sex_ps2)

    size_test_ps2 <- t.test(body_size ~ Sex, data = div_ps2_pk_adults)
    cat("t-test p-value:", format.pval(size_test_ps2$p.value, digits = 3), "\n")
  }

  # ANCOVA models (sex + body size)
  cat("\n=== ANCOVA MODELS (ADULTS ONLY) ===\n")
  cat("Model: Shannon ~ Sex + body_size + Sex:body_size\n\n")

  # PS1 ANCOVA
  cat("PRIMERSET 1:\n")
  ancova_sex_ps1 <- lm(Shannon ~ Sex + body_size + Sex:body_size,
                       data = div_ps1_pk_adults)
  print(summary(ancova_sex_ps1))
  cat("\nANOVA table:\n")
  print(anova(ancova_sex_ps1))

  # PS2 ANCOVA
  if (nrow(div_ps2_pk_adults) > 5 && length(unique(div_ps2_pk_adults$Sex)) == 2) {
    cat("\n\nPRIMERSET 2:\n")
    ancova_sex_ps2 <- lm(Shannon ~ Sex + body_size + Sex:body_size,
                         data = div_ps2_pk_adults)
    print(summary(ancova_sex_ps2))
    cat("\nANOVA table:\n")
    print(anova(ancova_sex_ps2))
  }

  # Visualization
  div_ps1_pk_adults$PrimerSet <- "Primerset 1"
  div_ps2_pk_adults$PrimerSet <- "Primerset 2"
  div_combined_adults <- bind_rows(div_ps1_pk_adults, div_ps2_pk_adults)

  # Also prepare all-individuals dataset
  div_ps1_pk$PrimerSet <- "Primerset 1"
  div_ps2_pk$PrimerSet <- "Primerset 2"
  div_combined_all <- bind_rows(div_ps1_pk, div_ps2_pk)

  # Panel 1: Shannon by sex (adults only)
  p_sex <- ggplot(div_combined_adults,
                  aes(x = Sex, y = Shannon, fill = Sex)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.5) +
    facet_wrap(~ PrimerSet) +
    theme_minimal() +
    labs(title = "Diet Diversity by Sex (P. kingsleyae Adults)",
         x = "Sex", y = "Shannon Diversity") +
    theme(legend.position = "none")

  # Panel 2: Body size by sex (adults only)
  p_size_sex <- ggplot(div_combined_adults,
                       aes(x = Sex, y = body_size, fill = Sex)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.5) +
    facet_wrap(~ PrimerSet) +
    theme_minimal() +
    labs(title = "Body Size by Sex (P. kingsleyae Adults)",
         x = "Sex",
         y = "Standard Length (mm)") +
    theme(legend.position = "none")

  # Panel 3: Shannon vs body size with sex indicated
  p_scatter <- ggplot(div_combined_all,
                      aes(x = body_size, y = Shannon, color = Sex)) +
    geom_point(alpha = 0.6, size = 2) +
    geom_smooth(method = "lm", se = TRUE) +
    facet_wrap(~ PrimerSet) +
    theme_minimal() +
    labs(title = "Diet Diversity vs Body Size (All P. kingsleyae)",
         x = "Standard Length (mm)",
         y = "Shannon Diversity",
         color = "Sex")

  # Combine plots
  print(p_sex / p_size_sex / p_scatter)

} else {
  cat("Insufficient data for sex analysis\n")
}
```

## 3.3 Body Size Effects

Test for linear relationship between body size and diet diversity across all *P. kingsleyae* individuals.

```{r level3-body-size, fig.width=10, fig.height=6}
cat("\n=== LEVEL 3.3: BODY SIZE EFFECTS ===\n")
cat("Testing Shannon ~ body_size for all P. kingsleyae individuals\n\n")

# PS1
cat("PRIMERSET 1:\n")
lm_size_ps1 <- lm(Shannon ~ body_size, data = div_ps1_pk)
print(summary(lm_size_ps1))
cat("\nPearson correlation:\n")
cor_size_ps1 <- cor.test(div_ps1_pk$Shannon, div_ps1_pk$body_size)
cat("  r:", round(cor_size_ps1$estimate, 3), "\n")
cat("  95% CI: [", round(cor_size_ps1$conf.int[1], 3), ",",
    round(cor_size_ps1$conf.int[2], 3), "]\n")
cat("  P-value:", format.pval(cor_size_ps1$p.value, digits = 3), "\n")

# PS2
cat("\n\nPRIMERSET 2:\n")
lm_size_ps2 <- lm(Shannon ~ body_size, data = div_ps2_pk)
print(summary(lm_size_ps2))
cat("\nPearson correlation:\n")
cor_size_ps2 <- cor.test(div_ps2_pk$Shannon, div_ps2_pk$body_size)
cat("  r:", round(cor_size_ps2$estimate, 3), "\n")
cat("  95% CI: [", round(cor_size_ps2$conf.int[1], 3), ",",
    round(cor_size_ps2$conf.int[2], 3), "]\n")
cat("  P-value:", format.pval(cor_size_ps2$p.value, digits = 3), "\n")

# Scatterplot with regression line
if (exists("div_combined_all")) {
  p_size_effect <- ggplot(div_combined_all,
                          aes(x = body_size, y = Shannon)) +
    geom_point(aes(color = Sex), alpha = 0.6, size = 3) +
    geom_smooth(method = "lm", se = TRUE, color = "black",
                linetype = "dashed", linewidth = 1) +
    facet_wrap(~ PrimerSet) +
    theme_minimal() +
    labs(title = "Body Size Effect on Diet Diversity (P. kingsleyae)",
         subtitle = paste0("PS1: r = ", round(cor_size_ps1$estimate, 3),
                          ", p ", format.pval(cor_size_ps1$p.value, digits = 2),
                          " | PS2: r = ", round(cor_size_ps2$estimate, 3),
                          ", p ", format.pval(cor_size_ps2$p.value, digits = 2)),
         x = "Standard Length (mm)",
         y = "Shannon Diversity",
         color = "Sex")

  print(p_size_effect)
}
```

## 3.4 Parasitism Effects

Test whether parasitism affects diet diversity and composition in *P. kingsleyae*.

```{r level3-parasitism, fig.width=12, fig.height=8}
cat("\n=== LEVEL 3.4: PARASITISM EFFECTS ===\n\n")

# Check if parasitism data exists
check_parasitism <- function(ps) {
  meta <- data.frame(sample_data(ps))

  # Look for parasite-related columns
  parasite_cols <- grep("parasite|parasit", names(meta),
                        ignore.case = TRUE, value = TRUE)

  if (length(parasite_cols) == 0) {
    cat("No parasitism columns found in metadata\n")
    return(NULL)
  }

  cat("Found parasitism columns:", paste(parasite_cols, collapse = ", "), "\n")
  return(parasite_cols)
}

# Check both primer sets
cat("PRIMERSET 1:\n")
parasite_cols_ps1 <- check_parasitism(ps1_pk)

cat("\nPRIMERSET 2:\n")
parasite_cols_ps2 <- check_parasitism(ps2_pk)

# Analyze parasitism effects if data exists
if (!is.null(parasite_cols_ps1) && length(parasite_cols_ps1) > 0) {

  # Use the first parasite column found
  parasite_col <- parasite_cols_ps1[1]
  cat("\n=== ANALYZING PARASITISM USING COLUMN:", parasite_col, "===\n\n")

  # Prepare data with parasitism
  prepare_parasite_data <- function(ps, parasite_col) {
    meta <- as.data.frame(sample_data(ps))
    div <- estimate_richness(ps, measures = c("Shannon"))

    div$Parasite <- meta[[parasite_col]]
    div$body_size <- as.numeric(meta$SL)
    div$Sample_Name <- meta$Sample_Name
    div$Sex <- meta$Sex

    # Remove missing data
    div <- div[!is.na(div$Parasite) & !is.na(div$body_size), ]

    # Standardize parasite values (yes/no, presence/absence, etc.)
    div$Parasite <- tolower(as.character(div$Parasite))
    div$Parasite <- ifelse(div$Parasite %in% c("yes", "present", "1", "true"),
                           "Yes",
                           ifelse(div$Parasite %in% c("no", "absent", "0", "false"),
                                 "No", div$Parasite))

    return(div)
  }

  div_para_ps1 <- prepare_parasite_data(ps1_pk, parasite_col)
  div_para_ps2 <- prepare_parasite_data(ps2_pk, parasite_col)

  cat("PRIMERSET 1:\n")
  cat("  Samples with parasite data:", nrow(div_para_ps1), "\n")
  cat("  Parasitism status:\n")
  print(table(div_para_ps1$Parasite))

  cat("\nPRIMERSET 2:\n")
  cat("  Samples with parasite data:", nrow(div_para_ps2), "\n")
  cat("  Parasitism status:\n")
  print(table(div_para_ps2$Parasite))

  # Check if parasitism is categorical or continuous
  is_categorical <- length(unique(div_para_ps1$Parasite)) < 10

  if (is_categorical) {
    cat("\n=== CATEGORICAL PARASITISM ANALYSIS ===\n")

    # Kruskal-Wallis test for alpha diversity
    cat("\nAlpha Diversity (Shannon) by Parasitism Status:\n\n")

    # PS1
    cat("PRIMERSET 1:\n")
    kw_ps1 <- kruskal.test(Shannon ~ Parasite, data = div_para_ps1)
    print(kw_ps1)

    # Summary statistics
    para_summary_ps1 <- div_para_ps1 %>%
      group_by(Parasite) %>%
      summarise(
        n = n(),
        mean_shannon = mean(Shannon, na.rm = TRUE),
        sd_shannon = sd(Shannon, na.rm = TRUE),
        .groups = "drop"
      )
    print(para_summary_ps1)

    # PS2
    cat("\nPRIMERSET 2:\n")
    kw_ps2 <- kruskal.test(Shannon ~ Parasite, data = div_para_ps2)
    print(kw_ps2)

    para_summary_ps2 <- div_para_ps2 %>%
      group_by(Parasite) %>%
      summarise(
        n = n(),
        mean_shannon = mean(Shannon, na.rm = TRUE),
        sd_shannon = sd(Shannon, na.rm = TRUE),
        .groups = "drop"
      )
    print(para_summary_ps2)

    # ANCOVA controlling for body size
    cat("\n=== ANCOVA: Shannon ~ Parasite + body_size ===\n\n")

    # PS1
    cat("PRIMERSET 1:\n")
    ancova_para_ps1 <- lm(Shannon ~ Parasite + body_size, data = div_para_ps1)
    print(summary(ancova_para_ps1))
    cat("\nANOVA table:\n")
    print(anova(ancova_para_ps1))

    # PS2
    cat("\n\nPRIMERSET 2:\n")
    ancova_para_ps2 <- lm(Shannon ~ Parasite + body_size, data = div_para_ps2)
    print(summary(ancova_para_ps2))
    cat("\nANOVA table:\n")
    print(anova(ancova_para_ps2))

  } else {
    cat("\n=== CONTINUOUS PARASITISM ANALYSIS ===\n")
    cat("Parasitism appears to be continuous. Using correlation/regression.\n\n")

    # Convert to numeric if possible
    div_para_ps1$Parasite_numeric <- as.numeric(div_para_ps1$Parasite)
    div_para_ps2$Parasite_numeric <- as.numeric(div_para_ps2$Parasite)

    # ANCOVA with continuous parasite measure
    cat("PRIMERSET 1:\n")
    ancova_para_ps1 <- lm(Shannon ~ Parasite_numeric + body_size,
                         data = div_para_ps1)
    print(summary(ancova_para_ps1))

    cat("\n\nPRIMERSET 2:\n")
    ancova_para_ps2 <- lm(Shannon ~ Parasite_numeric + body_size,
                         data = div_para_ps2)
    print(summary(ancova_para_ps2))
  }

  # Beta diversity analysis
  cat("\n\n=== BETA DIVERSITY: PERMANOVA ===\n")
  cat("Testing: Bray-Curtis ~ Parasitism\n\n")

  # Create phyloseq subsets with parasitism metadata
  prepare_parasite_phyloseq <- function(ps, parasite_col) {
    meta <- data.frame(sample_data(ps))

    # Standardize parasite column
    meta$Parasite_status <- tolower(as.character(meta[[parasite_col]]))
    meta$Parasite_status <- ifelse(
      meta$Parasite_status %in% c("yes", "present", "1", "true"),
      "Yes",
      ifelse(meta$Parasite_status %in% c("no", "absent", "0", "false"),
             "No", meta$Parasite_status)
    )

    # Update sample data
    sample_data(ps) <- meta

    # Remove samples with missing parasite data
    ps_filt <- subset_samples(ps, !is.na(Parasite_status) &
                                   Parasite_status != "")

    return(ps_filt)
  }

  ps1_pk_para <- prepare_parasite_phyloseq(ps1_pk, parasite_col)
  ps2_pk_para <- prepare_parasite_phyloseq(ps2_pk, parasite_col)

  # Run beta diversity analysis
  if (nsamples(ps1_pk_para) > 5) {
    cat("PRIMERSET 1:\n")
    beta_para_ps1 <- run_beta_diversity_analysis(ps1_pk_para, "~ Parasite_status")
    print(beta_para_ps1$permanova)

    cat("\n\nPRIMERSET 2:\n")
    beta_para_ps2 <- run_beta_diversity_analysis(ps2_pk_para, "~ Parasite_status")
    print(beta_para_ps2$permanova)
  } else {
    cat("Insufficient samples with parasite data for beta diversity analysis\n")
  }

  # Visualization
  div_para_ps1$PrimerSet <- "Primerset 1"
  div_para_ps2$PrimerSet <- "Primerset 2"
  div_para_combined <- bind_rows(div_para_ps1, div_para_ps2)

  # Boxplot of Shannon by parasitism status
  p_para_box <- ggplot(div_para_combined,
                       aes(x = Parasite, y = Shannon, fill = Parasite)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.5, size = 2) +
    facet_wrap(~ PrimerSet) +
    theme_minimal() +
    labs(title = "Diet Diversity by Parasitism Status (P. kingsleyae)",
         x = "Parasitism Status",
         y = "Shannon Diversity") +
    theme(legend.position = "none")

  # Scatterplot of Shannon vs body size colored by parasitism
  p_para_scatter <- ggplot(div_para_combined,
                           aes(x = body_size, y = Shannon, color = Parasite)) +
    geom_point(alpha = 0.6, size = 3) +
    geom_smooth(method = "lm", se = TRUE) +
    facet_wrap(~ PrimerSet) +
    theme_minimal() +
    labs(title = "Body Size, Parasitism, and Diet Diversity",
         x = "Standard Length (mm)",
         y = "Shannon Diversity",
         color = "Parasitism")

  print(p_para_box / p_para_scatter)

} else {
  cat("\nNo parasitism data available in metadata. Skipping analysis.\n")
}
```

---

# LEVEL 4: BP vs. TP Within Paramormyrops kingsleyae

**The Core Question: Does EOD waveform associate with dietary niche partitioning?**

**Design consideration:** Waveform is confounded with geography in this dataset. BP waveforms are found in southern rivers (Bambomo, Apassa), while TP waveforms are found in northern rivers (Mengono, Iboundji). This level implements multiple analytical approaches to separate geographic effects from waveform effects.

**Analysis strategy:**
1. Two-way ANOVA/PERMANOVA (Region × Waveform) to assess relative contributions
2. Variance partitioning to quantify pure vs. shared effects
3. Waveform similarity across sites to test whether same-waveform fish in different regions are more similar than different-waveform fish in the same region
4. Prey selectivity analysis to distinguish composition differences from selection behavior

## 4.1 Prey Composition: BP vs. TP Emphasis

This section reuses the 4-panel visualization from 3.1 but emphasizes the BP vs. TP comparison within each region.

```{r level4-composition-emphasis, fig.width=14, fig.height=10}
cat("=== LEVEL 4.1: PREY COMPOSITION - BP vs. TP EMPHASIS ===\n\n")

# The 4-panel plot from 3.1 already shows Region × Waveform organization
# Here we add specific statistical comparisons

# Calculate mean composition similarity within groups
calc_group_similarity <- function(ps) {
  # Get OTU table
  otu_mat <- as.matrix(otu_table(ps))
  if (taxa_are_rows(ps)) {
    otu_mat <- t(otu_mat)
  }

  # Calculate Bray-Curtis distance
  dist_mat <- vegdist(otu_mat, method = "bray")

  # Get metadata
  meta <- data.frame(sample_data(ps))
  meta$Group <- paste(meta$Region, meta$Waveform_Type, sep = "_")

  # Calculate within-group vs. between-group similarities
  dist_df <- as.matrix(dist_mat) %>%
    as.data.frame() %>%
    rownames_to_column("Sample1") %>%
    pivot_longer(-Sample1, names_to = "Sample2", values_to = "Distance") %>%
    filter(Sample1 < Sample2)  # Remove duplicates and self-comparisons

  # Add group info
  dist_df$Group1 <- meta$Group[match(dist_df$Sample1, rownames(meta))]
  dist_df$Group2 <- meta$Group[match(dist_df$Sample2, rownames(meta))]
  dist_df$Region1 <- meta$Region[match(dist_df$Sample1, rownames(meta))]
  dist_df$Region2 <- meta$Region[match(dist_df$Sample2, rownames(meta))]
  dist_df$Waveform1 <- meta$Waveform_Type[match(dist_df$Sample1, rownames(meta))]
  dist_df$Waveform2 <- meta$Waveform_Type[match(dist_df$Sample2, rownames(meta))]

  # Categorize comparisons
  dist_df <- dist_df %>%
    mutate(
      Comparison = case_when(
        Group1 == Group2 ~ "Same Region, Same Waveform",
        Region1 == Region2 & Waveform1 != Waveform2 ~ "Same Region, Different Waveform",
        Region1 != Region2 & Waveform1 == Waveform2 ~ "Different Region, Same Waveform",
        Region1 != Region2 & Waveform1 != Waveform2 ~ "Different Region, Different Waveform",
        TRUE ~ "Other"
      )
    )

  return(dist_df)
}

# Calculate for both primer sets
dist_comp_ps1 <- calc_group_similarity(ps1_pk)
dist_comp_ps2 <- calc_group_similarity(ps2_pk)

# Summary statistics
cat("PRIMERSET 1 - Mean Bray-Curtis distances:\n")
dist_summary_ps1 <- dist_comp_ps1 %>%
  group_by(Comparison) %>%
  summarise(
    n = n(),
    mean_dist = mean(Distance),
    sd_dist = sd(Distance),
    .groups = "drop"
  ) %>%
  arrange(mean_dist)

print(dist_summary_ps1)

cat("\n\nPRIMERSET 2 - Mean Bray-Curtis distances:\n")
dist_summary_ps2 <- dist_comp_ps2 %>%
  group_by(Comparison) %>%
  summarise(
    n = n(),
    mean_dist = mean(Distance),
    sd_dist = sd(Distance),
    .groups = "drop"
  ) %>%
  arrange(mean_dist)

print(dist_summary_ps2)

# Visualization
dist_comp_ps1$PrimerSet <- "Primerset 1"
dist_comp_ps2$PrimerSet <- "Primerset 2"
dist_comp_combined <- bind_rows(dist_comp_ps1, dist_comp_ps2)

# Reorder comparison levels for logical display
dist_comp_combined$Comparison <- factor(
  dist_comp_combined$Comparison,
  levels = c("Same Region, Same Waveform",
             "Same Region, Different Waveform",
             "Different Region, Same Waveform",
             "Different Region, Different Waveform")
)

p_comp_distances <- ggplot(dist_comp_combined,
                            aes(x = Comparison, y = Distance, fill = Comparison)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~ PrimerSet) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  labs(
    title = "P. kingsleyae: Dietary Dissimilarity by Region and Waveform",
    subtitle = "Lower values = more similar diets",
    x = "Comparison Type",
    y = "Bray-Curtis Distance"
  ) +
  scale_fill_viridis_d()

print(p_comp_distances)

# Statistical test: Are different waveforms in same region more different than
# same waveforms in different regions?
cat("\n=== KEY STATISTICAL TEST ===\n")
cat("Question: Do waveforms predict diet better than geography?\n")
cat("Test: Same Waveform/Different Region vs. Different Waveform/Same Region\n\n")

test_ps1 <- dist_comp_ps1 %>%
  filter(Comparison %in% c("Same Region, Different Waveform",
                          "Different Region, Same Waveform"))

if (nrow(test_ps1) > 0) {
  cat("PRIMERSET 1:\n")
  wt_ps1 <- wilcox.test(Distance ~ Comparison, data = test_ps1)
  print(wt_ps1)

  cat("\nInterpretation:\n")
  if (wt_ps1$p.value < 0.05) {
    mean_same_wave <- mean(test_ps1$Distance[test_ps1$Comparison == "Different Region, Same Waveform"])
    mean_diff_wave <- mean(test_ps1$Distance[test_ps1$Comparison == "Same Region, Different Waveform"])

    if (mean_same_wave < mean_diff_wave) {
      cat("Same waveform in different regions are MORE similar than\n")
      cat("different waveforms in same region (p < 0.05).\n")
      cat("→ WAVEFORM predicts diet better than GEOGRAPHY\n")
    } else {
      cat("Different waveforms in same region are MORE similar than\n")
      cat("same waveform in different regions (p < 0.05).\n")
      cat("→ GEOGRAPHY predicts diet better than WAVEFORM\n")
    }
  } else {
    cat("No significant difference (p >= 0.05).\n")
    cat("Cannot distinguish waveform from geographic effects.\n")
  }
}

test_ps2 <- dist_comp_ps2 %>%
  filter(Comparison %in% c("Same Region, Different Waveform",
                          "Different Region, Same Waveform"))

if (nrow(test_ps2) > 0) {
  cat("\n\nPRIMERSET 2:\n")
  wt_ps2 <- wilcox.test(Distance ~ Comparison, data = test_ps2)
  print(wt_ps2)

  cat("\nInterpretation:\n")
  if (wt_ps2$p.value < 0.05) {
    mean_same_wave <- mean(test_ps2$Distance[test_ps2$Comparison == "Different Region, Same Waveform"])
    mean_diff_wave <- mean(test_ps2$Distance[test_ps2$Comparison == "Same Region, Different Waveform"])

    if (mean_same_wave < mean_diff_wave) {
      cat("Same waveform in different regions are MORE similar than\n")
      cat("different waveforms in same region (p < 0.05).\n")
      cat("→ WAVEFORM predicts diet better than GEOGRAPHY\n")
    } else {
      cat("Different waveforms in same region are MORE similar than\n")
      cat("same waveform in different regions (p < 0.05).\n")
      cat("→ GEOGRAPHY predicts diet better than WAVEFORM\n")
    }
  } else {
    cat("No significant difference (p >= 0.05).\n")
    cat("Cannot distinguish waveform from geographic effects.\n")
  }
}
```

## 4.2 Alpha Diversity: Two-Way ANOVA

Test whether Shannon diversity differs by Region and/or Waveform using two-way ANOVA with interaction. This approach allows us to assess:

1. Main effect of Region
2. Main effect of Waveform
3. Region × Waveform interaction
4. Marginal effects (Type II/III SS approach)

```{r level4-alpha-2way, fig.width=12, fig.height=8}
cat("=== LEVEL 4.2: ALPHA DIVERSITY - TWO-WAY ANOVA ===\n\n")

# Calculate diversity metrics
div1_pk <- estimate_richness(ps1_pk, measures = c("Shannon"))
meta1_pk <- data.frame(sample_data(ps1_pk), stringsAsFactors = FALSE)

# Add grouping variables
div1_pk$Region <- ifelse(meta1_pk$Locality_Name %in% c("Mengono Creek", "Iboundji Creek"),
                        "North", "South")
div1_pk$Waveform_Type <- meta1_pk$Waveform_Type
div1_pk$Locality_Name <- meta1_pk$Locality_Name
div1_pk$Sample_Name <- meta1_pk$Sample_Name

# Convert to factors
div1_pk$Region <- factor(div1_pk$Region, levels = c("North", "South"))
div1_pk$Waveform_Type <- factor(div1_pk$Waveform_Type)

# Same for PS2
div2_pk <- estimate_richness(ps2_pk, measures = c("Shannon"))
meta2_pk <- data.frame(sample_data(ps2_pk), stringsAsFactors = FALSE)

div2_pk$Region <- ifelse(meta2_pk$Locality_Name %in% c("Mengono Creek", "Iboundji Creek"),
                        "North", "South")
div2_pk$Waveform_Type <- meta2_pk$Waveform_Type
div2_pk$Locality_Name <- meta2_pk$Locality_Name
div2_pk$Sample_Name <- meta2_pk$Sample_Name

div2_pk$Region <- factor(div2_pk$Region, levels = c("North", "South"))
div2_pk$Waveform_Type <- factor(div2_pk$Waveform_Type)

# PRIMERSET 1 Analysis
cat("=== PRIMERSET 1 ===\n\n")

# Interactive model (Region × Waveform)
cat("Two-Way ANOVA with Interaction (Region × Waveform):\n")
anova_pk1_int <- aov(Shannon ~ Region * Waveform_Type, data = div1_pk)
print(summary(anova_pk1_int))

# Additive model (Region + Waveform)
cat("\n\nAdditive Model (Region + Waveform):\n")
anova_pk1_add <- aov(Shannon ~ Region + Waveform_Type, data = div1_pk)
print(summary(anova_pk1_add))

# Marginal effects using Type II SS
cat("\n\nMarginal Effects (Type II Sums of Squares):\n")
if (requireNamespace("car", quietly = TRUE)) {
  print(car::Anova(anova_pk1_add, type = "II"))
} else {
  cat("car package not available. Showing sequential tests:\n\n")

  cat("Region only:\n")
  anova_pk1_region <- aov(Shannon ~ Region, data = div1_pk)
  print(summary(anova_pk1_region))

  cat("\nWaveform only:\n")
  anova_pk1_wave <- aov(Shannon ~ Waveform_Type, data = div1_pk)
  print(summary(anova_pk1_wave))
}

# Summary statistics by group
cat("\n\nDescriptive Statistics:\n")
div1_pk %>%
  group_by(Region, Waveform_Type) %>%
  summarise(
    n = n(),
    mean_shannon = mean(Shannon),
    sd_shannon = sd(Shannon),
    .groups = "drop"
  ) %>%
  kable(digits = 3, booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# PRIMERSET 2 Analysis
cat("\n\n=== PRIMERSET 2 ===\n\n")

# Interactive model
cat("Two-Way ANOVA with Interaction (Region × Waveform):\n")
anova_pk2_int <- aov(Shannon ~ Region * Waveform_Type, data = div2_pk)
print(summary(anova_pk2_int))

# Additive model
cat("\n\nAdditive Model (Region + Waveform):\n")
anova_pk2_add <- aov(Shannon ~ Region + Waveform_Type, data = div2_pk)
print(summary(anova_pk2_add))

# Marginal effects
cat("\n\nMarginal Effects (Type II Sums of Squares):\n")
if (requireNamespace("car", quietly = TRUE)) {
  print(car::Anova(anova_pk2_add, type = "II"))
} else {
  cat("car package not available. Showing sequential tests:\n\n")

  cat("Region only:\n")
  anova_pk2_region <- aov(Shannon ~ Region, data = div2_pk)
  print(summary(anova_pk2_region))

  cat("\nWaveform only:\n")
  anova_pk2_wave <- aov(Shannon ~ Waveform_Type, data = div2_pk)
  print(summary(anova_pk2_wave))
}

# Summary statistics
cat("\n\nDescriptive Statistics:\n")
div2_pk %>%
  group_by(Region, Waveform_Type) %>%
  summarise(
    n = n(),
    mean_shannon = mean(Shannon),
    sd_shannon = sd(Shannon),
    .groups = "drop"
  ) %>%
  kable(digits = 3, booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Visualization: Interaction plot
div1_pk$PrimerSet <- "Primerset 1"
div2_pk$PrimerSet <- "Primerset 2"
div_combined <- bind_rows(div1_pk, div2_pk)

p_alpha_interact <- ggplot(div_combined,
                           aes(x = Region, y = Shannon, color = Waveform_Type,
                               group = Waveform_Type)) +
  stat_summary(fun = mean, geom = "point", size = 4) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.1) +
  facet_wrap(~ PrimerSet) +
  theme_minimal() +
  labs(
    title = "P. kingsleyae: Shannon Diversity by Region and Waveform",
    subtitle = "Interaction plot with mean ± SE",
    x = "Region",
    y = "Shannon Diversity",
    color = "Waveform"
  ) +
  scale_color_brewer(palette = "Set1")

print(p_alpha_interact)

# Boxplot version
p_alpha_box <- ggplot(div_combined,
                      aes(x = Region, y = Shannon, fill = Waveform_Type)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.3) +
  facet_wrap(~ PrimerSet) +
  theme_minimal() +
  labs(
    title = "P. kingsleyae: Shannon Diversity Distribution",
    x = "Region",
    y = "Shannon Diversity",
    fill = "Waveform"
  ) +
  scale_fill_brewer(palette = "Set1")

print(p_alpha_box)

cat("\n=== RESULT INTERPRETATION ===\n")
cat("Expected pattern if waveform matters: Significant Waveform main effect\n")
cat("Expected pattern if confounded: Significant Region effect, NS Waveform effect\n")
```

## 4.3 Beta Diversity: Two-Way PERMANOVA

Apply PERMANOVA to test Region and Waveform effects on diet composition (Bray-Curtis dissimilarity). Use marginal tests (by="margin") to assess independent contributions.

```{r level4-beta-2way, fig.width=14, fig.height=10}
cat("=== LEVEL 4.3: BETA DIVERSITY - TWO-WAY PERMANOVA ===\n\n")

# Prepare data for PERMANOVA - PS1
otu1_pk <- as.matrix(otu_table(ps1_pk))
if (taxa_are_rows(ps1_pk)) {
  otu1_pk <- t(otu1_pk)
}

meta1_pk_beta <- data.frame(sample_data(ps1_pk), stringsAsFactors = FALSE)
meta1_pk_beta$Region <- factor(
  ifelse(meta1_pk_beta$Locality_Name %in% c("Mengono Creek", "Iboundji Creek"),
         "North", "South"),
  levels = c("North", "South")
)
meta1_pk_beta$Waveform_Type <- factor(meta1_pk_beta$Waveform_Type)

# Ensure rownames match
rownames(meta1_pk_beta) <- rownames(otu1_pk)

dist1_pk <- vegdist(otu1_pk, method = "bray")

# Prepare data for PERMANOVA - PS2
otu2_pk <- as.matrix(otu_table(ps2_pk))
if (taxa_are_rows(ps2_pk)) {
  otu2_pk <- t(otu2_pk)
}

meta2_pk_beta <- data.frame(sample_data(ps2_pk), stringsAsFactors = FALSE)
meta2_pk_beta$Region <- factor(
  ifelse(meta2_pk_beta$Locality_Name %in% c("Mengono Creek", "Iboundji Creek"),
         "North", "South"),
  levels = c("North", "South")
)
meta2_pk_beta$Waveform_Type <- factor(meta2_pk_beta$Waveform_Type)

rownames(meta2_pk_beta) <- rownames(otu2_pk)

dist2_pk <- vegdist(otu2_pk, method = "bray")

# PRIMERSET 1 - PERMANOVA
cat("=== PRIMERSET 1 ===\n\n")

# Additive model with sequential testing
cat("Additive Model - Sequential (Region first):\n")
perm_pk1_seq_region <- adonis2(dist1_pk ~ Region + Waveform_Type,
                               data = meta1_pk_beta,
                               permutations = 999,
                               by = "terms")
print(perm_pk1_seq_region)

cat("\n\nAdditive Model - Sequential (Waveform first):\n")
perm_pk1_seq_wave <- adonis2(dist1_pk ~ Waveform_Type + Region,
                             data = meta1_pk_beta,
                             permutations = 999,
                             by = "terms")
print(perm_pk1_seq_wave)

# Marginal effects (most appropriate for confounded variables)
cat("\n\nAdditive Model - Marginal Effects (Type III):\n")
perm_pk1_margin <- adonis2(dist1_pk ~ Region + Waveform_Type,
                           data = meta1_pk_beta,
                           permutations = 999,
                           by = "margin")
print(perm_pk1_margin)

# Interactive model
cat("\n\nInteractive Model (Region × Waveform):\n")
perm_pk1_int <- adonis2(dist1_pk ~ Region * Waveform_Type,
                        data = meta1_pk_beta,
                        permutations = 999)
print(perm_pk1_int)

# Extract R² values for reporting
r2_region_ps1 <- perm_pk1_margin$R2[1]
r2_wave_ps1 <- perm_pk1_margin$R2[2]
p_region_ps1 <- perm_pk1_margin$`Pr(>F)`[1]
p_wave_ps1 <- perm_pk1_margin$`Pr(>F)`[2]

cat("\n\nEffect Size Summary (Marginal R²):\n")
cat(sprintf("  Region: R² = %.3f, p = %.3f\n", r2_region_ps1, p_region_ps1))
cat(sprintf("  Waveform: R² = %.3f, p = %.3f\n", r2_wave_ps1, p_wave_ps1))

# PRIMERSET 2 - PERMANOVA
cat("\n\n=== PRIMERSET 2 ===\n\n")

cat("Additive Model - Sequential (Region first):\n")
perm_pk2_seq_region <- adonis2(dist2_pk ~ Region + Waveform_Type,
                               data = meta2_pk_beta,
                               permutations = 999,
                               by = "terms")
print(perm_pk2_seq_region)

cat("\n\nAdditive Model - Sequential (Waveform first):\n")
perm_pk2_seq_wave <- adonis2(dist2_pk ~ Waveform_Type + Region,
                             data = meta2_pk_beta,
                             permutations = 999,
                             by = "terms")
print(perm_pk2_seq_wave)

cat("\n\nAdditive Model - Marginal Effects (Type III):\n")
perm_pk2_margin <- adonis2(dist2_pk ~ Region + Waveform_Type,
                           data = meta2_pk_beta,
                           permutations = 999,
                           by = "margin")
print(perm_pk2_margin)

cat("\n\nInteractive Model (Region × Waveform):\n")
perm_pk2_int <- adonis2(dist2_pk ~ Region * Waveform_Type,
                        data = meta2_pk_beta,
                        permutations = 999)
print(perm_pk2_int)

# Extract R² values
r2_region_ps2 <- perm_pk2_margin$R2[1]
r2_wave_ps2 <- perm_pk2_margin$R2[2]
p_region_ps2 <- perm_pk2_margin$`Pr(>F)`[1]
p_wave_ps2 <- perm_pk2_margin$`Pr(>F)`[2]

cat("\n\nEffect Size Summary (Marginal R²):\n")
cat(sprintf("  Region: R² = %.3f, p = %.3f\n", r2_region_ps2, p_region_ps2))
cat(sprintf("  Waveform: R² = %.3f, p = %.3f\n", r2_wave_ps2, p_wave_ps2))

# PCoA visualization
pcoa1_pk <- cmdscale(dist1_pk, eig = TRUE, k = 2)
var_exp1_pk <- pcoa1_pk$eig / sum(pcoa1_pk$eig) * 100

pcoa2_pk <- cmdscale(dist2_pk, eig = TRUE, k = 2)
var_exp2_pk <- pcoa2_pk$eig / sum(pcoa2_pk$eig) * 100

# Create PCoA data frames
pcoa1_df <- data.frame(
  PCoA1 = pcoa1_pk$points[, 1],
  PCoA2 = pcoa1_pk$points[, 2],
  meta1_pk_beta,
  PrimerSet = "Primerset 1"
)

pcoa2_df <- data.frame(
  PCoA1 = pcoa2_pk$points[, 1],
  PCoA2 = pcoa2_pk$points[, 2],
  meta2_pk_beta,
  PrimerSet = "Primerset 2"
)

pcoa_combined <- bind_rows(pcoa1_df, pcoa2_df)

# Plot: Color by Region, Shape by Waveform
p_pcoa_region_wave <- ggplot(pcoa_combined,
                              aes(x = PCoA1, y = PCoA2,
                                  color = Region, shape = Waveform_Type)) +
  geom_point(size = 4, alpha = 0.7) +
  facet_wrap(~ PrimerSet, scales = "free") +
  theme_minimal() +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "P. kingsleyae: Diet Composition by Region and Waveform",
    subtitle = "PCoA on Bray-Curtis dissimilarity",
    color = "Region",
    shape = "Waveform"
  ) +
  theme(legend.position = "right")

print(p_pcoa_region_wave)

# Alternative: Color by Waveform, Shape by Region
p_pcoa_wave_region <- ggplot(pcoa_combined,
                              aes(x = PCoA1, y = PCoA2,
                                  color = Waveform_Type, shape = Region)) +
  geom_point(size = 4, alpha = 0.7) +
  facet_wrap(~ PrimerSet, scales = "free") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "P. kingsleyae: Diet Composition (Waveform Emphasis)",
    subtitle = "PCoA on Bray-Curtis dissimilarity",
    color = "Waveform",
    shape = "Region"
  ) +
  theme(legend.position = "right")

print(p_pcoa_wave_region)

cat("\n=== RESULT INTERPRETATION ===\n")
cat("Marginal R² values indicate independent contribution of each factor\n")
cat("after accounting for the other.\n")
cat("If Waveform R² > 0.10 and significant: suggests dietary niche partitioning\n")
cat("If Region R² >> Waveform R²: geographic effects dominate\n")
```

## 4.4 Variance Partitioning

Decompose variation in diet composition into pure Region effect, pure Waveform effect, and shared (confounded) variation using variance partitioning.

```{r level4-varpart, fig.width=10, fig.height=8}
cat("=== LEVEL 4.4: VARIANCE PARTITIONING ===\n\n")

# This analysis partitions the variation explained by Region and Waveform
# into pure and shared components

# Prepare beta diversity results for variance partitioning
beta1_pk <- list(
  distance = dist1_pk,
  metadata = meta1_pk_beta
)

beta2_pk <- list(
  distance = dist2_pk,
  metadata = meta2_pk_beta
)

# Run variance partitioning (uses helper function from gut_analysis_2026_helpers.R)
cat("=== PRIMERSET 1 ===\n\n")
varpart_pk1 <- variance_partition_permanova(
  beta1_pk,
  var1_formula = "~ Region",
  var2_formula = "~ Waveform_Type"
)

cat("Variance Partitioning Results:\n")
print(varpart_pk1$varpart)

cat("\n\nSequential PERMANOVA (Region first):\n")
print(varpart_pk1$sequential_var1_first)

cat("\n\nSequential PERMANOVA (Waveform first):\n")
print(varpart_pk1$sequential_var2_first)

cat("\n\n#### Marginal PERMANOVA:\n\n")
kable(as.data.frame(varpart_pk1$marginal),
            digits = 3,
            booktabs = TRUE,
            caption = "Marginal PERMANOVA: Primerset 1") %>%
      kable_styling(latex_options = c("striped", "HOLD_position"),
                    font_size = 9,
                    full_width = FALSE)

# Extract variance components
# varpart returns: [a] = pure Region, [b] = pure Waveform, [c] = shared
vp1 <- varpart_pk1$varpart
if (!is.null(vp1$part$indfract)) {
  pure_region_ps1 <- vp1$part$indfract$Adj.R.squared[1]
  pure_wave_ps1 <- vp1$part$indfract$Adj.R.squared[2]
  shared_ps1 <- vp1$part$indfract$Adj.R.squared[3]

  cat("\n\nVariance Components (Adjusted R²):\n")
  cat(sprintf("  Pure Region: %.4f\n", pure_region_ps1))
  cat(sprintf("  Pure Waveform: %.4f\n", pure_wave_ps1))
  cat(sprintf("  Shared (confounded): %.4f\n", shared_ps1))
  cat(sprintf("  Total explained: %.4f\n",
              pure_region_ps1 + pure_wave_ps1 + shared_ps1))
}

cat("\n\n=== PRIMERSET 2 ===\n\n")
varpart_pk2 <- variance_partition_permanova(
  beta2_pk,
  var1_formula = "~ Region",
  var2_formula = "~ Waveform_Type"
)

cat("Variance Partitioning Results:\n")
print(varpart_pk2$varpart)

cat("\n\nSequential PERMANOVA (Region first):\n")
print(varpart_pk2$sequential_var1_first)

cat("\n\nSequential PERMANOVA (Waveform first):\n")
print(varpart_pk2$sequential_var2_first)

cat("\n\n#### Marginal PERMANOVA:\n\n")
kable(as.data.frame(varpart_pk2$marginal),
            digits = 3,
            booktabs = TRUE,
            caption = "Marginal PERMANOVA: Primerset 2") %>%
      kable_styling(latex_options = c("striped", "HOLD_position"),
                    font_size = 9,
                    full_width = FALSE)

vp2 <- varpart_pk2$varpart
if (!is.null(vp2$part$indfract)) {
  pure_region_ps2 <- vp2$part$indfract$Adj.R.squared[1]
  pure_wave_ps2 <- vp2$part$indfract$Adj.R.squared[2]
  shared_ps2 <- vp2$part$indfract$Adj.R.squared[3]

  cat("\n\nVariance Components (Adjusted R²):\n")
  cat(sprintf("  Pure Region: %.4f\n", pure_region_ps2))
  cat(sprintf("  Pure Waveform: %.4f\n", pure_wave_ps2))
  cat(sprintf("  Shared (confounded): %.4f\n", shared_ps2))
  cat(sprintf("  Total explained: %.4f\n",
              pure_region_ps2 + pure_wave_ps2 + shared_ps2))
}

# Create summary visualization
if (!is.null(vp1$part$indfract) && !is.null(vp2$part$indfract)) {

  varpart_summary <- data.frame(
    PrimerSet = rep(c("PS1", "PS2"), each = 3),
    Component = rep(c("Pure Region", "Pure Waveform", "Shared"), 2),
    R2 = c(pure_region_ps1, pure_wave_ps1, shared_ps1,
           pure_region_ps2, pure_wave_ps2, shared_ps2)
  )

  # Handle negative values (set to 0 for visualization)
  varpart_summary$R2_plot <- pmax(varpart_summary$R2, 0)

  p_varpart <- ggplot(varpart_summary,
                      aes(x = PrimerSet, y = R2_plot, fill = Component)) +
    geom_bar(stat = "identity", position = "stack") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set2") +
    labs(
      title = "Variance Partitioning: Region vs. Waveform",
      subtitle = "Decomposition of explained variation in diet composition",
      x = "Primer Set",
      y = "Proportion of Variance (Adjusted R²)",
      fill = "Component"
    ) +
    geom_text(
      data = varpart_summary %>%
        group_by(PrimerSet) %>%
        mutate(
          pos = cumsum(R2_plot) - 0.5 * R2_plot,
          label = sprintf("%.3f", R2)
        ),
      aes(y = pos, label = label),
      size = 3
    )

  print(p_varpart)
}

cat("\n=== INTERPRETATION ===\n")
cat("Pure Waveform variance indicates effect independent of geography.\n")
cat("Large Shared variance indicates confounding between Region and Waveform.\n")
cat("If Pure Waveform > 0.05: suggests waveform has independent effect on diet.\n")
```

## 4.5 Waveform Similarity Across Sites - CRITICAL TEST

This is the key test to determine whether waveform predicts diet better than geography. Compare dietary similarity of:
- Same waveform in different regions vs.
- Different waveforms in same region

If same-waveform-different-region < different-waveform-same-region → waveform matters more than geography.

```{r level4-waveform-similarity, fig.width=12, fig.height=8}
cat("=== LEVEL 4.5: WAVEFORM SIMILARITY ACROSS SITES ===\n")
cat("THE CRITICAL TEST FOR WAVEFORM EFFECT\n\n")

# This analysis was already performed in section 4.1
# Here we formalize it as the key statistical test

# Extract pairwise distances and categorize
analyze_waveform_geography <- function(dist_mat, metadata) {

  # Convert distance matrix to data frame
  dist_df <- as.matrix(dist_mat) %>%
    as.data.frame() %>%
    rownames_to_column("Sample1") %>%
    pivot_longer(-Sample1, names_to = "Sample2", values_to = "Distance") %>%
    filter(Sample1 < Sample2)  # Remove self and duplicate comparisons

  # Add metadata
  dist_df$Region1 <- metadata$Region[match(dist_df$Sample1, rownames(metadata))]
  dist_df$Region2 <- metadata$Region[match(dist_df$Sample2, rownames(metadata))]
  dist_df$Waveform1 <- metadata$Waveform_Type[match(dist_df$Sample1, rownames(metadata))]
  dist_df$Waveform2 <- metadata$Waveform_Type[match(dist_df$Sample2, rownames(metadata))]

  # Create comparison categories
  dist_df <- dist_df %>%
    mutate(
      Same_Region = Region1 == Region2,
      Same_Waveform = Waveform1 == Waveform2,
      Comparison = case_when(
        Same_Region & Same_Waveform ~ "Same Region, Same Waveform",
        Same_Region & !Same_Waveform ~ "Same Region, Different Waveform",
        !Same_Region & Same_Waveform ~ "Different Region, Same Waveform",
        !Same_Region & !Same_Waveform ~ "Different Region, Different Waveform"
      )
    )

  return(dist_df)
}

# Analyze both primer sets
dist_analysis_ps1 <- analyze_waveform_geography(dist1_pk, meta1_pk_beta)
dist_analysis_ps2 <- analyze_waveform_geography(dist2_pk, meta2_pk_beta)

# KEY STATISTICAL TEST
cat("=== THE KEY COMPARISON ===\n")
cat("H0: Same waveform in different regions has same dietary similarity as\n")
cat("    different waveforms in same region\n")
cat("HA: These groups differ in dietary similarity\n\n")

# Extract the two key comparison groups for PS1
key_comp_ps1 <- dist_analysis_ps1 %>%
  filter(Comparison %in% c("Same Region, Different Waveform",
                          "Different Region, Same Waveform"))

if (nrow(key_comp_ps1) > 0) {
  cat("=== PRIMERSET 1 ===\n\n")

  # Summary statistics
  key_summary_ps1 <- key_comp_ps1 %>%
    group_by(Comparison) %>%
    summarise(
      n = n(),
      mean_dist = mean(Distance),
      median_dist = median(Distance),
      sd_dist = sd(Distance),
      .groups = "drop"
    )

  print(key_summary_ps1)

  # Statistical test (Wilcoxon rank-sum, non-parametric)
  cat("\n\nWilcoxon Rank-Sum Test:\n")
  wt_key_ps1 <- wilcox.test(Distance ~ Comparison, data = key_comp_ps1)
  print(wt_key_ps1)

  # Effect size (rank-biserial correlation)
  if (requireNamespace("rcompanion", quietly = TRUE)) {
    eff_ps1 <- rcompanion::wilcoxonR(x = key_comp_ps1$Distance,
                                    g = key_comp_ps1$Comparison)
    cat("\nEffect size (rank-biserial r):", round(eff_ps1, 3), "\n")
  }

  # Interpretation
  cat("\n\nINTERPRETATION:\n")
  mean_same_wave_ps1 <- key_summary_ps1$mean_dist[
    key_summary_ps1$Comparison == "Different Region, Same Waveform"
  ]
  mean_diff_wave_ps1 <- key_summary_ps1$mean_dist[
    key_summary_ps1$Comparison == "Same Region, Different Waveform"
  ]

  if (wt_key_ps1$p.value < 0.05) {
    if (mean_same_wave_ps1 < mean_diff_wave_ps1) {
      cat("RESULT: Fish with same waveform in different regions have MORE\n")
      cat("        similar diets than fish with different waveforms in the\n")
      cat("        same region (p < 0.05).\n\n")
      cat("CONCLUSION: WAVEFORM predicts diet better than GEOGRAPHY.\n")
      cat("            This supports dietary niche partitioning by waveform.\n")
    } else {
      cat("RESULT: Fish in same region have MORE similar diets than fish\n")
      cat("        with same waveform in different regions (p < 0.05).\n\n")
      cat("CONCLUSION: GEOGRAPHY predicts diet better than WAVEFORM.\n")
      cat("            Dietary differences are driven by site-level factors.\n")
    }
  } else {
    cat("RESULT: No significant difference (p = ",
        format.pval(wt_key_ps1$p.value, digits = 3), ")\n\n")
    cat("CONCLUSION: Cannot distinguish waveform effect from geographic effect.\n")
    cat("            The confounding is too strong to separate these factors.\n")
  }
}

# Repeat for PS2
key_comp_ps2 <- dist_analysis_ps2 %>%
  filter(Comparison %in% c("Same Region, Different Waveform",
                          "Different Region, Same Waveform"))

if (nrow(key_comp_ps2) > 0) {
  cat("\n\n=== PRIMERSET 2 ===\n\n")

  key_summary_ps2 <- key_comp_ps2 %>%
    group_by(Comparison) %>%
    summarise(
      n = n(),
      mean_dist = mean(Distance),
      median_dist = median(Distance),
      sd_dist = sd(Distance),
      .groups = "drop"
    )

  print(key_summary_ps2)

  cat("\n\nWilcoxon Rank-Sum Test:\n")
  wt_key_ps2 <- wilcox.test(Distance ~ Comparison, data = key_comp_ps2)
  print(wt_key_ps2)

  if (requireNamespace("rcompanion", quietly = TRUE)) {
    eff_ps2 <- rcompanion::wilcoxonR(x = key_comp_ps2$Distance,
                                    g = key_comp_ps2$Comparison)
    cat("\nEffect size (rank-biserial r):", round(eff_ps2, 3), "\n")
  }

  cat("\n\nINTERPRETATION:\n")
  mean_same_wave_ps2 <- key_summary_ps2$mean_dist[
    key_summary_ps2$Comparison == "Different Region, Same Waveform"
  ]
  mean_diff_wave_ps2 <- key_summary_ps2$mean_dist[
    key_summary_ps2$Comparison == "Same Region, Different Waveform"
  ]

  if (wt_key_ps2$p.value < 0.05) {
    if (mean_same_wave_ps2 < mean_diff_wave_ps2) {
      cat("RESULT: Fish with same waveform in different regions have MORE\n")
      cat("        similar diets than fish with different waveforms in the\n")
      cat("        same region (p < 0.05).\n\n")
      cat("CONCLUSION: WAVEFORM predicts diet better than GEOGRAPHY.\n")
      cat("            This supports dietary niche partitioning by waveform.\n")
    } else {
      cat("RESULT: Fish in same region have MORE similar diets than fish\n")
      cat("        with same waveform in different regions (p < 0.05).\n\n")
      cat("CONCLUSION: GEOGRAPHY predicts diet better than WAVEFORM.\n")
      cat("            Dietary differences are driven by site-level factors.\n")
    }
  } else {
    cat("RESULT: No significant difference (p = ",
        format.pval(wt_key_ps2$p.value, digits = 3), ")\n\n")
    cat("CONCLUSION: Cannot distinguish waveform effect from geographic effect.\n")
    cat("            The confounding is too strong to separate these factors.\n")
  }
}

# Visualization
dist_analysis_ps1$PrimerSet <- "Primerset 1"
dist_analysis_ps2$PrimerSet <- "Primerset 2"
dist_combined_analysis <- bind_rows(dist_analysis_ps1, dist_analysis_ps2)

# Reorder for clarity
dist_combined_analysis$Comparison <- factor(
  dist_combined_analysis$Comparison,
  levels = c("Same Region, Same Waveform",
             "Same Region, Different Waveform",
             "Different Region, Same Waveform",
             "Different Region, Different Waveform")
)

p_key_test <- ggplot(
  dist_combined_analysis %>%
    filter(Comparison %in% c("Same Region, Different Waveform",
                            "Different Region, Same Waveform")),
  aes(x = Comparison, y = Distance, fill = Comparison)
) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.2, alpha = 0.7) +
  facet_wrap(~ PrimerSet) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  labs(
    title = "THE KEY TEST: Waveform vs. Geography",
    subtitle = "Lower distance = more similar diets",
    x = "",
    y = "Bray-Curtis Distance"
  ) +
  scale_fill_brewer(palette = "Set2")

print(p_key_test)

# All four comparison categories
p_all_comparisons <- ggplot(dist_combined_analysis,
                            aes(x = Comparison, y = Distance, fill = Comparison)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.3, alpha = 0.7, outlier.alpha = 0.3) +
  facet_wrap(~ PrimerSet) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  labs(
    title = "All Pairwise Comparisons: Region × Waveform",
    x = "",
    y = "Bray-Curtis Distance"
  ) +
  scale_fill_viridis_d()

print(p_all_comparisons)

# Statistical test across all four categories
cat("\n\n=== OVERALL COMPARISON (All Four Categories) ===\n\n")

cat("PRIMERSET 1:\n")
kw_all_ps1 <- kruskal.test(Distance ~ Comparison, data = dist_analysis_ps1)
print(kw_all_ps1)

if (kw_all_ps1$p.value < 0.05 && requireNamespace("FSA", quietly = TRUE)) {
  cat("\n#### Post-hoc pairwise comparisons (Dunn test):\n\n")
  dunn_ps1 <- FSA::dunnTest(Distance ~ Comparison, data = dist_analysis_ps1,
                           method = "bonferroni")
  # Abbreviate comparison names for table width
  dunn_df <- as.data.frame(dunn_ps1$res)
  dunn_df$Comparison <- gsub("Different Region, Different Waveform", "DR-DW", dunn_df$Comparison)
  dunn_df$Comparison <- gsub("Different Region, Same Waveform", "DR-SW", dunn_df$Comparison)
  dunn_df$Comparison <- gsub("Same Region, Different Waveform", "SR-DW", dunn_df$Comparison)
  dunn_df$Comparison <- gsub("Same Region, Same Waveform", "SR-SW", dunn_df$Comparison)

  cat("Abbreviations: DR=Different Region, SR=Same Region, DW=Different Waveform, SW=Same Waveform\n\n")

  kable(dunn_df, digits = 4, booktabs = TRUE,
        caption = "Dunn Test: Primerset 1") %>%
    kable_styling(latex_options = c("striped", "HOLD_position", "scale_down"),
                  full_width = FALSE, font_size = 9)
}

cat("\n\nPRIMERSET 2:\n")
kw_all_ps2 <- kruskal.test(Distance ~ Comparison, data = dist_analysis_ps2)
print(kw_all_ps2)

if (kw_all_ps2$p.value < 0.05 && requireNamespace("FSA", quietly = TRUE)) {
  cat("\n#### Post-hoc pairwise comparisons (Dunn test):\n\n")
  dunn_ps2 <- FSA::dunnTest(Distance ~ Comparison, data = dist_analysis_ps2,
                           method = "bonferroni")
  # Abbreviate comparison names for table width
  dunn_df <- as.data.frame(dunn_ps2$res)
  dunn_df$Comparison <- gsub("Different Region, Different Waveform", "DR-DW", dunn_df$Comparison)
  dunn_df$Comparison <- gsub("Different Region, Same Waveform", "DR-SW", dunn_df$Comparison)
  dunn_df$Comparison <- gsub("Same Region, Different Waveform", "SR-DW", dunn_df$Comparison)
  dunn_df$Comparison <- gsub("Same Region, Same Waveform", "SR-SW", dunn_df$Comparison)

  cat("Abbreviations: DR=Different Region, SR=Same Region, DW=Different Waveform, SW=Same Waveform\n\n")

  kable(dunn_df, digits = 4, booktabs = TRUE,
        caption = "Dunn Test: Primerset 2") %>%
    kable_styling(latex_options = c("striped", "HOLD_position", "scale_down"),
                  full_width = FALSE, font_size = 9)
}
```

## 4.6 Prey Selectivity Analysis

Integrate environmental prey availability data to distinguish between:
1. **Compositional differences**: BP and TP eat different prey
2. **Selectivity differences**: BP and TP show different preferences for the same available prey

This requires D-frame net data from the field sites. If available, calculate Ivlev's electivity index (E) and test whether selectivity profiles differ between waveforms.

```{r level4-selectivity, fig.width=12, fig.height=10}
cat("=== LEVEL 4.6: PREY SELECTIVITY ANALYSIS ===\n\n")

# Check if environmental data is available
env_file <- file.path(root, "input_data", "gabon_inverts_compiled_cleaned_2026_final.csv")

if (file.exists(env_file)) {

  cat("Environmental prey availability data found.\n")
  cat("Loading and processing...\n\n")

  # Load environmental data
  env_raw <- read.csv(env_file, stringsAsFactors = FALSE)
  colnames(env_raw) <- trimws(colnames(env_raw))

  # Clean and aggregate environmental data
  env_data <- env_raw %>%
    mutate(Count = as.numeric(Count)) %>%
    mutate(Family_clean = sapply(Family, clean_family_name)) %>%
    filter(!is.na(Family_clean), !is.na(Count), Count > 0) %>%
    select(Site = Name, Family = Family_clean, Count)

  # Map locality names to site names in environmental data
  # This mapping depends on your data - adjust as needed
  locality_to_site <- c(
    "Bambomo" = "Bambomo Creek",
    "Apassa" = "Ipassa Creek",
    "Mengono" = "Mengono Creek",
    "Iboundji" = "Iboundji Creek"
  )

  # Get family-level gut contents for P. kingsleyae
  ps1_pk_family_counts <- extract_family_counts(ps1_pk)
  ps2_pk_family_counts <- extract_family_counts(ps2_pk)

  # Add metadata
  meta1_sel <- data.frame(sample_data(ps1_pk))
  meta1_sel$SampleID <- rownames(meta1_sel)

  meta2_sel <- data.frame(sample_data(ps2_pk))
  meta2_sel$SampleID <- rownames(meta2_sel)

  ps1_pk_family_counts <- ps1_pk_family_counts %>%
    left_join(meta1_sel %>%
                select(SampleID, Locality_Name, Waveform_Type, Region),
              by = "SampleID")

  ps2_pk_family_counts <- ps2_pk_family_counts %>%
    left_join(meta2_sel %>%
                select(SampleID, Locality_Name, Waveform_Type, Region),
              by = "SampleID")

  # Function to calculate selectivity for an individual fish
  calc_individual_selectivity <- function(sample_id, diet_data, env_data_site, metadata) {

    # Get diet for this individual fish
    diet_individual <- diet_data %>%
      filter(SampleID == sample_id)

    if (nrow(diet_individual) == 0) return(NULL)

    # Calculate selectivity
    selectivity <- calculate_selectivity(
      diet_counts = diet_individual %>% select(Family, Count),
      env_counts = env_data_site %>% select(Family, Count)
    )

    # Add metadata
    meta_row <- metadata[metadata$SampleID == sample_id, ]
    selectivity$SampleID <- sample_id
    selectivity$Locality <- meta_row$Locality_Name
    selectivity$Waveform <- meta_row$Waveform_Type
    selectivity$Region <- meta_row$Region

    return(selectivity)
  }

  # Calculate selectivity for EACH INDIVIDUAL FISH
  selectivity_results_ps1 <- list()
  selectivity_results_ps2 <- list()

  # Get unique samples with locality info
  samples_ps1 <- ps1_pk_family_counts %>%
    distinct(SampleID, Locality_Name, Waveform_Type, Region)

  for (i in 1:nrow(samples_ps1)) {
    sample_id <- samples_ps1$SampleID[i]
    locality <- samples_ps1$Locality_Name[i]

    # Get site name for environmental data
    site_name <- locality_to_site[locality]
    if (is.na(site_name)) next

    # Get environmental data for this site
    env_site <- env_data %>% filter(Site == site_name)
    if (nrow(env_site) == 0) next

    # Calculate selectivity for this individual
    selectivity <- calc_individual_selectivity(
      sample_id = sample_id,
      diet_data = ps1_pk_family_counts,
      env_data_site = env_site,
      metadata = meta1_sel
    )

    if (!is.null(selectivity)) {
      selectivity_results_ps1[[sample_id]] <- selectivity
    }
  }

  # Combine results
  if (length(selectivity_results_ps1) > 0) {
    selectivity_ps1_df <- bind_rows(selectivity_results_ps1)

    cat("PRIMERSET 1 - Selectivity Analysis (Individual-level):\n")
    cat("  Individuals analyzed:", n_distinct(selectivity_ps1_df$SampleID), "\n")
    cat("  Total family-individual combinations:", nrow(selectivity_ps1_df), "\n\n")

    # Summary statistics by Waveform
    cat("Mean Ivlev's E by Waveform:\n")
    selectivity_summary_ps1 <- selectivity_ps1_df %>%
      group_by(Waveform, Region) %>%
      summarise(
        n_individuals = n_distinct(SampleID),
        n_families = n_distinct(Family),
        mean_E = mean(Ivlev_E, na.rm = TRUE),
        mean_abs_E = mean(abs(Ivlev_E), na.rm = TRUE),
        n_preference = sum(Selection_Type == "Preference"),
        n_avoidance = sum(Selection_Type == "Avoidance"),
        .groups = "drop"
      )

    print(selectivity_summary_ps1)

    # Test: Do BP and TP differ in selectivity profiles?
    # Create wide matrix of Ivlev's E values (one row per individual)
    selectivity_wide_ps1 <- selectivity_ps1_df %>%
      select(SampleID, Region, Waveform, Locality, Family, Ivlev_E) %>%
      group_by(SampleID, Region, Waveform, Locality, Family) %>%
      summarise(Ivlev_E = mean(Ivlev_E, na.rm = TRUE), .groups = "drop") %>%
      pivot_wider(names_from = Family, values_from = Ivlev_E, values_fill = 0)

    if (nrow(selectivity_wide_ps1) > 3) {  # Need at least 4 samples
      selectivity_matrix_ps1 <- as.matrix(selectivity_wide_ps1[, -(1:4)])
      rownames(selectivity_matrix_ps1) <- selectivity_wide_ps1$SampleID

      selectivity_meta_ps1 <- selectivity_wide_ps1 %>%
        select(SampleID, Region, Waveform, Locality)

      # Distance matrix
      selectivity_dist_ps1 <- vegdist(selectivity_matrix_ps1, method = "euclidean")

      # PERMANOVA: Does waveform predict selectivity?
      cat("\n\nPERMANOVA: Individual Selectivity ~ Waveform + Region\n")
      cat("Sample size: ", nrow(selectivity_meta_ps1), " individuals\n")

      if (n_distinct(selectivity_meta_ps1$Waveform) > 1 &&
          n_distinct(selectivity_meta_ps1$Region) > 0) {

        # Check degrees of freedom
        n_samples <- nrow(selectivity_meta_ps1)
        n_waveforms <- n_distinct(selectivity_meta_ps1$Waveform)
        n_regions <- n_distinct(selectivity_meta_ps1$Region)
        model_df <- (n_waveforms - 1) + (n_regions - 1)
        residual_df <- n_samples - model_df - 1

        cat("Model df:", model_df, ", Residual df:", residual_df, "\n\n")

        if (residual_df > 0) {
          perm_selectivity_ps1 <- adonis2(
            selectivity_dist_ps1 ~ Waveform + Region,
            data = selectivity_meta_ps1,
            permutations = 999,
            by = "margin"
          )
          print(perm_selectivity_ps1)

          cat("\nInterpretation:\n")
          if (!is.na(perm_selectivity_ps1$`Pr(>F)`[1]) &&
              perm_selectivity_ps1$`Pr(>F)`[1] < 0.05) {
            cat("Waveform DOES predict selectivity (p < 0.05)\n")
            cat("→ BP and TP show different prey preferences\n")
          } else {
            cat("Waveform does NOT predict selectivity (p >= 0.05)\n")
            cat("→ BP and TP show similar prey preferences\n")
            cat("   (compositional differences driven by availability, not selection)\n")
          }
        } else {
          cat("No residual component - insufficient samples for this model.\n")
          cat("Try testing Waveform alone:\n\n")

          perm_selectivity_wf_ps1 <- adonis2(
            selectivity_dist_ps1 ~ Waveform,
            data = selectivity_meta_ps1,
            permutations = 999
          )
          print(perm_selectivity_wf_ps1)
        }
      } else {
        cat("Insufficient waveform variation for PERMANOVA.\n")
      }

      # PCoA visualization
      pcoa_sel_ps1 <- cmdscale(selectivity_dist_ps1, eig = TRUE, k = 2)

      pcoa_sel_df_ps1 <- data.frame(
        PCoA1 = pcoa_sel_ps1$points[, 1],
        PCoA2 = pcoa_sel_ps1$points[, 2],
        selectivity_meta_ps1
      )

      p_selectivity_ps1 <- ggplot(pcoa_sel_df_ps1,
                                  aes(x = PCoA1, y = PCoA2,
                                      color = Waveform, shape = Region)) +
        geom_point(size = 4, alpha = 0.7) +
        theme_minimal() +
        scale_color_brewer(palette = "Set1") +
        labs(
          title = "PS1: Individual Prey Selectivity Profiles (BP vs. TP)",
          subtitle = "PCoA on Ivlev's E values (each point = one fish)",
          color = "Waveform",
          shape = "Region"
        )

      print(p_selectivity_ps1)
    } else {
      cat("Insufficient samples for PERMANOVA (need at least 4 individuals).\n")
    }

  } else {
    cat("Could not calculate selectivity for PS1 (no matching environmental data)\n")
  }

  # ===== PRIMERSET 2 ANALYSIS =====

  # Get unique samples with locality info for PS2
  samples_ps2 <- ps2_pk_family_counts %>%
    distinct(SampleID, Locality_Name, Waveform_Type, Region)

  for (i in 1:nrow(samples_ps2)) {
    sample_id <- samples_ps2$SampleID[i]
    locality <- samples_ps2$Locality_Name[i]

    # Get site name for environmental data
    site_name <- locality_to_site[locality]
    if (is.na(site_name)) next

    # Get environmental data for this site
    env_site <- env_data %>% filter(Site == site_name)
    if (nrow(env_site) == 0) next

    # Calculate selectivity for this individual
    selectivity <- calc_individual_selectivity(
      sample_id = sample_id,
      diet_data = ps2_pk_family_counts,
      env_data_site = env_site,
      metadata = meta2_sel
    )

    if (!is.null(selectivity)) {
      selectivity_results_ps2[[sample_id]] <- selectivity
    }
  }

  # Combine and analyze PS2 results
  if (length(selectivity_results_ps2) > 0) {
    selectivity_ps2_df <- bind_rows(selectivity_results_ps2)

    cat("\n\n==============================================\n")
    cat("PRIMERSET 2 - Selectivity Analysis (Individual-level):\n")
    cat("  Individuals analyzed:", n_distinct(selectivity_ps2_df$SampleID), "\n")
    cat("  Total family-individual combinations:",
        nrow(selectivity_ps2_df), "\n\n")

    # Summary statistics by Waveform
    cat("Mean Ivlev's E by Waveform:\n")
    selectivity_summary_ps2 <- selectivity_ps2_df %>%
      group_by(Waveform, Region) %>%
      summarise(
        n_individuals = n_distinct(SampleID),
        n_families = n_distinct(Family),
        mean_E = mean(Ivlev_E, na.rm = TRUE),
        mean_abs_E = mean(abs(Ivlev_E), na.rm = TRUE),
        n_preference = sum(Selection_Type == "Preference"),
        n_avoidance = sum(Selection_Type == "Avoidance"),
        .groups = "drop"
      )

    print(selectivity_summary_ps2)

    # Create wide matrix for PERMANOVA
    selectivity_wide_ps2 <- selectivity_ps2_df %>%
      select(SampleID, Region, Waveform, Locality, Family, Ivlev_E) %>%
      group_by(SampleID, Region, Waveform, Locality, Family) %>%
      summarise(Ivlev_E = mean(Ivlev_E, na.rm = TRUE), .groups = "drop") %>%
      pivot_wider(names_from = Family, values_from = Ivlev_E, values_fill = 0)

    if (nrow(selectivity_wide_ps2) > 3) {
      selectivity_matrix_ps2 <- as.matrix(selectivity_wide_ps2[, -(1:4)])
      rownames(selectivity_matrix_ps2) <- selectivity_wide_ps2$SampleID

      selectivity_meta_ps2 <- selectivity_wide_ps2 %>%
        select(SampleID, Region, Waveform, Locality)

      # Distance matrix
      selectivity_dist_ps2 <- vegdist(selectivity_matrix_ps2, method = "euclidean")

      # PERMANOVA
      cat("\n\nPERMANOVA: Individual Selectivity ~ Waveform + Region\n")
      cat("Sample size: ", nrow(selectivity_meta_ps2), " individuals\n")

      if (n_distinct(selectivity_meta_ps2$Waveform) > 1 &&
          n_distinct(selectivity_meta_ps2$Region) > 0) {

        n_samples <- nrow(selectivity_meta_ps2)
        n_waveforms <- n_distinct(selectivity_meta_ps2$Waveform)
        n_regions <- n_distinct(selectivity_meta_ps2$Region)
        model_df <- (n_waveforms - 1) + (n_regions - 1)
        residual_df <- n_samples - model_df - 1

        cat("Model df:", model_df, ", Residual df:", residual_df, "\n\n")

        if (residual_df > 0) {
          perm_selectivity_ps2 <- adonis2(
            selectivity_dist_ps2 ~ Waveform + Region,
            data = selectivity_meta_ps2,
            permutations = 999,
            by = "margin"
          )
          print(perm_selectivity_ps2)

          cat("\nInterpretation:\n")
          if (!is.na(perm_selectivity_ps2$`Pr(>F)`[1]) &&
              perm_selectivity_ps2$`Pr(>F)`[1] < 0.05) {
            cat("Waveform DOES predict selectivity (p < 0.05)\n")
            cat("→ BP and TP show different prey preferences\n")
          } else {
            cat("Waveform does NOT predict selectivity (p >= 0.05)\n")
            cat("→ BP and TP show similar prey preferences\n")
          }
        } else {
          cat("No residual component - insufficient samples for this model.\n")
          cat("Try testing Waveform alone:\n\n")

          perm_selectivity_wf_ps2 <- adonis2(
            selectivity_dist_ps2 ~ Waveform,
            data = selectivity_meta_ps2,
            permutations = 999
          )
          print(perm_selectivity_wf_ps2)
        }
      } else {
        cat("Insufficient waveform variation for PERMANOVA.\n")
      }

      # PCoA visualization
      pcoa_sel_ps2 <- cmdscale(selectivity_dist_ps2, eig = TRUE, k = 2)

      pcoa_sel_df_ps2 <- data.frame(
        PCoA1 = pcoa_sel_ps2$points[, 1],
        PCoA2 = pcoa_sel_ps2$points[, 2],
        selectivity_meta_ps2
      )

      p_selectivity_ps2 <- ggplot(pcoa_sel_df_ps2,
                                  aes(x = PCoA1, y = PCoA2,
                                      color = Waveform, shape = Region)) +
        geom_point(size = 4, alpha = 0.7) +
        theme_minimal() +
        scale_color_brewer(palette = "Set1") +
        labs(
          title = "PS2: Individual Prey Selectivity Profiles (BP vs. TP)",
          subtitle = "PCoA on Ivlev's E values (each point = one fish)",
          color = "Waveform",
          shape = "Region"
        )

      print(p_selectivity_ps2)
    } else {
      cat("Insufficient samples for PERMANOVA (need at least 4 individuals).\n")
    }

  } else {
    cat("Could not calculate selectivity for PS2 (no matching environmental data)\n")
  }

} else {
  cat("Environmental prey availability data not found at:\n")
  cat(env_file, "\n\n")
  cat("Skipping prey selectivity analysis.\n")
  cat("This analysis requires D-frame net collection data.\n")
}
```

## 4.7 Mean Selectivity Comparison

Calculate mean absolute selectivity (|E|) per fish to measure overall "choosiness", then compare BP vs. TP.

```{r level4-mean-selectivity, fig.width=10, fig.height=6}
cat("=== LEVEL 4.7: MEAN SELECTIVITY COMPARISON ===\n\n")

if (exists("selectivity_ps1_df") && nrow(selectivity_ps1_df) > 0) {

  # Calculate mean absolute selectivity per individual fish
  mean_selectivity <- selectivity_ps1_df %>%
    group_by(SampleID, Region, Waveform, Locality) %>%
    summarise(
      mean_abs_E = mean(abs(Ivlev_E), na.rm = TRUE),
      n_families = n(),
      .groups = "drop"
    )

  cat("Mean Absolute Selectivity by Individual:\n")
  cat("(Higher |E| = more selective/choosy)\n\n")
  print(head(mean_selectivity, 10))

  cat("\n\nSummary by Waveform:\n")
  print(mean_selectivity %>%
    group_by(Waveform) %>%
    summarise(
      n_fish = n(),
      mean_selectivity = mean(mean_abs_E),
      sd_selectivity = sd(mean_abs_E),
      .groups = "drop"
    ))

  # Test: Do BP and TP differ in overall selectivity?
  if (n_distinct(mean_selectivity$Waveform) > 1) {
    cat("\n\nWilcoxon Test: Mean |E| ~ Waveform\n")
    wt_mean_sel <- wilcox.test(mean_abs_E ~ Waveform, data = mean_selectivity)
    print(wt_mean_sel)

    cat("\n\nInterpretation:\n")
    if (wt_mean_sel$p.value < 0.05) {
      cat("BP and TP differ significantly in overall selectivity (p < 0.05)\n")

      bp_mean <- mean(mean_selectivity$mean_abs_E[mean_selectivity$Waveform == "BP"])
      tp_mean <- mean(mean_selectivity$mean_abs_E[mean_selectivity$Waveform == "TP"])

      if (bp_mean > tp_mean) {
        cat("→ BP fish are MORE selective (choosier) than TP fish\n")
      } else {
        cat("→ TP fish are MORE selective (choosier) than BP fish\n")
      }
    } else {
      cat("No significant difference in overall selectivity (p >= 0.05)\n")
      cat("→ BP and TP are equally selective\n")
    }

    # Visualization
    p_mean_sel <- ggplot(mean_selectivity,
                         aes(x = Waveform, y = mean_abs_E, fill = Waveform)) +
      geom_boxplot(alpha = 0.7) +
      geom_jitter(width = 0.1, size = 3, alpha = 0.6) +
      theme_minimal() +
      scale_fill_brewer(palette = "Set1") +
      labs(
        title = "Mean Absolute Selectivity: BP vs. TP",
        subtitle = "Higher values = more selective (choosier)",
        x = "Waveform",
        y = "Mean |Ivlev's E|"
      ) +
      theme(legend.position = "none")

    print(p_mean_sel)
  }

} else {
  cat("PS1: Selectivity data not available.\n")
}

# ===== PRIMERSET 2 MEAN SELECTIVITY =====

if (exists("selectivity_ps2_df") && nrow(selectivity_ps2_df) > 0) {

  cat("\n\n==============================================\n")
  cat("PRIMERSET 2 - Mean Selectivity\n")
  cat("==============================================\n\n")

  # Calculate mean absolute selectivity per individual fish
  mean_selectivity_ps2 <- selectivity_ps2_df %>%
    group_by(SampleID, Region, Waveform, Locality) %>%
    summarise(
      mean_abs_E = mean(abs(Ivlev_E), na.rm = TRUE),
      n_families = n(),
      .groups = "drop"
    )

  cat("Mean Absolute Selectivity by Individual:\n")
  cat("(Higher |E| = more selective/choosy)\n\n")
  print(head(mean_selectivity_ps2, 10))

  cat("\n\nSummary by Waveform:\n")
  print(mean_selectivity_ps2 %>%
    group_by(Waveform) %>%
    summarise(
      n_fish = n(),
      mean_selectivity = mean(mean_abs_E),
      sd_selectivity = sd(mean_abs_E),
      .groups = "drop"
    ))

  # Test: Do BP and TP differ in overall selectivity?
  if (n_distinct(mean_selectivity_ps2$Waveform) > 1) {
    cat("\n\nWilcoxon Test: Mean |E| ~ Waveform\n")
    wt_mean_sel_ps2 <- wilcox.test(mean_abs_E ~ Waveform,
                                    data = mean_selectivity_ps2)
    print(wt_mean_sel_ps2)

    cat("\n\nInterpretation:\n")
    if (wt_mean_sel_ps2$p.value < 0.05) {
      cat("BP and TP differ significantly in overall selectivity (p < 0.05)\n")

      bp_mean <- mean(mean_selectivity_ps2$mean_abs_E[
        mean_selectivity_ps2$Waveform == "BP"])
      tp_mean <- mean(mean_selectivity_ps2$mean_abs_E[
        mean_selectivity_ps2$Waveform == "TP"])

      if (bp_mean > tp_mean) {
        cat("→ BP fish are MORE selective (choosier) than TP fish\n")
      } else {
        cat("→ TP fish are MORE selective (choosier) than BP fish\n")
      }
    } else {
      cat("No significant difference in overall selectivity (p >= 0.05)\n")
      cat("→ BP and TP are equally selective\n")
    }

    # Visualization
    p_mean_sel_ps2 <- ggplot(mean_selectivity_ps2,
                             aes(x = Waveform, y = mean_abs_E,
                                 fill = Waveform)) +
      geom_boxplot(alpha = 0.7) +
      geom_jitter(width = 0.1, size = 3, alpha = 0.6) +
      theme_minimal() +
      scale_fill_brewer(palette = "Set1") +
      labs(
        title = "PS2: Mean Absolute Selectivity: BP vs. TP",
        subtitle = "Higher values = more selective (choosier)",
        x = "Waveform",
        y = "Mean |Ivlev's E|"
      ) +
      theme(legend.position = "none")

    print(p_mean_sel_ps2)
  }

} else {
  cat("PS2: Selectivity data not available.\n")
}
```

## 4.8 Synthesis and Interpretation

For detailed interpretation guidelines, manuscript writing templates, and analytical strategy framework, see:

**[analysis_plan.md](./analysis_plan.md)**

```{r level4-synthesis-brief}
cat("=== LEVEL 4.8: SYNTHESIS ===\n\n")
cat("See analysis_plan.md for complete interpretation framework.\n\n")
cat("Analysis complete. Results ready for interpretation.\n")
```

---

**END OF DOCUMENT**
